### 2016-11-18 bare_conf.tex
Lets start today

%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{../png/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Michael Shell}
\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
Georgia Institute of Technology\\
Atlanta, Georgia 30332--0250\\
Email: http://www.michaelshell.org/contact.html}
\and
\IEEEauthorblockN{Homer Simpson}
\IEEEauthorblockA{Twentieth Century Fox\\
Springfield, USA\\
Email: homer@thesimpsons.com}
\and
\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
\IEEEauthorblockA{Starfleet Academy\\
San Francisco, California 96678--2391\\
Telephone: (800) 555--1212\\
Fax: (888) 555--1212}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area 
\maketitle
% As a general rule, do not put math, special symbols or citations 
% in the abstract

\begin{abstract} This work is focused on the prediction of affective
states using non-obtrusive sensors. Non-obtrusive sensors are preferred in this
work because of a hypothesis that states that an obtrusive sensor would alter
the affective states of a user. Specifically, the method uses data gathered from
a keyboard and mouse dynamics plugin embedded in a tutoring system designed for
the teaching of computer programming in the Python language. The tutoring system
requires from the user to practice their programming skills in order to progress
through the course, and the user’s creativity plays an important role in the
solving of programming exercises. Also a pre-processing method is applied, which
generates feature vectors consisting of 39 features representing each
interaction of the users by recording their keyboard and mouse dynamics. 149
feature vectors were generated and used as input of four different
classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% and kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states during their interaction with a programming tutoring system.
\end{abstract}

% no keywords

% For peer review papers, you can put extra information on the cover % page as needed: 
% \ifCLASSOPTIONpeerreview % \begin{center} \bfseries EDICS Category:  3-BBND \end{center} 
% \fi % % For peerreview papers, this IEEEtran command inserts a page break and 
% creates the second title. It will be ignored for other modes. \IEEEpeerreviewmaketitle

\section{Introduction} 
Introductory programming courses are
generally regarded as difficult \cite{robins2003learning,lahtinen2005study} 
and there is a common conception that they often have a high
failure rate \cite{bennedsen2007failure}. There are multiple factors involved.
Jenkins \cite{jenkins2001motivation} argues that some are deeply related with the expectations,
attitudes, and previous experiences of the teaching staff and students. Another
factor is the nature of the subject that involves the learning of new abstract
constructs, syntax and tools. Also groups are often heterogeneous and thus it is
difficult to design courses that are beneficial for everyone. Many students
begin to program when they are at their first year of university and until then
are confronted with a totally new topic that does not respond to their habitual
study approaches. Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
problem-solving intensive, and it requires high precision because even the
slightest perturbation can render a program totally worthless. These
difficulties often frustrate students. 
Jenkins \cite{jenkins2001motivation,jenkins2002difficulty} notes that many
students expect the course to be difficult and come with the notion that they
will have to struggle, others may have a stereotyped image of a programmer,
these beliefs can negatively affect their initial motivation. During the course,
novice programmers experience many emotions, for instance frustration and
confusion when they cannot find a bug in a program, but also joy when they
successfully run a challenging program for the first time. They can also become
bored if they found the exercises too repetitive or too easy. Learning to
program is a difficult task, where emotions play a significant role. Research
has gone far to understand the role of emotion in learning, both in the process
and the outcome; for instance, in e-learning  \cite{kort2001affective,rossin2009effects}
and in programming  \cite{rodrigo2009affective,jenkins2001motivation,
bosch2013emotions,khan2007mood}.

It is expected that flow/engagement is the ideal affective state in which
students tend to be most capable of acquiring meaningful information through the
learning process. Flow is defined by Csikszentmihalyi \cite{csikszentmihalyi1990flow}
as a mental state in which a person performing an activity is fully immersed in a feeling of
energized focus, full involvement, and enjoyment. Engagement is also a positive
affect where students are thought to be more involved behaviourally,
intellectually, and emotionally in their learning tasks \cite{bangert2002teacher}.
When designing instruction, an objective is to assign learning activities
that result challenging to students, but not much as to frustrate them. For
this, experienced instructors gauge a student’s affective state and then assign
an activity with the appropriate level of difficulty. In affect-aware
intelligent learning environments (ILEs), emotions have a central role in the
students' user model. However, recognition of emotions is a difficult task, even
for humans. Instructors use their social intelligence often to recognize
students’ affective states. In class a good instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. ILEs can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, as many researchers have
demonstrated affect-aware computers can provide better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. In order to perform recognition of the users' affective states,
sensors must be used to gather data. Some sensors rely on physiological readings
and must be in physical contact with students. These sensors can be considered
intrusive or invasive, and can disrupt the student’s learning experience 
\cite{zhai2008stress,sidney2005integrating,arroyo2009emotion}.
Other sensors such as video cameras can also be considered invasive, since they always transmit the
identity, appearance and behaviour along with emotional data \cite{picard2001toward}.

In this work, a method for the recognition of affective states through keystroke
and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
of keystrokes according to their affective state when programming.  The method
has three main components: a Javascript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using a web based learning environment. Using only the data from the
student’s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a naïve Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

If a relationship could be modelled, ILEs could adapt the instruction using the
same devices required to input the program to the computer. This could be
important because writing programs in a computer is an important learning
activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
\cite{lahtinen2005study} is reported that students rated ``working alone on programming
coursework'' as a more useful situation than lectures.

The structure of this work is organized as follows: Section Related Work
presents a series of works related to the proposed method in this paper; Section
Proposed Method describes the proposed method for the recognition of affective
states in a web based learning environment for the teaching of programming
languages; Section Experiment explains the experimental evaluation of the
proposed method, following by a Results sections and finally, a Conclusion and
Future Work are discussed.

\section{Related Work} 
Affect recognition is an active field of research. Many
methods have been proposed, some require the intervention of the user to fill up
questionnaires or forms; these selections remain static until the user changes
the values. These methods are easy to implement, but cannot detect dynamic
changes.  A more dynamic approach requires the use of sensors to capture
affective states as they change. The context, the environment and the learning
activity determine what kind of sensors can be used. The most common learning
environment is the classroom, a physical space with context to facilitate
learning. But learning is possible in a wide variety of settings, such as
outside-of-school locations and outdoor environments these are sometimes
referred as ubiquitous learning environments, an example of such environments is
the work of Yang \cite{yang2006context} in a context-aware environment, but missing affective
information. There are also virtual learning environments such as the one proposed by Dillenburg
\cite{dillenbourg2002virtual}  where learners can have avatars, and virtual places
where they can play roles and socialize.

In the general context of learning there have been some approaches for affect
recognition. The work of Kapoor y Picard \cite{kapoor2005multimodal} uses a multi modal
approach using sensory information from facial expressions and postural shifts
of the learner combined with information about the learner's activity on the
computer; the learning activity was solving a puzzle on a computer. They report
using multimodal Gaussian Process approach achieving an accuracy of over 86\%.
Sidney, et al., \cite{sidney2005integrating,} enhances the intelligent tutor system AutoTutor with
affective recognition using nonintrusive sensory information from facial
expressions, gross body movements, and conversational cues from logs.  The
subject material of that system consisted in lectures of computer literacy.

Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
with an agent in a virtual environment. In this work an agent called Steve
responds emotionally to other agents and interactive users. The agent simulates
his emotions through different multimedia modes including facial expressions and
speech. Affective reasoning agents were used for training, by putting students
in work related situations. Agents could react to the behavior of students, for
instance if a student was being careless in a task and was in a dangerous
situation, Steve would show distress or fear. Also in a virtual environment the
work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
investigating the affective transitions that occur throughout narrative-centered
learning experiences. The analysis of affective state transitions in this work
replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al. 
\cite{rodrigo2009affective} where
also engagement/flow dominated self-reported affect. For outdoor environments
Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
recognition.  The results about emotion recognition from physiological signals
achieved a best-case accuracy (86.3\%) for four types of learning emotions.

Bosch, D'Mello \& Mills \cite{bosch2013emotions} analysed the relationship between affective states
and performance of novice programmers when they were learning the basics of
computer programming in the Python language. The results of their study
indicated that the more common emotions students experienced were engaged
(23\%), confusion (22\%), frustration (14\%), and boredom (12\%). It was useful
to consider these results, as it presented evidence of what affective states
need to be targeted in order to obtain less biased data. For example, if a less
common emotion was chosen, a classifier could opt to classify any feature vector
as not experiencing such emotion. Nevertheless, the classifier would obtain
accurate results, although the classifier would be inaccurate at determining if
a feature vector was actually experiencing the given affective state. Similar to
the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
behaviours relate to student's achievement within a basic computer science
course. The authors found that confusion, boredom and engagement in IDE-related
(on-task) conversation are associated with lower achievement.

There has been very little research reported on the effectiveness of the use of
keyboard and mouse dynamics as a sensory channel for affective recognition, and
the few have not been focused on programming. The preliminary work by Zimmermann
et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
mouse) with an emotional state. Several physiological parameters were measured
concurrent with the task. The parameters included respiration, pulse, skin
conductance level, and corrugator activity. The task assigned to subjects was to
shop on an e-commerce website for office-supplies and finally write a predefined
message to the shop operator. Subjects where watching videos in order to change
their emotional state during the experiment. The method was validated using
self-reported emotions using the self-assessment-manikin (SAM), devised by Lang
\cite{lang1980behavioral}, designed to assess the dimensions valence, arousal and dominance
directly by means of three sets of graphical manikins. Results show that the
film clips were effective in inducing the expected mood changes, but no further
empirical results are presented. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
uses sensory data based on the time elapsed between each key press, the task
given to users was to write a free-text and used linguistic features in order to
recognize both physical and emotional stress. The results show a classification
accuracy of 62.5\% for physical stress and 75\% from emotional stress; authors
argue that these results are comparable with other approaches in affective
computing. They also stress that their methods must be validated further in
other contexts. Moods may have an impact on programmer’s performance according
to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
on the basis of information regarding the programmer’s use of the keyboard and
mouse, and to integrate them into development environments that can improve
programmer performance. They briefly describe a further experiment that could
use keyboard and mouse but only as future work. There are other studies about
the behaviour of programmers, which are not directly concerned with affective
states but are nevertheless important to their performance. Eye tracking in
computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
also used for assessing learner´s comprehension of C++ and Python programs in
Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed 
an automated technique to
assess, analyse and visualize students learning computer programming. Blikstein
employs different quantitative techniques to extract students’ behaviours and
categorize them in terms of programming experience.

Research works related to Keystroke Dynamics (KD) are carried out either using
fixed-texts, or free-texts Gunetti y Picardi \cite{gunetti2005keystroke}.
KD performed on fixed-texts
involves the recognition of typing patterns when typing a pre-established fixed-
length text, e.g., a password. In the other case, free-text KD achieves the
recognition of typing patterns when typing a text of arbitrary-length, e.g., a
description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
most of
the research regarding KD is done on fixed-text input, the reason being that
fixed-text KD usually yields better results than free-text KD. Yet, the authors
of this work share the opinion with Janakiraman, R., and Sim, T., that it would
be more useful if KD can handle free text as well as fixed text, this is also a
requirement if the text is a program.

Although the use of Keystroke Dynamics (KD) is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed-text their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models.

Regarding Mouse Dynamics (MD), some research has been conducted for the
identification of affective states, although, as with the case of KD, MD is
mainly used as a biometric measure for authentication. Salmeron-Majadas, Santos
and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
five different classification algorithms. Bakhtiyari y Husain \cite{bakhtiyari2014fuzzy} discuss a
method based on fuzzy models for the recognition of emotions through KD, MD and
touch-screen interactions. For a broad review of emotion recognition methods
based on KD and MD, the work by Kolakowska \cite{kolakowska2013review} is recommended.

\section{Proposed Method} 
The goal of this work is to propose an affective
recognition method based on the sensory data provided by the keyboard and mouse
dynamics generated by a learner as he/she types a programing exercise. The main
components of the method are depicted in Figure 1. A brief explanation of the
whole process is explained next, details come later. The process starts when the
learner begins to type a program in a browser-based editor. As she types or
moves the mouse the dynamics are recorded. When the learner submits the code to
evaluation, the request includes the sensory data along with the code and
information about the session. In the server, the code is evaluated in an
external virtual machine that provides a sand box to prevent malicious or
erroneous code to halt the server. When the result is ready, it is recorded
along with the sensory data and sent to the preprocessing module. The output is
a feature vector, ready for classification. A previously trained classifier is
responsible for the classification, which outputs the predicted affective state.
The method does not consider other sensory data, but it could be integrated with
other sensory inputs in a multi-modal approach.  Each of the steps is explained
in detail next.

\subsubsection{Capturing the Keystroke and Mouse Data} 
While a student is trying
to solve an exercise, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a timestamp in milliseconds (using the method
getTime() of Javascript's built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device.  The interface of the web-based
editor is shown in Figure 2. Programming exercises are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. An example execution is shown in the left
side image of Figure 2. In this example the learner was asked to write a
function to add two numbers, in Python. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories at http://git.io/vJlUV also the code for the
sandbox is in http://git.io/vJlUj.

\subsubsection{Preprocessing of the Keystroke and Mouse Data} 
The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise. These events
have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman (2007, June) are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.

To calculate the average and standard deviations of these presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner's affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded how many attempts
a student required before successfully solving an exercise. This number of
attempts is also included in the feature vector. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying} and are shown in Table 1. (The code used for this step is available at:
https://github.com/amherag/keyboard-mouse-dynamics).

Once feature vectors are obtained from an experiment, the generated dataset is
normally used as training data for a classifier. Researchers of affective
recognition have used a wide variety of classification algorithms. As a proof of
concept four well known classification algorithms where compared: k-Nearest
Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
propagation, a naïve Bayes classifier and finally a decision trees algorithm for
rational data (J-48). Details for these algorithms, can be found in the textbook
by Tan, Steinbach, \& Kumar \cite{tan2006introduction}. The accuracies obtained by these algorithms
as well as the parameters used are reported in the next section.

\section{Experiment} 
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system. The context of use is a learning environment for programmers. In
particular for learning activities consisting in writing short programs
interactively. An experimental approach was adopted with this aim. Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared and came
to conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many exercises as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the exercises or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.

Figure 4. Web-based editor used for code evaluation. A tutorial was developed to
obtain the necessary data using Protoboard, a web-based programming learning
environment (the latest version can be found online at
http://python.protoboard.org/). Protoboard has the functionality described in
the proposed method and shown schematically in Figure 1. Users-log in to
Protoboard using their account credentials from a popular social network
(Facebook).  The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to solve the programming
exercises in the course. What follows after these videos, are 13 programming
exercises that the students need to solve in consecutively. Exercises are of
incremental difficulty. In Figures 2 and 3, the interface of this platform is
presented. For this experiment Protoboard was configured to allow unlimited
tries to solve an exercise.

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network. Learners were required to login with their
Facebook credentials, this had the advantage of not requiring the system to
maintain users’ credentials or profile information as this is kept outside in an
external service. Unfortunately this decision also kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services. Other requirement was that they where all
adults, this age is 18 years in Mexico where the experiment was conducted. The
requirement of the social network account stopped several prospects from
volunteering, but on the other hand gave researchers access to public
information about the subjects. All the learners were either current students or
graduates of Computer Systems Engineering. Their ages were in the range of 18 to
30 years old. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, all the participants were
required had completed at least one course in software programming. Results from
the initial survey are shown in Figure 4.

In order to determine what affective states a student was experiencing, an
Experience Sampling Method (ESM) was used (Csikszentmihalyi y Larson, 1987).
After the students successfully solve a programming exercise, they are presented
with an ESM survey that asks what they were feeling during their solving of the
exercise. A very brief description is given about what to do in this survey,
followed by statements the students need to answer according to how they were
feeling. As an example, the statement “I was feeling frustrated” is presented,
and a student needs to answer either “Strongly agree,” “Agree,” “Neutral,”
“Disagree,” and “Strongly Disagree.”

After the two weeks of the experiment, only four learners completed all of the
programming exercises and 22 did not completed any. Out of the total activities
available (videos, survey and questionnaires) only two completed all. Figure 5
shows the number of exercises and activities completed by each learner.

Figure 5. Web-based editor used for code evaluation.

Figure 6. Class distribution of emotions as reported by learners. The
participants' interaction generated a total of 142 feature vectors one for each
successfully completed exercise. The affective states reported by learners after
completing each exercise was grouped in three classes: ‘Yes’, ‘Neutral’ and
‘No’. The class distribution of the emotions reported is shown in Figure 6. This
results show that the most common emotions were flow/engagement (72\%) and
relaxation (61\%) while few learners reported distraction (5\%), frustration
(8\%) and boredom (2\%). This distribution is different from what was reported
by D’Mello et al. (2013) and Baker et al. (2007). Although Flow/Engagement was
also the predominant class, the distribution is skewed to the first two. There
are some possible reasons for this; first the majority of students had
experience in programming, so learning new one is not that problematic, a second
reason could be the freedom users had to abandon the activities, perhaps
frustrated or bored learners simply quit the tutorial. A high number of learners
(49\%) did not completed any exercise and the once who completed more where also
the more experienced.

Classifiers where implemented using the software Rapid Miner using a feature
selection module as part of the process. A genetic algorithm was used in the
user selection module. In the end, the subset for the frustration classifier
consists of 11 features, and the subset for the boredom classifier consists of
13 features. The performance of the classifiers is shown in Table 1. Table 1.
The components of the feature vector; adapted from Epp, Lippold y Mandryk
(2007).    Naïve Bayes J-48  k-NN  ANN Affect  Accuracy  k Accuracy  k Accuracy
k Accuracy  k Flow/Engaged  79.00\% (0.11) 0.50  80.48\% (0.10) 0.51
76.14(0.09) 0.43  78.43(0.12) 0.45 Relaxation  71.10\% (0.09) 0.40  72.57\%
(0.11) 0.45  69.14\%(0.08)  0.37  71.24\%(0.07)  0.34 Distraction  70.33\%
(0.13)  0.43  71.00\% (0.10) 0.43  74.00\%(0.06)  0.50  72.52\%(0.07)  0.39
Frustration 74.71\% (0.09) 0.49  73.86\% (0.12) 0.45  72.62\%(0.09)  0.42
78.0\%(0.09)  0.51 Boredom 74.71\% (0.13) 0.47  84.57\% (0.06) 0.59
83.81\%(0.06)  0.58  76.19\%(0.09)  0.45

The results obtained with this method are satisfactory, considering a free-text
was used. The accuracies and kappa coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk (2007). In the case of the kappa coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of kappa were close to 0.5, is important to consider the values of kappa
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\% marginally better than the others, and an accepted kappa
statistic for the kind of problem. Artificial neural network give the highest
accuracy in the classification of frustration.

\section{Conclusion}

% An example of a floating figure using the graphicx package. 
% Note that \label must occur AFTER (or within) \caption. 
% For figures, \caption should occur after the \includegraphics. 
% Note that IEEEtran v1.7 and later has special internal code that 
% is designed to preserve the operation of \label within \caption 
% even when the captionsoff option is in effect. However, because 
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}. % 
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class 
% option should be used if it is desired that the figures are to be % displayed while in draft mode. 
%
\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{KMD_Affective} 
% where an .eps filename suffix will be assumed under latex,  
% and a .pdf suffix will be assumed for pdflatex; or what has been declared 
% via \DeclareGraphicsExtensions. 
\caption{Simulation results for the network.}
\label{fig_sim} 
\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% use section* for acknowledgment 
\section*{Acknowledgment} 
This received funding from Tecnólogico Nacional de México through the project: Técnicas de
computación inteligente para el secuenciado adaptativo de ejercicios de
programación en la nube.

% trigger a \newpage just before the given reference 
% number - used to balance the columns on the last page 
% adjust value as needed - may need to be readjusted if 
% the document is modified later %\IEEEtriggeratref{8} 
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file 
% BibTeX documentation can be easily obtained at: 
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/ 
% The IEEEtran BibTeX style support page is at: 
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran} 
% argument is your BibTeX string definitions and bibliography database(s) 
\bibliography{programming} 
% 
% <OR> manually copy in the resultant .bbl file 
% set second argument of \begin to the number of references 
% (used to reserve space for the reference number labels box)

% that's all folks 
\end{document}

### 2016-11-22 bare_conf.tex
First draft completed

@@
-%% bare_conf.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 \documentclass[conference]{IEEEtran}
-% Some Computer Society conferences also require the compsoc mode option,
-% but others use the standard conference format.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/pkg/ifpdf
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
 
@@
 %
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of the IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 5.0 (2009-03-20) and later if using hyperref.sty.
-% The latest version can be obtained at:
-% http://www.ctan.org/pkg/cite
-% The documentation is contained in the cite.sty file itself.
+\usepackage{cite}
 
@@
   % declare the path(s) where your graphic files are
-\graphicspath{{../png/}{../jpeg/}}
+\graphicspath{{./img/}}
   % and their extensions so you won't have to specify these with
@@
 \else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
+
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at: 
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
 
-% *** MATH PACKAGES ***
-%
-%\usepackage{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics.
-%
-% Note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/amsmath
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/pkg/algorithms
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/pkg/algorithmicx
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/array
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/pkg/subfig
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure.
-% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
-% corrections already built into the system in which case a warning will
-% be issued if an attempt is made to load fixltx2e.sty as it is no longer
-% needed.
-% The latest version and documentation can be found at:
-% http://www.ctan.org/pkg/fixltx2e
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/stfloats
-% Do not use the stfloats baselinefloat ability as the IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that the IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/pkg/dblfloatfix
-
+\usepackage{stfloats}
 % *** PDF, URL AND HYPERLINK PACKAGES ***
 %
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/url
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+\usepackage{url}
 
@@
 \begin{document}
-%
+
 % paper title
@@
 
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
 % use for special paper notices
@@
 \maketitle
-% As a general rule, do not put math, special symbols or citations 
-% in the abstract
-
 \begin{abstract} This work is focused on the prediction of affective
@@
 % no keywords
-
-% For peer review papers, you can put extra information on the cover % page as needed: 
-% \ifCLASSOPTIONpeerreview % \begin{center} \bfseries EDICS Category:  3-BBND \end{center} 
-% \fi % % For peerreview papers, this IEEEtran command inserts a page break and 
-% creates the second title. It will be ignored for other modes. \IEEEpeerreviewmaketitle
 
@@
 using multimodal Gaussian Process approach achieving an accuracy of over 86\%.
-Sidney, et al., \cite{sidney2005integrating,} enhances the intelligent tutor system AutoTutor with
+Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
 affective recognition using nonintrusive sensory information from facial
@@
 
+\begin{figure}[!t] 
+\centering 
+\includegraphics[width=2.5in]{KMDAffective.png} 
+\caption{Code Execution and Affect Recognition Architecture.}
+\label{fig_process} 
+\end{figure}
+
 Research works related to Keystroke Dynamics (KD) are carried out either using
@@
 dynamics generated by a learner as he/she types a programing exercise. The main
-components of the method are depicted in Figure 1. A brief explanation of the
+components of the method are depicted in Figure \ref{fig_process} A brief explanation of the
 whole process is explained next, details come later. The process starts when the
@@
 in detail next.
+
+\begin{table*}[!t]
+\centering
+\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+    \begin{tabular}{ | l | l || l | l | }
+    \hline
+    ID          & Description           & ID               & Description  \\
+    \hline
+    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
+    \hline
+    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
+    \hline
+    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
+    \hline
+    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
+    \hline
+    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
+    \hline
+    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
+    \hline
+    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
+    \hline
+    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
+    \hline
+    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
+    \hline
+    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
+    \hline
+    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
+    \hline
+    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
+    \hline
+    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
+    \hline
+    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
+   \hline
+    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
+    \hline
+    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
+    \hline
+    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
+    \hline
+    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
+    \hline
+    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
+    \hline
+    20 & Standard deviation of the previous feature. &   &  \\
+    \hline
+  \end{tabular}
+\end{table*}
 
@@
 in a mobile device or any web-enabled device.  The interface of the web-based
-editor is shown in Figure 2. Programming exercises are evaluated using unit
+editor is shown in Figure \ref{fig_editor}. Programming exercises are evaluated using unit
 tests; the results of the evaluation are shown to users. If all tests the
 program is considered to be correct. An example execution is shown in the left
-side image of Figure 2. In this example the learner was asked to write a
+side image of Figure \ref{fig_editor}. In this example the learner was asked to write a
 function to add two numbers, in Python. The source code for the Protoboard web
@@
 
+\begin{figure}[!t] 
+\centering 
+\includegraphics[width=2.5in]{editorRresult.png} 
+\caption{Web-based editor used for code evaluation.}
+\label{fig_editor} 
+\end{figure}
+\begin{figure}[b!] 
+\centering 
+\includegraphics[width=2.5in]{KeyDyn.png} 
+\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
+\label{fig_KD} 
+\end{figure}
 \subsubsection{Preprocessing of the Keystroke and Mouse Data} 
@@
 events, as it is a common practice when dealing with keystroke dynamics. In this
-work, the definitions proposed by Sim y Janakiraman (2007, June) are used. Held
+work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
 time (Ht) is defined as the time between a key-down and a key-up of the same
@@
 
-Figure 4. Web-based editor used for code evaluation. A tutorial was developed to
+\begin{figure}[!t] 
+\centering 
+\includegraphics[width=2.5in]{Completed.png} 
+\caption{Completed activities and exercises and by learner in ascending order.}
+\label{fig_completed} 
+\end{figure}
+
+\begin{figure}[!t] 
+\centering 
+\includegraphics[width=2.5in]{classDist.png} 
+\caption{Class distribution of emotions as reported by learners.}
+\label{fig_class} 
+\end{figure}
+
+A tutorial was developed to
 obtain the necessary data using Protoboard, a web-based programming learning
@@
 the initial survey are shown in Figure 4.
+\begin{table*}[!t]
+\centering
+\caption{The performance of the classifiers.}
+    \begin{tabular}{ | c | l | l | l | l | }
+    \hline
+    Affect          & Naïve Bayes           & J-48                & k-NN              & ANN \\
+    \hline
+    Flow/engaged    & 79.00\% (0.11) 0.50   & 80.48\% (0.10) 0.51 & 76.14(0.09) 0.43    & 78.43(0.12) 0.45 \\
+    \hline
+    Relaxation      & 71.10\% (0.09) 0.40   & 72.57\% (0.11) 0.45 & 69.14\%(0.08)  0.37 & 71.24\%(0.07)  0.34\\
+    \hline
+    Distraction     & 70.33\% (0.13)  0.43  & 71.00\% (0.10) 0.43 & 74.00\%(0.06)  0.50 & 72.52\%(0.07)  0.39\\
+    \hline
+    Frustration     & 74.71\% (0.09) 0.49   & 73.86\% (0.12) 0.45 & 72.62\%(0.09)  0.42 & 78.0\%(0.09)  0.51 \\
+    \hline
+    Boredom         & 74.71\% (0.13) 0.47   & 84.57\% (0.06) 0.59 & 83.81\%(0.06)  0.58 & 76.19\%(0.09)  0.45\\
+    \hline
+    \end{tabular}
+\end{table*}
 
@@
 
-Figure 5. Web-based editor used for code evaluation.
-
-Figure 6. Class distribution of emotions as reported by learners. The
-participants' interaction generated a total of 142 feature vectors one for each
+The participants' interaction generated a total of 142 feature vectors one for each
 successfully completed exercise. The affective states reported by learners after
@@
 (8\%) and boredom (2\%). This distribution is different from what was reported
-by D’Mello et al. (2013) and Baker et al. (2007). Although Flow/Engagement was
+by D’Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
+Although Flow/Engagement was
 also the predominant class, the distribution is skewed to the first two. There
@@
 consists of 11 features, and the subset for the boredom classifier consists of
-13 features. The performance of the classifiers is shown in Table 1. Table 1.
-The components of the feature vector; adapted from Epp, Lippold y Mandryk
-(2007).    Naïve Bayes J-48  k-NN  ANN Affect  Accuracy  k Accuracy  k Accuracy
-k Accuracy  k Flow/Engaged  79.00\% (0.11) 0.50  80.48\% (0.10) 0.51
-76.14(0.09) 0.43  78.43(0.12) 0.45 Relaxation  71.10\% (0.09) 0.40  72.57\%
-(0.11) 0.45  69.14\%(0.08)  0.37  71.24\%(0.07)  0.34 Distraction  70.33\%
-(0.13)  0.43  71.00\% (0.10) 0.43  74.00\%(0.06)  0.50  72.52\%(0.07)  0.39
-Frustration 74.71\% (0.09) 0.49  73.86\% (0.12) 0.45  72.62\%(0.09)  0.42
-78.0\%(0.09)  0.51 Boredom 74.71\% (0.13) 0.47  84.57\% (0.06) 0.59
-83.81\%(0.06)  0.58  76.19\%(0.09)  0.45
+13 features. The performance of the classifiers is shown in Table 1.
 
@@
 \section{Conclusion}
+The aim of this research was to evaluate the use of keyboard and mouse dynamics 
+as an appropriate sensory input for an affective recognition system. The context
+of use would be learning environment for programmers. In particular for learning
+activities consisting in writing short programs interactively. An experimental
+approach was adopted with this aim. The proposed method in this work obtained
+satisfactory results. It is usual for a classification method based on free-text
+to obtain an accuracy and kappa below their counterparts of fixed-text, and in this
+work the results obtained were similar to those works applied to fixed-text 
+dynamics. A possible explanation of this would be the addition of the mouse 
+dynamics features and the additional preprocessing performed on the feature vectors. 
+While these are promising results further experiments are needed. Other experiments
+should focus on novice programmers.
 
-% An example of a floating figure using the graphicx package. 
-% Note that \label must occur AFTER (or within) \caption. 
-% For figures, \caption should occur after the \includegraphics. 
-% Note that IEEEtran v1.7 and later has special internal code that 
-% is designed to preserve the operation of \label within \caption 
-% even when the captionsoff option is in effect. However, because 
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}. % 
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class 
-% option should be used if it is desired that the figures are to be % displayed while in draft mode. 
-%
-\begin{figure}[!t] 
-\centering 
-\includegraphics[width=2.5in]{KMD_Affective} 
-% where an .eps filename suffix will be assumed under latex,  
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared 
-% via \DeclareGraphicsExtensions. 
-\caption{Simulation results for the network.}
-\label{fig_sim} 
-\end{figure}
-
-% Note that the IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
+Future work will focus on extracting other features from the learners’ interaction
+like behavioral data, as proposed by Blikstein (2011); the data includes the size
+of the program and changes in the code. Other authors also propose treating the use
+of certain keys as features e.g., the delete key because is manly used for correcting
+sentences. Also a multimodal approach is needed.
 
@@
 
-% can use a bibliography generated by BibTeX as a .bbl file 
-% BibTeX documentation can be easily obtained at: 
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/ 
-% The IEEEtran BibTeX style support page is at: 
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
 \bibliographystyle{IEEEtran} 
@@
 \bibliography{programming} 
-% 
-% <OR> manually copy in the resultant .bbl file 
-% set second argument of \begin to the number of references 
-% (used to reserve space for the reference number labels box)
-
-% that's all folks 
 \end{document}

### 2016-11-23 bare_conf.tex
New abstract needed

@@
 % affiliations
-\author{\IEEEauthorblockN{Michael Shell}
+\author{\IEEEauthorblockN{Kevin Bacon}
 \IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
@@
 Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
+Email: http://www.kev.org/contact.html}
 \and
@@
 % make the title area
+
 \maketitle
-\begin{abstract} This work is focused on the prediction of affective
-states using non-obtrusive sensors. Non-obtrusive sensors are preferred in this
-work because of a hypothesis that states that an obtrusive sensor would alter
-the affective states of a user. Specifically, the method uses data gathered from
-a keyboard and mouse dynamics plugin embedded in a tutoring system designed for
-the teaching of computer programming in the Python language. The tutoring system
-requires from the user to practice their programming skills in order to progress
-through the course, and the user’s creativity plays an important role in the
-solving of programming exercises. Also a pre-processing method is applied, which
-generates feature vectors consisting of 39 features representing each
-interaction of the users by recording their keyboard and mouse dynamics. 149
-feature vectors were generated and used as input of four different
-classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
-Networks. These classification models are then compared by their performances
-when classifying learners into five different affective states: boredom,
-frustration, distraction, relaxation and engagement. Each of the classification
-models achieved an accuracy of around 75\% and kappa of around 0.45, and, in
-average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
-that the data gathered from the non-obtrusive sensors can successfully be used
-as another input to classification models in order to predict an individual's
-affective states during their interaction with a programming tutoring system.
+\begin{abstract} Learning a programming language is a hard thing to do.
+
 \end{abstract}

### 2016-11-23 paper.tex
Few changes here and there
Abstract

--- 2016-11-23 bare_conf.tex
@@
 \maketitle
-\begin{abstract} Learning a programming language is a hard thing to do.
+\begin{abstract}
+
+Learning to program is often regarded as a difficult task, courses are 
+problem-solving intensive and often students feel frustrated when doing programming
+exercises. Instructors aim to design learning activities that result
+challenging to students, but not to the extreme as to frustrate them. When
+selecting an appropiate programing excercise, experienced instructors gauge a
+student’s affective state and skills to then assign an activity with the  appropriate level
+of difficulty. This work is focused on the prediction of the affective states
+of programmers with different levels of expertise when learning a new
+programming language. For this, an interactive web-based programming platform
+is proposed. The platform is designed to collect data from the students'
+interaction for data analysis. Current work is focused on the prediction of
+affective states using non-obtrusive sensors. Specifically, the proposed method uses data
+gathered from both keyboard and mouse dynamics. The data set was generated
+as students learned the basics of Python programming. Feature vectors 
+were generated with 39 features representing the dynamics of the interaction. 
+In total 149 feature vectors were generated and used as input of four different
+classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
+Networks. These classification models are then compared by their performances
+when classifying learners into five different affective states: boredom,
+frustration, distraction, relaxation and engagement. Each of the classification
+models achieved an accuracy of around 75% with a kappa of around 0.45, and, in
+average, J-48 was the better algorithm with an accuracy of 75.97%. Results show
+that the data gathered from the non-obtrusive sensors can successfully be used
+as another input to classification models in order to predict an individual's
+affective states.
 
@@
 intellectually, and emotionally in their learning tasks \cite{bangert2002teacher}.
+
 When designing instruction, an objective is to assign learning activities
@@
 this, experienced instructors gauge a student’s affective state and then assign
-an activity with the appropriate level of difficulty. In affect-aware
-intelligent learning environments (ILEs), emotions have a central role in the
-students' user model. However, recognition of emotions is a difficult task, even
+an activity with the appropriate level of difficulty. However, recognition of 
+emotions is a difficult task, even
 for humans. Instructors use their social intelligence often to recognize

### 2016-11-23 paper.tex
Definition of ILEs acronym
Interactive Learning Environment, could also be Intelligent but not yet.

@@
 faces of students in the classroom to see if they are confused, bored or
-engaged; then decides what to do next. ILEs can be enhanced if they can offer an
+engaged; then decides what to do next. Interactive Learning Environments (ILEs)
+can be enhanced if they can offer an
 automatic perception of emotions. The recognition and simulation of human
@@
 An experiment was conducted where students solved a series of programming
-exercises using a web based learning environment. Using only the data from the
+exercises using the web based learning environment. Using only the data from the
 student’s keyboard and mouse dynamics six affective states where recognized for

### 2016-11-23 paper.tex
Changes galore and suggestions

--- 2016-11-23 bare_conf.tex
@@
 \maketitle
-\begin{abstract} Learning a programming language is a hard thing to do.
+\begin{abstract} 
+Learning a programming language is a hard thing to do. %Teaching or learning?
 
@@
 \section{Introduction}
+
 Introductory programming courses are
@@
 and there is a common conception that they often have a high
-failure rate \cite{bennedsen2007failure}. There are multiple factors involved.
-Jenkins \cite{jenkins2001motivation} argues that some are deeply related with the expectations,
+failure rate \cite{bennedsen2007failure}. There are multiple factors
+involved in this lack of success.
+Jenkins \cite{jenkins2001motivation} argues that there is a deep  relation with the expectations,
 attitudes, and previous experiences of the teaching staff and students. Another
 factor is the nature of the subject that involves the learning of new abstract
-constructs, syntax and tools. Also groups are often heterogeneous and thus it is
+constructs, syntax and tools. % Same guy here? - JJ
+Also groups are often heterogeneous and thus it is
 difficult to design courses that are beneficial for everyone. Many students
-begin to program when they are at their first year of university and until then
-are confronted with a totally new topic that does not respond to their habitual
-study approaches. Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
-problem-solving intensive, and it requires high precision because even the
+begin to program when they are at their first year of university which
+means they are tackling a totally new topic that does not respond to their habitual
+study approaches. % same reference? You should itemize the claims - JJ
+On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
+problem-solving intensive, which means that a  high precision is required because even the
 slightest perturbation can render a program totally worthless. These
@@
 
-It is expected that flow/engagement is the ideal affective state in which
+So it is expected that flow/engagement is the ideal affective state in which
 students tend to be most capable of acquiring meaningful information through the
@@
 
+%But the problem is how to recognize this state so that it can be used
+%to (select appropriate content| give grades| whatever ) - JJ
 In this work, a method for the recognition of affective states through keystroke
 and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
-of keystrokes according to their affective state when programming.  The method
+of keystrokes according to their affective state when programming. 
+% since Bosch or whatever has already proved that this is possible (or
+% whatever) --> Establish the state of the art and how you plan to
+% advance from it - JJ
+ The method
 has three main components: a Javascript library to capture keyboard and mouse
@@
 
-The structure of this work is organized as follows: Section Related Work
+The structure of this work is organized as follows: next Section 
 presents a series of works related to the proposed method in this paper; Section
-Proposed Method describes the proposed method for the recognition of affective
+\ref{sec:method} describes the proposed method for the recognition of affective
 states in a web based learning environment for the teaching of programming
-languages; Section Experiment explains the experimental evaluation of the
-proposed method, following by a Results sections and finally, a Conclusion and
-Future Work are discussed.
+languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
+proposed method. % There is no results section so far
+Finally, we draw some conclusions about our methodology and
+experimients.
 
 \section{Related Work}
+
 Affect recognition is an active field of research. Many
@@
 computer; the learning activity was solving a puzzle on a computer. They report
-using multimodal Gaussian Process approach achieving an accuracy of over 86\%.
+using a multimodal Gaussian Process approach achieving an accuracy of over 86\%.
 Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
@@
 \section{Proposed Method}
+\label{sec:method}
+
 The goal of this work is to propose an affective
@@
 available as Github repositories at http://git.io/vJlUV also the code for the
-sandbox is in http://git.io/vJlUj.
+sandbox is in http://git.io/vJlUj. % Anonimyze this, I guess - JJ
 
@@
 
-To calculate the average and standard deviations of these presses, the delays
+To calculate the average and standard deviations of these key presses, the delays
 between a key-down and a key-up event of the left button clicks are used. In
@@
 \cite{epp2011identifying} and are shown in Table 1. (The code used for this step is available at:
-https://github.com/amherag/keyboard-mouse-dynamics).
+https://github.com/amherag/keyboard-mouse-dynamics). %anonymize
 
@@
 \section{Experiment}
+\label{sec:exp}
 The aim of this research is to evaluate the use of keyboard

### 2016-11-30 paper.tex

@@
 \begin{abstract}
-Learning a programming language is a hard thing to do. %Teaching or learning?
+
+Learning to program is often regarded as a difficult task, courses are 
+problem-solving intensive and often students feel frustrated when doing programming
+exercises. Instructors aim to design learning activities that result
+challenging to students, but not to the extreme as to frustrate them. When
+selecting an appropiate programing excercise, experienced instructors gauge a
+student’s affective state and skills to then assign an activity with the  appropriate level
+of difficulty. This work is focused on the prediction of the affective states
+of programmers with different levels of expertise when learning a new
+programming language. For this, an interactive web-based programming platform
+is proposed. The platform is designed to collect data from the students'
+interaction for data analysis. Current work is focused on the prediction of
+affective states using non-obtrusive sensors. Specifically, the proposed method uses data
+gathered from both keyboard and mouse dynamics. The data set was generated
+as students learned the basics of Python programming. Feature vectors 
+were generated with 39 features representing the dynamics of the interaction. 
+In total 149 feature vectors were generated and used as input of four different
+classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
+Networks. These classification models are then compared by their performances
+when classifying learners into five different affective states: boredom,
+frustration, distraction, relaxation and engagement. Each of the classification
+models achieved an accuracy of around 75% with a kappa of around 0.45, and, in
+average, J-48 was the better algorithm with an accuracy of 75.97%. Results show
+that the data gathered from the non-obtrusive sensors can successfully be used
+as another input to classification models in order to predict an individual's
+affective states.
 

### 2016-11-30 paper.tex
Revise figure references
They are hard-wired, maybe wrong and should anyway be substituted by references.  Check it out @mariosky

@@
 particular for learning activities consisting in writing short programs
-interactively. An experimental approach was adopted with this aim. Sensory and
+interactively. An experimental approach was adopted with this aim: Sensory and
 quantitative data was collected from learners as they were enrolled in a basic
 course of Python programming. This data was then pre-processed using the method
-described earlier, and together with the quantitative data obtained from users,
-classifiers were trained and validated. The results were then compared and came
-to conclusions regarding the effectiveness of the method. The goal given to
+described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
+classifiers were trained and validated. The results were then compared
+and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
 subjects was to solve as many exercises as they could in a period of two weeks.
@@
 A tutorial was developed to
-obtain the necessary data using Protoboard, a web-based programming learning
-environment (the latest version can be found online at
-http://python.protoboard.org/). Protoboard has the functionality described in
-the proposed method and shown schematically in Figure 1. Users-log in to
+obtain the necessary data using Protoboard, a web-based environment,
+whose latest version can be found online at
+http://python.protoboard.org/, used for learning to
+program. Protoboard has the functionality described in Section
+\ref{sec:method} and shown schematically in Figure 1. % This should be
+                                % a reference and I'm not sure which
+                                % one, if it's now Fig 1 or 2. - JJ
+Users log in to 
 Protoboard using their account credentials from a popular social network
@@
 exercises that the students need to solve in consecutively. Exercises are of
-incremental difficulty. In Figures 2 and 3, the interface of this platform is
+incremental difficulty. In Figures 2 and 3, % 3 is keystroke
+                                % dynamics. Please revise all these
+                                % Figure thing. 
+ the interface of this platform is
 presented. For this experiment Protoboard was configured to allow unlimited

### 2016-11-30 paper.tex
More problems with figures
And paragraph rearranged to avoid repetitions and eliminating non-necessary information.

@@
 Protoboard using their account credentials from a popular social network
-(Facebook).  The web tutorial begins with three introductory videos that explain
+(Facebook), with the advantage that we did not have to keep these
+credentials on site.  The web tutorial begins with three introductory videos that explain
 the fundamentals of programming in Python, and how to solve the programming
@@
 presented. For this experiment Protoboard was configured to allow unlimited
-tries to solve an exercise.
+attempts to solve an exercise.
 
@@
 as a response to three announcements in a special interest group of programming
-students in a social network. Learners were required to login with their
-Facebook credentials, this had the advantage of not requiring the system to
-maintain users’ credentials or profile information as this is kept outside in an
-external service. Unfortunately this decision also kept some prospects from
+students in a social network, but the fact that they needed Facebook kept some prospects from
 volunteering; a subject reported that he normally creates temporary email
-accounts to try new web services. Other requirement was that they where all
-adults, this age is 18 years in Mexico where the experiment was conducted. The
-requirement of the social network account stopped several prospects from
-volunteering, but on the other hand gave researchers access to public
-information about the subjects. All the learners were either current students or
+accounts to try new web services; however, the use of Facebook gave researchers access to public
+information about the subjects:  all the learners were either current students or
 graduates of Computer Systems Engineering. Their ages were in the range of 18 to
-30 years old. Although no experience in software programming was needed, as the
+30 years. Although no experience in software programming was needed, as the
 web tutorial's course is of a very basic level, all the participants were
 required had completed at least one course in software programming. Results from
-the initial survey are shown in Figure 4.
+the initial survey are shown in Figure 4. %Please use reference. Fig 4
+                                %is completed activities. I think this
+                                %should be Fig 5. 
 \begin{table*}[!t]

### 2016-11-30 paper.tex
Comments to conclusions and errors in Abstract

@@
 frustration, distraction, relaxation and engagement. Each of the classification
-models achieved an accuracy of around 75% with a kappa of around 0.45, and, in
-average, J-48 was the better algorithm with an accuracy of 75.97%. Results show
+models achieved an accuracy of around 75\% with a kappa of around 0.45, and, in
+average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
 that the data gathered from the non-obtrusive sensors can successfully be used
@@
 are some possible reasons for this; first the majority of students had
-experience in programming, so learning new one is not that problematic, a second
+prior experience in programming, so learning new one is not that problematic, a second
 reason could be the freedom users had to abandon the activities, perhaps
@@
 Classifiers where implemented using the software Rapid Miner using a feature
-selection module as part of the process. A genetic algorithm was used in the
-user selection module. In the end, the subset for the frustration classifier
+selection module % which one? - JJ
+as part of the process. A genetic algorithm was used in the
+user selection module. % Which one? You really have to be _way_ more
+                       % specific in this conference - JJ
+In the end, the subset for the frustration classifier
 consists of 11 features, and the subset for the boredom classifier consists of
@@
 
+%Start with a reference to the Table (and check numbering) - JJ
 The results obtained with this method are satisfactory, considering a free-text
-was used. The accuracies and kappa coefficients obtained are close to what is
+was used. % don't get this. What free text? A program? - JJ
+The accuracies and kappa coefficients obtained are close to what is
 usually obtained in fixed-text methods, for example, the results presented in
-Epp, Lippold y Mandryk (2007). In the case of the kappa coefficients, it is
+Epp, Lippold y Mandryk (2007). %A reference!!! - JJ
+ In the case of the kappa coefficients, it is
 usual to see values below 0.2 in methods involving free-text. In this case, some
@@
 produce competitive accuracy. In this case the J-48 classifier obtained an
-accuracy of 80.48\% marginally better than the others, and an accepted kappa
-statistic for the kind of problem. Artificial neural network give the highest
+accuracy of 80.48\%, which is marginally better than the others, and an accepted kappa
+statistic for the kind of problem. The artificial neural network % Which one? - JJ
+give the highest
 accuracy in the classification of frustration.
@@
 should focus on novice programmers.
+% But you have to comment on how this addresses the problem. Which
+% should be in the title that is not there yet - JJ
+% You also have to draw a _hard_ conclusion. Is programming using free
+% text better than what? Typing a fixed text?
 

### 2016-11-30 paper.tex
Correcting typo with naïve Bayes

@@
 In total 149 feature vectors were generated and used as input of four different
-classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
+classification algorithms: Na\"ive Bayes, J-48, k-NN, and Artificial Neural
 Networks. These classification models are then compared by their performances
@@
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
-classifier, feed forward neural network, and a naïve Bayes. The output of each
+classifier, feed forward neural network, and a na\"ive Bayes. The output of each
 classifier determined if the student experienced the affective state during the
@@
 Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
-propagation, a naïve Bayes classifier and finally a decision trees algorithm for
+propagation, a na\"ive Bayes classifier and finally a decision trees algorithm for
 rational data (J-48). Details for these algorithms, can be found in the textbook
@@
     \hline
-    Affect          & Naïve Bayes           & J-48                & k-NN              & ANN \\
+    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
     \hline
@@
 % text better than what? Typing a fixed text? 
+% Is 80% effective?
 

### 2016-12-01 paper.tex
Taking care of JJ comments
Please @JJ check the problem section.   For this comment: But the
problem is how to recognize this state so that it can be used to
(select appropriate content| give grades| whatever )

@@
 and there is a common conception that they often have a high
-failure rate \cite{bennedsen2007failure}. There are multiple factors
-involved in this lack of success.
-Jenkins \cite{jenkins2001motivation} argues that there is a deep  relation with the expectations,
-attitudes, and previous experiences of the teaching staff and students. Another
-factor is the nature of the subject that involves the learning of new abstract
-constructs, syntax and tools. % Same guy here? - JJ
-Also groups are often heterogeneous and thus it is
-difficult to design courses that are beneficial for everyone. Many students
-begin to program when they are at their first year of university which
-means they are tackling a totally new topic that does not respond to their habitual
-study approaches. % same reference? You should itemize the claims - JJ
+failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that there 
+are multiple factors involved in this lack of success: 
+\begin{itemize}
+\item There is a deep relation with the expectations, attitudes, 
+and previous experiences of the teaching staff and students. 
+\item The nature of the subject that involves the 
+learning of new abstract constructs, syntax and tools.
+\item Groups are often heterogeneous and thus it is difficult to design courses 
+that are beneficial for everyone. Many students begin to program when they 
+are at their first year of university which means they are tackling a totally
+new topic that does not respond to their habitual study approaches. 
+\end{itemize}
+ % Same guy here? - JJ
+% same reference? You should itemize the claims - JJ
+% Hope its fixed now - MG
 On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
@@
 
-So it is expected that flow/engagement is the ideal affective state in which
+In these works the emotion of flow/engagement is regarded as the ideal affective state in which
 students tend to be most capable of acquiring meaningful information through the
@@
 
-When designing instruction, an objective is to assign learning activities
+In order to promote engagement, instruction is designed with the
+objective of assigning learning activities
 that result challenging to students, but not much as to frustrate them. For
@@
 
+In programming courses, students solve programming problems by implementing their
+proposed solution in a particular language; in order to do this, they must write the code,
+some times compile it, and finally run it to see how it works. 
+While doing this, data can be collected, all the dynamics of actually writing 
+the code in the keyboard, errors, corrections, ellapsed time, number of attempts,
+compiling errors, exceptions, among others. Data mining of this data, could
+bring us a better understanding of the learning process, and also a better model to 
+both select the appropriate content and evaluate of the learner’s competencies. 
 %But the problem is how to recognize this state so that it can be used
 %to (select appropriate content| give grades| whatever ) - JJ
-In this work, a method for the recognition of affective states through keystroke
-and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
+% I highlited the Data Ming task for the track, and that there are other options
+% for Data Analytics   
+In this work, a method for the recognition of affective states through the
+mining of keystroke and mouse dynamics data is proposed. The hypothesis is that students vary the dynamic
 of keystrokes according to their affective state when programming.
+%%
+%%
+
 % since Bosch or whatever has already proved that this is possible (or
@@
 % advance from it - JJ
+
+For this, an interactive web-based programming platform
+is proposed. The platform is designed to collect data from the students'
+interaction for data analysis.
+
  The method

### 2016-12-01 paper.tex
A little more about the proposed approach in the Intro

@@
 bring us a better understanding of the learning process, and also a better model to 
-both select the appropriate content and evaluate of the learner’s competencies. 
+both the selection of the appropriate content and to evaluate the learner’s competencies.
 %But the problem is how to recognize this state so that it can be used
@@
 In this work, a method for the recognition of affective states through the
-mining of keystroke and mouse dynamics data is proposed. The hypothesis is that students vary the dynamic
-of keystrokes according to their affective state when programming.
-%%
-%%
-
+mining of keystroke and mouse dynamics data is proposed. 
+The hypothesis is that students can vary the dynamic of keystrokes according 
+to their affective state when programming. A correlation of the 
+user’s interactions (keyboard and mouse) with an emotional state, has been 
+established in the preliminary work by Zimmermann
+et al. \cite{zimmermann2003affective}, but the work experiments where
+conducted by asking users to write a predefined message. 
 % since Bosch or whatever has already proved that this is possible (or
@@
 % advance from it - JJ
-
 For this, an interactive web-based programming platform
@@
 interaction for data analysis.
-
- The method
+The proposed method
 has three main components: a Javascript library to capture keyboard and mouse

### 2016-12-01 paper.tex
Anonymizing

@@
 based learning environment including the KD and MD functions are open source and
-available as Github repositories at http://git.io/vJlUV also the code for the
-sandbox is in http://git.io/vJlUj. % Anonimyze this, I guess - JJ
+available as Github repositories at http://--------.--/----- also the code for the
+sandbox is in http://--------.--/-----. % Anonimyze this, I guess - JJ
 
@@
 \cite{epp2011identifying} and are shown in Table 1. (The code used for this step is available at:
-https://github.com/amherag/keyboard-mouse-dynamics). %anonymize
+https://------.---/-------/-----------------------). %anonymize
 
@@
 usually obtained in fixed-text methods, for example, the results presented in
-Epp, Lippold y Mandryk (2007). %A reference!!! - JJ
+Epp, Lippold y Mandryk \cite{epp2011identifying}. %A reference!!! - JJ  
+												 %Added	
  In the case of the kappa coefficients, it is
@@
 % use section* for acknowledgment 
-\section*{Acknowledgment} 
-This received funding from Tecnólogico Nacional de México through the project: Técnicas de
-computación inteligente para el secuenciado adaptativo de ejercicios de
-programación en la nube.
+\section*{Acknowledgment} This work received funding from some places, but this is 
+only to keep the space neeed. this is only to keep the space neeed. this is  only to 
+keep the space neeed. only to keep the space neeed. this is only to keep the space neeed. 
+this is  only to only to keep the space neeed. this is only to keep the space neeed. 
+this is  only to  keep the space neeed. only to keep the space neeed. this is only to
+this is  only to
 

### 2016-12-01 paper.tex
Figure and Table numbering corrected

@@
 \caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+\label{tab_features}
     \begin{tabular}{ | l | l || l | l | }
@@
 consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
-\cite{epp2011identifying} and are shown in Table 1. (The code used for this step is available at:
+\cite{epp2011identifying} and are shown in Table \ref{tab_features}. (The code used for this step is available at:
 https://------.---/-------/-----------------------). %anonymize
@@
 A tutorial was developed to
-obtain the necessary data using Protoboard, a web-based environment,
+obtain the necessary data using the proposed platform Protoboard, a web-based environment,
 whose latest version can be found online at 
-http://python.protoboard.org/, used for learning to
-program. Protoboard has the functionality described in Section
-\ref{sec:method} and shown schematically in Figure 1. % This should be
+http://------.----------.---/. 
+The main functionality  used for this experiemnt is
+shown schematically in Figure \ref{fig_process} . % This should be
                                 % a reference and I'm not sure which
                                 % one, if it's now Fig 1 or 2. - JJ
+                                % Corrected -MG
 Users log in to 
-Protoboard using their account credentials from a popular social network
-(Facebook), with the advantage that we did not have to keep these
-credentials on site.  The web tutorial begins with three introductory videos that explain
-the fundamentals of programming in Python, and how to solve the programming
-exercises in the course. What follows after these videos, are 13 programming
-exercises that the students need to solve in consecutively. Exercises are of
-incremental difficulty. In Figures 2 and 3, % 3 is keystroke
+Protoboard  by creaing an account or by using their account credentials from the
+social network Facebook. The web tutorial begins with three introductory videos that explain
+the fundamentals of programming in Python, and how to use the editor to
+execute thier code. What follows after these videos, are 13 programming
+exercises that students need to solve in sequential order. Exercises are of
+incremental difficulty. In Figure \ref{fig_editor}, % 3 is keystroke
                                 % dynamics. Please revise all these
                                 % Figure thing. 
- the interface of this platform is
+the user interface of the programming editor is
 presented. For this experiment Protoboard was configured to allow unlimited
-attempts to solve an exercise.
+attempts to solve each exercise.
 
@@
 accounts to try new web services; however, the use of Facebook gave researchers access to public
-information about the subjects:  all the learners were either current students or
-graduates of Computer Systems Engineering. Their ages were in the range of 18 to
+information about the subjects. Their ages were in the range of 18 to
 30 years. Although no experience in software programming was needed, as the
-web tutorial's course is of a very basic level, all the participants were
-required had completed at least one course in software programming. Results from
-the initial survey are shown in Figure 4. %Please use reference. Fig 4
+web tutorial's course is of a very basic level, participants had
+different levels of experience. Results from
+the initial survey is shown in Table \ref{tab_results}. %Please use reference. Fig 4
                                 %is completed activities. I think this
@@
 \centering
-\caption{The performance of the classifiers.}
+\caption{ Survey question: How many years of programming experience do you have? }
+\label{tab_results}
+    \begin{tabular}{ | c | l | }
+    \hline
+    Years Programming          & Number of Learners \\
+    \hline
+   		0   &  8 \\
+    \hline
+    	1   &  13\\
+    \hline
+    	2-4  & 15\\
+    \hline
+    	5+   & 22\\
+    \hline
+    \end{tabular}
+
+\end{table*}
+
+\begin{table*}[!t]
+\centering
+\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy (kappa) }
+\label{tab_performance}
     \begin{tabular}{ | c | l | l | l | l | }
@@
     \hline
-    Flow/engaged    & 79.00\% (0.11) 0.50   & 80.48\% (0.10) 0.51 & 76.14(0.09) 0.43    & 78.43(0.12) 0.45 \\
+    Flow/engaged    & 79.00\% 0.50 & 80.48\% 0.51 & 76.14\% 0.43 & 78.43\% 0.45 \\
     \hline
-    Relaxation      & 71.10\% (0.09) 0.40   & 72.57\% (0.11) 0.45 & 69.14\%(0.08)  0.37 & 71.24\%(0.07)  0.34\\
+    Relaxation      & 71.10\% 0.40 & 72.57\% 0.45 & 69.14\% 0.37 & 71.24\% 0.34\\
     \hline
-    Distraction     & 70.33\% (0.13)  0.43  & 71.00\% (0.10) 0.43 & 74.00\%(0.06)  0.50 & 72.52\%(0.07)  0.39\\
+    Distraction     & 70.33\% 0.43 & 71.00\% 0.43 & 74.00\% 0.50 & 72.52\% 0.39\\
     \hline
-    Frustration     & 74.71\% (0.09) 0.49   & 73.86\% (0.12) 0.45 & 72.62\%(0.09)  0.42 & 78.0\%(0.09)  0.51 \\
+    Frustration     & 74.71\% 0.49 & 73.86\% 0.45 & 72.62\% 0.42 & 78.00\% 0.50\\
     \hline
-    Boredom         & 74.71\% (0.13) 0.47   & 84.57\% (0.06) 0.59 & 83.81\%(0.06)  0.58 & 76.19\%(0.09)  0.45\\
+    Boredom         & 74.71\% 0.47 & 84.57\% 0.59 & 83.81\% 0.58 & 76.19\% 0.45\\
     \hline
@@
 
-In order to determine what affective states a student was experiencing, an
-Experience Sampling Method (ESM) was used (Csikszentmihalyi y Larson, 1987).
+In order to determine what affective states a student was experiencing, the
+Experience Sampling Method (ESM) \cite{kubey1996experience} was used .
 After the students successfully solve a programming exercise, they are presented
@@
 programming exercises and 22 did not completed any. Out of the total activities
-available (videos, survey and questionnaires) only two completed all. Figure 5
+available (videos, survey and questionnaires) only two completed all. Figure \ref{fig_completed}
 shows the number of exercises and activities completed by each learner.
@@
 consists of 11 features, and the subset for the boredom classifier consists of
-13 features. The performance of the classifiers is shown in Table 1.
+13 features. The performance of the classifiers is shown in Table \ref{tab_performance}.
 

### 2016-12-01 paper.tex
Better description of parameters used.

@@
 interaction for data analysis. Current work is focused on the prediction of
-affective states using non-obtrusive sensors. Specifically, the proposed method uses data
-gathered from both keyboard and mouse dynamics. The data set was generated
+affective states using non-obtrusive sensors. Specifically,  the aim of this research 
+is to evaluate the use of keyboard and mouse dynamics as an appropriate 
+sensory input for an affective recognition
+system. The proposed method uses a data mining approach in data 
+genearated from both keyboard and mouse dynamics of students as they 
+work in programming assignments. The data set was generated
 as students learned the basics of Python programming. Feature vectors 
@@
 languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
-proposed method. % There is no results section so far
+proposed method, and results are discussed in Section \ref{sec:res}  % There is no results section so far
 Finally, we draw some conclusions about our methodology and
@@
 In order to determine what affective states a student was experiencing, the
-Experience Sampling Method (ESM) \cite{kubey1996experience} was used .
-After the students successfully solve a programming exercise, they are presented
+Experience Sampling Method (ESM) \cite{kubey1996experience} was used.
+After the students successfully solve a programming exercise, they were presented
 with an ESM survey that asks what they were feeling during their solving of the
@@
 
-Classifiers where implemented using the software Rapid Miner using a feature
-selection module % which one? - JJ
-as part of the process. A genetic algorithm was used in the
-user selection module. % Which one? You really have to be _way_ more
+As part of the data mining process a genetic algorithm was used for feature
+selection with the following parameters: population size 700, 30 generations with
+tournament selection, tournament size was 25\% of population size. A uniform crossover 
+with 0.5 probability and mutation of 1/number of features, and a minimum feature size of 5.
+  % Which one? You really have to be _way_ more
                        % specific in this conference - JJ
@@
 consists of 11 features, and the subset for the boredom classifier consists of
-13 features. The performance of the classifiers is shown in Table \ref{tab_performance}.
+13 features.
 
-%Start with a reference to the Table (and check numbering) - JJ
-The results obtained with this method are satisfactory, considering a free-text
-was used. % don't get this. What free text? A program? - JJ
+The selected classifiers where implemented using Rapid Miner, with the following patameters:
+
+\section{Results}
+\label{sec:res} 
+In Table \ref{tab_performance} the performance of each of the four classifiers is together 
+with the kappa coefficients. A 10-Fold cross validation was used. 
 The accuracies and kappa coefficients obtained are close to what is
@@
 												 %Added	
+In a fixed-text method, subjects are asked to write specific texts.											 
+The results obtained with this method are satisfactory, considering that learners
+were writing a program, a task comparable with free-text writing, where there
+is no restriction on what has to be writen. % don't get this. What free text? A program? - JJ
 In the case of the kappa coefficients, it is
@@
 Future work will focus on extracting other features from the learners’ interaction
-like behavioral data, as proposed by Blikstein (2011); the data includes the size
+like behavioral data, as proposed by Blikstein \cite{blikstein2011using} ; the data includes the size
 of the program and changes in the code. Other authors also propose treating the use

### 2016-12-02 paper.tex
Last changes

@@
 % Do not put math or special symbols in the title.
-\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}
+\title{ 
+Recognition of Student Engagement and Frustration with Programming Assignments
+using Keystroke and Mouse Dynamics}
 
@@
 challenging to students, but not to the extreme as to frustrate them. When
-selecting an appropiate programing excercise, experienced instructors gauge a
+selecting an appropriate programing exercise, experienced instructors gage a
 student’s affective state and skills to then assign an activity with the  appropriate level
@@
 system. The proposed method uses a data mining approach in data 
-genearated from both keyboard and mouse dynamics of students as they 
+generated from both keyboard and mouse dynamics of students as they 
 work in programming assignments. The data set was generated
@@
 energized focus, full involvement, and enjoyment. Engagement is also a positive
-affect where students are thought to be more involved behaviourally,
+affect where students are thought to be more involved behaviorally,
 intellectually, and emotionally in their learning tasks \cite{bangert2002teacher}.
@@
 that result challenging to students, but not much as to frustrate them. For
-this, experienced instructors gauge a student’s affective state and then assign
+this, experienced instructors gage a student’s affective state and then assign
 an activity with the appropriate level of difficulty. However, recognition of 
@@
 Other sensors such as video cameras can also be considered invasive, since they always transmit the
-identity, appearance and behaviour along with emotional data \cite{picard2001toward}.
+identity, appearance and behavior along with emotional data \cite{picard2001toward}.
 
@@
 While doing this, data can be collected, all the dynamics of actually writing 
-the code in the keyboard, errors, corrections, ellapsed time, number of attempts,
+the code in the keyboard, errors, corrections, elapsed time, number of attempts,
 compiling errors, exceptions, among others. Data mining of this data, could
@@
 The proposed method
-has three main components: a Javascript library to capture keyboard and mouse
+has three main components: a JavaScript library to capture keyboard and mouse
 dynamics from a browser-based editor, a pre-processing step whose output is a
@@
 
-If a relationship could be modelled, ILEs could adapt the instruction using the
+If a relationship could be modeled, ILEs could adapt the instruction using the
 same devices required to input the program to the computer. This could be
@@
 Finally, we draw some conclusions about our methodology and
-experimients.
+experiments.
 
@@
 
-Bosch, D'Mello \& Mills \cite{bosch2013emotions} analysed the relationship between affective states
+Bosch, D'Mello \& Mills \cite{bosch2013emotions} analyzed the relationship between affective states
 and performance of novice programmers when they were learning the basics of
@@
 the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
-behaviours relate to student's achievement within a basic computer science
+behaviors relate to student's achievement within a basic computer science
 course. The authors found that confusion, boredom and engagement in IDE-related
@@
 concurrent with the task. The parameters included respiration, pulse, skin
-conductance level, and corrugator activity. The task assigned to subjects was to
+conductance level, and corrugation activity. The task assigned to subjects was to
 shop on an e-commerce website for office-supplies and finally write a predefined
@@
 use keyboard and mouse but only as future work. There are other studies about
-the behaviour of programmers, which are not directly concerned with affective
+the behavior of programmers, which are not directly concerned with affective
 states but are nevertheless important to their performance. Eye tracking in
@@
 an automated technique to
-assess, analyse and visualize students learning computer programming. Blikstein
-employs different quantitative techniques to extract students’ behaviours and
+assess, analyze and visualize students learning computer programming. Blikstein
+employs different quantitative techniques to extract students’ behaviors and
 categorize them in terms of programming experience.
@@
 which captures every keystroke, mouse movement and mouse button press. Each
-capture of these events records a timestamp in milliseconds (using the method
-getTime() of Javascript's built-in class Date) that describes when the event
+capture of these events records a time-stamp in milliseconds (using the method
+getTime() of JavaScript's built-in class Date) that describes when the event
 occurred. If the event is a keystroke, the script captures what key was
@@
 \centering 
-\includegraphics[width=2.5in]{editorRresult.png} 
+\includegraphics[width=3in]{editorRresult.png} 
 \caption{Web-based editor used for code evaluation.}
@@
 \centering 
-\includegraphics[width=2.5in]{KeyDyn.png} 
+\includegraphics[width=3in]{KeyDyn.png} 
 \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
@@
 obtain the necessary data using the proposed platform Protoboard, a web-based environment,
-whose latest version can be found online at 
+whose latest version can be found on-line at 
 http://------.----------.---/. 
@@
 Users log in to 
-Protoboard  by creaing an account or by using their account credentials from the
+Protoboard  by creating an account or by using their account credentials from the
 social network Facebook. The web tutorial begins with three introductory videos that explain
 the fundamentals of programming in Python, and how to use the editor to
-execute thier code. What follows after these videos, are 13 programming
+execute their code. What follows after these videos, are 13 programming
 exercises that students need to solve in sequential order. Exercises are of
@@
 tournament selection, tournament size was 25\% of population size. A uniform crossover 
-with 0.5 probability and mutation of 1/number of features, and a minimum feature size of 5.
+with 0.5 probability and mutation of 1/39, and a minimum feature size of 5.
   % Which one? You really have to be _way_ more
@@
 
-The selected classifiers where implemented using Rapid Miner, with the following patameters:
+The selected classifiers where implemented using Rapid Miner, with the following parameters:
+For J48 decision tree algorithm a confidence threshold for pruning of 0.25, with a
+minimum of 2 instances per leaf, and 3 folds for reduced error pruning. The feed-forward 
+neural network used a back-propagation learning algorithm with two hidden layer of 20 neurons
+each, with a learning rate and momentum of 0.6, and was trained for 200 generations.
 
@@
 were writing a program, a task comparable with free-text writing, where there
-is no restriction on what has to be writen. % don't get this. What free text? A program? - JJ
+is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
 In the case of the kappa coefficients, it is
@@
 
-Future work will focus on extracting other features from the learners’ interaction
+Future work will focus on extracting other features from the learners'  interaction
 like behavioral data, as proposed by Blikstein \cite{blikstein2011using} ; the data includes the size
@@
 of certain keys as features e.g., the delete key because is manly used for correcting
-sentences. Also a multimodal approach is needed.
+sentences. Also a multi-modal approach is needed.
 
@@
 \section*{Acknowledgment} This work received funding from some places, but this is 
-only to keep the space neeed. this is only to keep the space neeed. this is  only to 
-keep the space neeed. only to keep the space neeed. this is only to keep the space neeed. 
-this is  only to only to keep the space neeed. this is only to keep the space neeed. 
-this is  only to  keep the space neeed. only to keep the space neeed. this is only to
-this is  only to
+only to keep the needed space this is only to keep the space.
 

### 2016-12-07 paper.tex
Fixed a few typos
Please check them out @mariosky just in case you can upload again. Also tagged.

@@
 challenging to students, but not to the extreme as to frustrate them. When
-selecting an appropriate programing exercise, experienced instructors gage a
+selecting an appropriate programming exercise, experienced instructors gauge a
 student’s affective state and skills to then assign an activity with the  appropriate level
@@
 that result challenging to students, but not much as to frustrate them. For
-this, experienced instructors gage a student’s affective state and then assign
+this, experienced instructors gauge a student’s affective state and then assign
 an activity with the appropriate level of difficulty. However, recognition of 
@@
 Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
-affective recognition using nonintrusive sensory information from facial
+affective recognition using non-intrusive sensory information from facial
 expressions, gross body movements, and conversational cues from logs.  The
@@
 keyboard and mouse dynamics as a sensory channel for affective recognition, and
-the few have not been focused on programming. The preliminary work by Zimmermann
+the few have not been focused on programming. The preliminary work by Zimmerman
 et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
@@
 recognition method based on the sensory data provided by the keyboard and mouse
-dynamics generated by a learner as he/she types a programing exercise. The main
+dynamics generated by a learner as he/she types a programming exercise. The main
 components of the method are depicted in Figure \ref{fig_process} A brief explanation of the
@@
 http://------.----------.---/. 
-The main functionality  used for this experiemnt is
+The main functionality  used for this experiment is
 shown schematically in Figure \ref{fig_process} . % This should be

### 2017-02-14 paper.tex
New styles

--- 2016-12-02 paper.tex
@@
-\documentclass[conference]{IEEEtran}
+\documentclass[a4paper]{llncs}
+\usepackage{subcaption} % an alternative package for sub figures
+\usepackage{llncsdoc}
 
@@
 % Do not put math or special symbols in the title.
-\title{ 
-Recognition of Student Engagement and Frustration with Programming Assignments
-using Keystroke and Mouse Dynamics}
+\title{Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning}
 
-% author names and affiliations
-% use a multiple column layout for up to three different
-% affiliations
-\author{\IEEEauthorblockN{Kevin Bacon}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.kev.org/contact.html}
+\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1}, Juan-J. Merelo\inst{2} and Alejandra Mancilla Soto\inst{1}}
+\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico)\\
+    \email{\tt mario@tectijuana.edu.mx}
 \and
-\IEEEauthorblockN{Homer Simpson}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678--2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
+Geneura Team and CITIC, University of Granada (Spain)\\
+    \email{\tt jmerelo@ugr.es}
+}
 
 % make the title area 
+\maketitle
 
-\maketitle
 \begin{abstract}
@@
 
-\bibliographystyle{IEEEtran} 
+\bibliographystyle{splncs03} 
 % argument is your BibTeX string definitions and bibliography database(s) 

### 2017-02-14 paper.tex
First version new format, too long
Reduced to 14 pages without figures. Probably the state of the art can be cut by 75%

@@
 \documentclass[a4paper]{llncs}
-\usepackage{subcaption} % an alternative package for sub figures
+
 \usepackage{llncsdoc}
+
+\usepackage{graphicx,grffile}
+
+\graphicspath{{img/}}
 
@@
 \usepackage{cite}
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-\usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-\graphicspath{{./img/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-
-\fi
 
@@
 \author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1}, Juan-J. Merelo\inst{2} and Alejandra Mancilla Soto\inst{1}}
-\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico)\\
-    \email{\tt mario@tectijuana.edu.mx}
+\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}
 \and
-Geneura Team and CITIC, University of Granada (Spain)\\
+Geneura Team and CITIC, University of Granada (Spain), 
     \email{\tt jmerelo@ugr.es}
@@
 
-Learning to program is often regarded as a difficult task, courses are 
-problem-solving intensive and often students feel frustrated when doing programming
-exercises. Instructors aim to design learning activities that result
-challenging to students, but not to the extreme as to frustrate them. When
-selecting an appropriate programing exercise, experienced instructors gage a
+Learning to program is often regarded as a difficult task. When
+selecting an appropriate programing exercise, experienced instructors gauge a
 student’s affective state and skills to then assign an activity with the  appropriate level
@@
 sensory input for an affective recognition
-system. The proposed method uses a data mining approach in data 
+system. The proposed method uses feature vectors obtained by mining  data 
 generated from both keyboard and mouse dynamics of students as they 
-work in programming assignments. The data set was generated
-as students learned the basics of Python programming. Feature vectors 
-were generated with 39 features representing the dynamics of the interaction. 
-In total 149 feature vectors were generated and used as input of four different
-classification algorithms: Na\"ive Bayes, J-48, k-NN, and Artificial Neural
-Networks. These classification models are then compared by their performances
-when classifying learners into five different affective states: boredom,
-frustration, distraction, relaxation and engagement. Each of the classification
-models achieved an accuracy of around 75\% with a kappa of around 0.45, and, in
-average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
-that the data gathered from the non-obtrusive sensors can successfully be used
+work in basic Python programming assignments, which were used to train different
+classification algorithms to classify learners into five different affective states: boredom,
+frustration, distraction, relaxation and engagement.
+Accuracy achieved was around 75\% with J48 obtaining the best results, proving that data gathered from non-obtrusive sensors can successfully be used
 as another input to classification models in order to predict an individual's
 affective states.
-
 \end{abstract}
@@
 
-\begin{figure}[!t] 
-\centering 
-\includegraphics[width=2.5in]{KMDAffective.png} 
-\caption{Code Execution and Affect Recognition Architecture.}
-\label{fig_process} 
-\end{figure}
+% \begin{figure}[!t] 
+% \centering 
+% \includegraphics[width=2.5in]{KMDAffective.png} 
+% \caption{Code Execution and Affect Recognition Architecture.}
+% \label{fig_process} 
+% \end{figure}
 
@@
 with decision-tree classifiers for the identification of 15 affective states.
-Although their work was based on fixed-text their technique to extract a feature
+Although their work was based on fixed text, their technique to extract a feature
 vector was an inspiration for the proposed method in this work. As for free-text
@@
 
-\begin{table*}[!t]
-\centering
-\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
-\label{tab_features}
-    \begin{tabular}{ | l | l || l | l | }
-    \hline
-    ID          & Description           & ID               & Description  \\
-    \hline
-    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
-    \hline
-    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
-    \hline
-    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
-    \hline
-    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
-    \hline
-    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
-    \hline
-    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
-    \hline
-    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
-    \hline
-    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
-    \hline
-    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
-    \hline
-    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
-    \hline
-    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
-    \hline
-    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
-    \hline
-    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
-    \hline
-    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
-   \hline
-    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
-    \hline
-    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
-    \hline
-    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
-    \hline
-    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
-    \hline
-    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
-    \hline
-    20 & Standard deviation of the previous feature. &   &  \\
-    \hline
-  \end{tabular}
-\end{table*}
+% \begin{table*}[!t]
+% \centering
+% \caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+% \label{tab_features}
+%     \begin{tabular}{ | l | l || l | l | }
+%     \hline
+%     ID          & Description           & ID               & Description  \\
+%     \hline
+%     1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
+%     \hline
+%     2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
+%     \hline
+%     3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
+%     \hline
+%     4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
+%     \hline
+%     5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
+%     \hline
+%     6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
+%     \hline
+%     7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
+%     \hline
+%     8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
+%     \hline
+%     9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
+%     \hline
+%     10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
+%     \hline
+%     11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
+%     \hline
+%     12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
+%     \hline
+%     13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
+%     \hline
+%     14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
+%    \hline
+%     15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
+%     \hline
+%     16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
+%     \hline
+%     17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
+%     \hline
+%     18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
+%     \hline
+%     19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
+%     \hline
+%     20 & Standard deviation of the previous feature. &   &  \\
+%     \hline
+%   \end{tabular}
+% \end{table*}
 
@@
 
-\begin{figure}[!t] 
-\centering 
-\includegraphics[width=3in]{editorRresult.png} 
-\caption{Web-based editor used for code evaluation.}
-\label{fig_editor} 
-\end{figure}
-\begin{figure}[b!] 
-\centering 
-\includegraphics[width=3in]{KeyDyn.png} 
-\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
-\label{fig_KD} 
-\end{figure}
+% \begin{figure}[!t] 
+% \centering 
+% \includegraphics[width=3in]{editorRresult.png} 
+% \caption{Web-based editor used for code evaluation.}
+% \label{fig_editor} 
+% \end{figure}
+
+% \begin{figure}[b!] 
+% \centering 
+% \includegraphics[width=3in]{KeyDyn.png} 
+% \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
+% \label{fig_KD} 
+% \end{figure}
+
 \subsubsection{Preprocessing of the Keystroke and Mouse Data} 

### 2017-02-14 paper.tex
De-anonymizing

@@
 any time.
-
+%
 \begin{figure}[!t] 
 \centering 
-\includegraphics[width=2.5in]{Completed.png} 
-\caption{Completed activities and exercises and by learner in ascending order.}
+\includegraphics[width=2in]{Completed.png} 
+\includegraphics[width=2in]{classDist.png} 
+\caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}
 \label{fig_completed} 
-\end{figure}
-
-\begin{figure}[!t] 
-\centering 
-\includegraphics[width=2.5in]{classDist.png} 
-\caption{Class distribution of emotions as reported by learners.}
-\label{fig_class} 
 \end{figure}
@@
 % use section* for acknowledgment 
-\section*{Acknowledgment} This work received funding from some places, but this is 
-only to keep the needed space this is only to keep the space.
-
-% trigger a \newpage just before the given reference 
-% number - used to balance the columns on the last page 
-% adjust value as needed - may need to be readjusted if 
-% the document is modified later %\IEEEtriggeratref{8} 
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
+\section*{Acknowledgment} 
+This work has been supported in part by: de Ministerio espa\~{n}ol de Econom\'{\i}a y Competitividad under project TIN2014-56494-C4-3-P (UGR-EPHEMECH).
 

### 2017-02-14 paper.tex
Cut it down to 12 pages.
Please @mariosky and @pacastillo have a look. @pacastillo, fork and to a pull request, please.

@@
 languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
-proposed method, and results are discussed in Section \ref{sec:res}  % There is no results section so far
+proposed method and results.
 Finally, we draw some conclusions about our methodology and
@@
 
-Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
-with an agent in a virtual environment. In this work an agent called Steve
-responds emotionally to other agents and interactive users. The agent simulates
-his emotions through different multimedia modes including facial expressions and
-speech. Affective reasoning agents were used for training, by putting students
-in work related situations. Agents could react to the behavior of students, for
-instance if a student was being careless in a task and was in a dangerous
-situation, Steve would show distress or fear. Also in a virtual environment the
-work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
-investigating the affective transitions that occur throughout narrative-centered
-learning experiences. The analysis of affective state transitions in this work
-replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al. 
-\cite{rodrigo2009affective} where
-also engagement/flow dominated self-reported affect. For outdoor environments
-Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
-recognition.  The results about emotion recognition from physiological signals
-achieved a best-case accuracy (86.3\%) for four types of learning emotions.
-
 Bosch, D'Mello \& Mills \cite{bosch2013emotions} analyzed the relationship between affective states
-and performance of novice programmers when they were learning the basics of
-computer programming in the Python language. The results of their study
-indicated that the more common emotions students experienced were engaged
+and performance of novice programmers learning the basic Python. The results of their study
+indicated that the most common emotions students experienced were engaged
 (23\%), confusion (22\%), frustration (14\%), and boredom (12\%). It was useful
 to consider these results, as it presented evidence of what affective states
-need to be targeted in order to obtain less biased data. For example, if a less
-common emotion was chosen, a classifier could opt to classify any feature vector
-as not experiencing such emotion. Nevertheless, the classifier would obtain
-accurate results, although the classifier would be inaccurate at determining if
-a feature vector was actually experiencing the given affective state. Similar to
+need to be targeted in order to obtain less biased data. Similar to
 the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
@@
 course. The authors found that confusion, boredom and engagement in IDE-related
-(on-task) conversation are associated with lower achievement.
+(on-task) conversation are associated with lower achievement. Measuring mood is important, since it may have an impact on programmer’s performance according
+to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
+on the basis of information regarding the programmer’s use of the keyboard and
+mouse, and to integrate them into development environments that can improve
+programmer performance.
 
@@
 et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
-mouse) with an emotional state. Several physiological parameters were measured
-concurrent with the task. The parameters included respiration, pulse, skin
-conductance level, and corrugation activity. The task assigned to subjects was to
-shop on an e-commerce website for office-supplies and finally write a predefined
-message to the shop operator. Subjects where watching videos in order to change
-their emotional state during the experiment. The method was validated using
-self-reported emotions using the self-assessment-manikin (SAM), devised by Lang
-\cite{lang1980behavioral}, designed to assess the dimensions valence, arousal and dominance
-directly by means of three sets of graphical manikins. Results show that the
-film clips were effective in inducing the expected mood changes, but no further
-empirical results are presented. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
+mouse) with an emotional state, meauring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
 uses sensory data based on the time elapsed between each key press, the task
@@
 argue that these results are comparable with other approaches in affective
-computing. They also stress that their methods must be validated further in
-other contexts. Moods may have an impact on programmer’s performance according
-to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
-on the basis of information regarding the programmer’s use of the keyboard and
-mouse, and to integrate them into development environments that can improve
-programmer performance. They briefly describe a further experiment that could
-use keyboard and mouse but only as future work. There are other studies about
-the behavior of programmers, which are not directly concerned with affective
-states but are nevertheless important to their performance. Eye tracking in
-computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
-also used for assessing learner´s comprehension of C++ and Python programs in
-Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed 
-an automated technique to
-assess, analyze and visualize students learning computer programming. Blikstein
-employs different quantitative techniques to extract students’ behaviors and
-categorize them in terms of programming experience.
-
-% \begin{figure}[!t] 
-% \centering 
-% \includegraphics[width=2.5in]{KMDAffective.png} 
-% \caption{Code Execution and Affect Recognition Architecture.}
-% \label{fig_process} 
-% \end{figure}
+computing.
 
@@
 KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
-and engagement based on several classification models.
-
-Regarding Mouse Dynamics (MD), some research has been conducted for the
-identification of affective states, although, as with the case of KD, MD is
-mainly used as a biometric measure for authentication. Salmeron-Majadas, Santos
+and engagement based on several classification models. A similar situation exists in the case of Mouse Dynamics (MD), which is used mainly for authentication; however, Salmeron-Majadas, Santos
 and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
@@
 movements in the axes X and Y are also calculated. Lastly, a final feature is
-added to preprocessing of the data. The web tutorial recorded how many attempts
-a student required before successfully solving an exercise. This number of
-attempts is also included in the feature vector. The final feature vector
+added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
+a student required before successfully solving an exercise. The final feature vector
 consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
-\cite{epp2011identifying} and are shown in Table \ref{tab_features}. (The code used for this step is available at:
-https://------.---/-------/-----------------------). %anonymize
+\cite{epp2011identifying} and are shown in Table \ref{tab_features}.
 
@@
 propagation, a na\"ive Bayes classifier and finally a decision trees algorithm for
-rational data (J-48). Details for these algorithms, can be found in the textbook
-by Tan, Steinbach, \& Kumar \cite{tan2006introduction}. The accuracies obtained by these algorithms
-as well as the parameters used are reported in the next section.
+rational data (J-48) \cite{tan2006introduction}. Experiments and results will be presented next.
 
-\section{Experiment}
+\section{Experiment and results}
 \label{sec:exp}
@@
 web tutorial's course is of a very basic level, participants had
-different levels of experience. Results from
-the initial survey is shown in Table \ref{tab_results}. %Please use reference. Fig 4
-                                %is completed activities. I think this
-                                %should be Fig 5. 
-\begin{table*}[!t]
-\centering
-\caption{ Survey question: How many years of programming experience do you have? }
-\label{tab_results}
-    \begin{tabular}{ | c | l | }
-    \hline
-    Years Programming          & Number of Learners \\
-    \hline
-   		0   &  8 \\
-    \hline
-    	1   &  13\\
-    \hline
-    	2-4  & 15\\
-    \hline
-    	5+   & 22\\
-    \hline
-    \end{tabular}
+different levels of experience, from 0 (8 persons) to more than 5 (22).
 
-\end{table*}
+% \begin{table*}[!t]
+% \centering
+% \caption{ Survey question: How many years of programming experience do you have? }
+% \label{tab_results}
+%     \begin{tabular}{ | c | l | }
+%     \hline
+%     Years Programming          & Number of Learners \\
+%     \hline
+%    		0   &  8 \\
+%     \hline
+%     	1   &  13\\
+%     \hline
+%     	2-4  & 15\\
+%     \hline
+%     	5+   & 22\\
+%     \hline
+%     \end{tabular}
+
+% \end{table*}
 
@@
 \centering
-\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy (kappa) }
+\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy ($\kappa$) }
 \label{tab_performance}
@@
 
-\section{Results}
-\label{sec:res} 
-In Table \ref{tab_performance} the performance of each of the four classifiers is together 
-with the kappa coefficients. A 10-Fold cross validation was used. 
-The accuracies and kappa coefficients obtained are close to what is
+Table \ref{tab_performance} shows the performance of each of the four classifiers is together 
+with the $\kappa$ coefficients. A 10-Fold cross validation was used. 
+The accuracies and $\kappa$ coefficients obtained are close to what is
 usually obtained in fixed-text methods, for example, the results presented in
@@
 is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
-In the case of the kappa coefficients, it is
+In the case of the $\kappa$ coefficients, it is
 usual to see values below 0.2 in methods involving free-text. In this case, some
-values of kappa were close to 0.5, is important to consider the values of kappa
+values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
 in these results because the class distribution is not uniformly distributed. As
@@
 produce competitive accuracy. In this case the J-48 classifier obtained an
-accuracy of 80.48\%, which is marginally better than the others, and an accepted kappa
+accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
 statistic for the kind of problem. The artificial neural network % Which one? - JJ
@@
 satisfactory results. It is usual for a classification method based on free-text
-to obtain an accuracy and kappa below their counterparts of fixed-text, and in this
+to obtain an accuracy and $\kappa$ below their counterparts of fixed-text, and in this
 work the results obtained were similar to those works applied to fixed-text 

### 2017-02-15 paper.tex

@@
 
-\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1}, Juan-J. Merelo\inst{2} and Alejandra Mancilla Soto\inst{1}}
+\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1},
+  Juan-J. Merelo\inst{2}, Pedro Castillo\inst{2} and Alejandra Mancilla Soto\inst{1}}
 \institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}
@@
 Learning to program is often regarded as a difficult task. When
-selecting an appropriate programing exercise, experienced instructors gauge a
-student’s affective state and skills to then assign an activity with the  appropriate level
-of difficulty. This work is focused on the prediction of the affective states
-of programmers with different levels of expertise when learning a new
-programming language. For this, an interactive web-based programming platform
-is proposed. The platform is designed to collect data from the students'
-interaction for data analysis. Current work is focused on the prediction of
-affective states using non-obtrusive sensors. Specifically,  the aim of this research 
-is to evaluate the use of keyboard and mouse dynamics as an appropriate 
-sensory input for an affective recognition
-system. The proposed method uses feature vectors obtained by mining  data 
-generated from both keyboard and mouse dynamics of students as they 
-work in basic Python programming assignments, which were used to train different
-classification algorithms to classify learners into five different affective states: boredom,
-frustration, distraction, relaxation and engagement.
-Accuracy achieved was around 75\% with J48 obtaining the best results, proving that data gathered from non-obtrusive sensors can successfully be used
-as another input to classification models in order to predict an individual's
-affective states.
+selecting an appropriate programming exercise, experienced instructors
+gauge a student’s affective state and skills to then assign an
+activity with the appropriate level of difficulty. This work is
+focused on the prediction of the affective states of programmers with
+different levels of expertise when learning a new programming
+language. For this, an interactive web-based programming platform is
+proposed. The platform is designed to collect data from the students'
+interaction for data analysis. Current work is focused on the
+prediction of affective states using non-obtrusive
+sensors. Specifically, the aim of this research is to evaluate the use
+of keyboard and mouse dynamics as an appropriate sensory input for an
+affective recognition system. The proposed method uses feature vectors
+obtained by mining data generated from both keyboard and mouse
+dynamics of students as they work in basic Python programming
+assignments, which were used to train different classification
+algorithms to classify learners into five different affective states:
+boredom, frustration, distraction, relaxation and engagement. Accuracy
+achieved was around 75\% with J48 obtaining the best results, proving
+that data gathered from non-obtrusive sensors can successfully be used
+as another input to classification models in order to predict an
+individual's affective states.
 \end{abstract}
 
-% no keywords
+\keywords{Affective computing, neural networks, learning analytics}
 
@@
 keyboard and mouse dynamics as a sensory channel for affective recognition, and
-the few have not been focused on programming. The preliminary work by Zimmermann
+the few have not been focused on programming. The preliminary work by Zimmerman
 et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
-mouse) with an emotional state, meauring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
+mouse) with an emotional state, measuring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
 uses sensory data based on the time elapsed between each key press, the task
@@
 recognition method based on the sensory data provided by the keyboard and mouse
-dynamics generated by a learner as he/she types a programing exercise. The main
-components of the method are depicted in Figure \ref{fig_process} A brief explanation of the
+dynamics generated by a learner as he/she types a programming exercise. A brief explanation of the
 whole process is explained next, details come later. The process starts when the
@@
 interpreters or compilers in learner’s computer. The code could even be written
-in a mobile device or any web-enabled device.  The interface of the web-based
-editor is shown in Figure \ref{fig_editor}. Programming exercises are evaluated using unit
+in a mobile device or any web-enabled device. Programming exercises are evaluated using unit
 tests; the results of the evaluation are shown to users. If all tests the
-program is considered to be correct. An example execution is shown in the left
-side image of Figure \ref{fig_editor}. In this example the learner was asked to write a
-function to add two numbers, in Python. The source code for the Protoboard web
+program is considered to be correct. The source code for the Protoboard web
 based learning environment including the KD and MD functions are open source and
-available as Github repositories at http://--------.--/----- also the code for the
+available as Github repositories. %at http://--------.--/----- also the code for the
 sandbox is in http://--------.--/-----. % Anonimyze this, I guess - JJ
@@
 \subsubsection{Preprocessing of the Keystroke and Mouse Data}
+
 The raw data
@@
 consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
-\cite{epp2011identifying} and are shown in Table \ref{tab_features}.
+\cite{epp2011identifying}.
+%and are shown in Table \ref{tab_features}.
 
@@
 obtain the necessary data using the proposed platform Protoboard, a web-based environment,
-whose latest version can be found on-line at 
-http://------.----------.---/. 
-The main functionality  used for this experiemnt is
-shown schematically in Figure \ref{fig_process} . % This should be
+whose latest version has been released under an open source license. 
+% can be found on-line at 
+% http://------.----------.---/. 
+% The main functionality  used for this experiment is
+% shown schematically in Figure \ref{fig_process} . % This should be
                                 % a reference and I'm not sure which
@@
 exercises that students need to solve in sequential order. Exercises are of
-incremental difficulty. In Figure \ref{fig_editor}, % 3 is keystroke
-                                % dynamics. Please revise all these
-                                % Figure thing. 
-the user interface of the programming editor is
-presented. For this experiment Protoboard was configured to allow unlimited
+incremental difficulty. % In Figure \ref{fig_editor}, % 3 is keystroke
+%                                 % dynamics. Please revise all these
+%                                 % Figure thing. 
+% the user interface of the programming editor is
+% presented.
+For this experiment Protoboard was configured to allow unlimited
 attempts to solve each exercise.

### 2017-04-29 paper.tex
Adapted to IJCCI

@@
-\documentclass[a4paper]{llncs}
+\documentclass[a4paper,twoside]{article}
 
-\usepackage{llncsdoc}
+\usepackage{epsfig}
+\usepackage{subfigure}
+\usepackage{calc}
+\usepackage{amssymb}
+\usepackage{amstext}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{multicol}
+\usepackage{pslatex}
+\usepackage{apalike}
+\usepackage{SCITEPRESS}     % Please add other packages that you may need BEFORE the SCITEPRESS.sty package.
+
+\subfigtopskip=0pt
+\subfigcapskip=0pt
+\subfigbottomskip=0pt
 
@@
 
-\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1},
-  Juan-J. Merelo\inst{2}, Pedro Castillo\inst{2} and Alejandra Mancilla Soto\inst{1}}
-\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}
-\and
-Geneura Team and CITIC, University of Granada (Spain), 
-    \email{\tt jmerelo@ugr.es}
+\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
+  Juan-J. Merelo\sup{2}, Pedro Castillo\sup{2} and Alejandra Mancilla Soto\sup{1}}
+\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}}
+\affiliation{Geneura Team and CITIC, University of Granada (Spain), 
+    \email{\tt jmerelo@ugr.es}}
 }
 
-% make the title area 
-\maketitle
+\keywords{Affective computing, neural networks, learning analytics}
 
-\begin{abstract}
-
-Learning to program is often regarded as a difficult task. When
+\abstract{Learning to program is often regarded as a difficult task. When
 selecting an appropriate programming exercise, experienced instructors
@@
 as another input to classification models in order to predict an
-individual's affective states.
-\end{abstract}
+individual's affective states.}
 
-\keywords{Affective computing, neural networks, learning analytics}
+\onecolumn \maketitle \normalsize \vfill
 
@@
 
-% \begin{table*}[!t]
-% \centering
-% \caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
-% \label{tab_features}
-%     \begin{tabular}{ | l | l || l | l | }
-%     \hline
-%     ID          & Description           & ID               & Description  \\
-%     \hline
-%     1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
-%     \hline
-%     2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
-%     \hline
-%     3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
-%     \hline
-%     4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
-%     \hline
-%     5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
-%     \hline
-%     6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
-%     \hline
-%     7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
-%     \hline
-%     8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
-%     \hline
-%     9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
-%     \hline
-%     10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
-%     \hline
-%     11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
-%     \hline
-%     12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
-%     \hline
-%     13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
-%     \hline
-%     14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
-%    \hline
-%     15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
-%     \hline
-%     16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
-%     \hline
-%     17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
-%     \hline
-%     18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
-%     \hline
-%     19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
-%     \hline
-%     20 & Standard deviation of the previous feature. &   &  \\
-%     \hline
-%   \end{tabular}
-% \end{table*}
+\begin{table*}[!t]
+\centering
+\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+\label{tab_features}
+    \begin{tabular}{ | l | l || l | l | }
+    \hline
+    ID          & Description           & ID               & Description  \\
+    \hline
+    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
+    \hline
+    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
+    \hline
+    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
+    \hline
+    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
+    \hline
+    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
+    \hline
+    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
+    \hline
+    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
+    \hline
+    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
+    \hline
+    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
+    \hline
+    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
+    \hline
+    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
+    \hline
+    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
+    \hline
+    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
+    \hline
+    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
+   \hline
+    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
+    \hline
+    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
+    \hline
+    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
+    \hline
+    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
+    \hline
+    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
+    \hline
+    20 & Standard deviation of the previous feature. &   &  \\
+    \hline
+  \end{tabular}
+\end{table*}
 
@@
 
-% \begin{figure}[!t] 
-% \centering 
-% \includegraphics[width=3in]{editorRresult.png} 
-% \caption{Web-based editor used for code evaluation.}
-% \label{fig_editor} 
-% \end{figure}
+\begin{figure*}[h!t] 
+\centering 
+\includegraphics[width=3in]{editorRresult.png} 
+\caption{Web-based editor used for code evaluation.}
+\label{fig_editor} 
+\end{figure*}
 
-% \begin{figure}[b!] 
-% \centering 
-% \includegraphics[width=3in]{KeyDyn.png} 
-% \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
-% \label{fig_KD} 
-% \end{figure}
+\begin{figure*}[hb!] 
+\centering 
+\includegraphics[width=3in]{KeyDyn.png} 
+\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
+\label{fig_KD} 
+\end{figure*}
 
@@
 down, key-up or mouse-move events triggered during an exercise. These events
-have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
+have the dynamic shown in Figure \ref{fig_KD}; for example when typing the word ‘key’ a
 user first presses the letter K and triggers the key-down event, this is
@@
 %
-\begin{figure}[!t] 
+\begin{figure*}[!t] 
 \centering 
-\includegraphics[width=2in]{Completed.png} 
-\includegraphics[width=2in]{classDist.png} 
+\includegraphics{Completed.png} 
+\includegraphics{classDist.png} 
 \caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}
 \label{fig_completed} 
-\end{figure}
+\end{figure*}
 
@@
 % http://------.----------.---/. 
-% The main functionality  used for this experiment is
-% shown schematically in Figure \ref{fig_process} . % This should be
-                                % a reference and I'm not sure which
-                                % one, if it's now Fig 1 or 2. - JJ
-                                % Corrected -MG
+The main functionality  used for this experiment is
+shown schematically in Figure \ref{fig_process} . % Don't know what's
+                                % happened to this figure
 Users log in to 
@@
 exercises that students need to solve in sequential order. Exercises are of
-incremental difficulty. % In Figure \ref{fig_editor}, % 3 is keystroke
-%                                 % dynamics. Please revise all these
-%                                 % Figure thing. 
-% the user interface of the programming editor is
-% presented.
+incremental difficulty.  In Figure \ref{fig_editor}, 
+the user interface of the programming editor is
+presented.
 For this experiment Protoboard was configured to allow unlimited
@@
 
-% \begin{table*}[!t]
-% \centering
-% \caption{ Survey question: How many years of programming experience do you have? }
-% \label{tab_results}
-%     \begin{tabular}{ | c | l | }
-%     \hline
-%     Years Programming          & Number of Learners \\
-%     \hline
-%    		0   &  8 \\
-%     \hline
-%     	1   &  13\\
-%     \hline
-%     	2-4  & 15\\
-%     \hline
-%     	5+   & 22\\
-%     \hline
-%     \end{tabular}
+\begin{table*}[!t]
+\centering
+\caption{ Survey question: How many years of programming experience do you have? }
+\label{tab_results}
+    \begin{tabular}{ | c | l | }
+    \hline
+    Years Programming          & Number of Learners \\
+    \hline
+   		0   &  8 \\
+    \hline
+    	1   &  13\\
+    \hline
+    	2-4  & 15\\
+    \hline
+    	5+   & 22\\
+    \hline
+    \end{tabular}
 
-% \end{table*}
+\end{table*}
 
@@
 
-\bibliographystyle{splncs03} 
+\bibliographystyle{apalike} 
 % argument is your BibTeX string definitions and bibliography database(s) 
 \bibliography{programming}
+
+\vfill
 \end{document}

### 2017-04-29 paper.tex
Retrieved the figure back from original version
And some paragraphs from related work. Also reformatted figures to fit
them in the page. Closes #3.

@@
 expressions, gross body movements, and conversational cues from logs.  The
-subject material of that system consisted in lectures of computer literacy.
+subject material of that system consisted in lectures of computer
+literacy.
+
+Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
+with an agent in a virtual environment. In this work an agent called Steve
+responds emotionally to other agents and interactive users. The agent simulates
+his emotions through different multimedia modes including facial expressions and
+speech. Affective reasoning agents were used for training, by putting students
+in work related situations. Agents could react to the behavior of students, for
+instance if a student was being careless in a task and was in a dangerous
+situation, Steve would show distress or fear. Also in a virtual environment the
+work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
+investigating the affective transitions that occur throughout narrative-centered
+learning experiences. The analysis of affective state transitions in this work
+replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al. 
+\cite{rodrigo2009affective} where
+also engagement/flow dominated self-reported affect. For outdoor environments
+Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
+recognition.  The results about emotion recognition from physiological signals
+achieved a best-case accuracy (86.3\%) for four types of learning emotions.
 
@@
 computing.
+
+They also stress that their methods must be validated further in
+other contexts. Moods may have an impact on programmer’s performance according
+to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
+on the basis of information regarding the programmer’s use of the keyboard and
+mouse, and to integrate them into development environments that can improve
+programmer performance. They briefly describe a further experiment that could
+use keyboard and mouse but only as future work. There are other studies about
+the behaviour of programmers, which are not directly concerned with affective
+states but are nevertheless important to their performance. Eye tracking in
+computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
+also used for assessing learner´s comprehension of C++ and Python programs in
+Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed 
+an automated technique to
+assess, analyse and visualize students learning computer programming. Blikstein
+employs different quantitative techniques to extract students’ behaviours and
+categorize them in terms of programming experience.
+
+\begin{figure*}[h!tbp] 
+\centering 
+\includegraphics[width=0.9\textwidth]{KMDAffective.png} 
+\caption{Code Execution and Affect Recognition Architecture.}
+\label{fig_process} 
+\end{figure*}
 
@@
 \centering 
-\includegraphics[width=3in]{editorRresult.png} 
+\includegraphics[width=0.9\textwidth]{editorRresult.png} 
 \caption{Web-based editor used for code evaluation.}
@@
 \centering 
-\includegraphics[width=3in]{KeyDyn.png} 
+\includegraphics[width=0.9\textwidth]{KeyDyn.png} 
 \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
@@
 \centering 
-\includegraphics{Completed.png} 
-\includegraphics{classDist.png} 
+\includegraphics[width=0.9\textwidth]{Completed.png} 
+\includegraphics[width=0.9\textwidth]{classDist.png} 
 \caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}

### 2017-06-12 paper.tex
Some corrections
Reset column width

@@
 \author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
-  Juan-J. Merelo\sup{2}, Pedro Castillo\sup{2} and Alejandra Mancilla Soto\sup{1}}
+  Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
 \affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}}
@@
 \end{itemize}
- % Same guy here? - JJ
-% same reference? You should itemize the claims - JJ
-% Hope its fixed now - MG
+
 On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
@@
 \centering
-\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+\caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
 \label{tab_features}
-    \begin{tabular}{ | l | l || l | l | }
+    \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
     \hline
@@
 available as Github repositories. %at http://--------.--/----- also the code for the
-sandbox is in http://--------.--/-----. % Anonimyze this, I guess - JJ
+sandbox is in \url{https://github.com/mariosky/sandbox}.
 
@@
 \includegraphics[width=0.9\textwidth]{editorRresult.png} 
-\caption{Web-based editor used for code evaluation.}
+\caption{Web-based editor used for code evaluation. The lower panel
+  indicates (in Spanish) the tests passed successfully and the program
+  output.}
 \label{fig_editor} 
@@
 obtain the necessary data using the proposed platform Protoboard, a web-based environment,
-whose latest version has been released under an open source license. 
-% can be found on-line at 
-% http://------.----------.---/. 
+whose latest version has been released under an open source license,
+and can be found online at \url{https://github.com/mariosky/protoboard}.
 The main functionality  used for this experiment is
@@
 usually obtained in fixed-text methods, for example, the results presented in
-Epp, Lippold y Mandryk \cite{epp2011identifying}. %A reference!!! - JJ  
-												 %Added	
+Epp, Lippold y Mandryk \cite{epp2011identifying}. 	
 In a fixed-text method, subjects are asked to write specific texts.											 
@@
 \section{Conclusion}
+
 The aim of this research was to evaluate the use of keyboard and mouse dynamics 
 as an appropriate sensory input for an affective recognition system. The context
-of use would be learning environment for programmers. In particular for learning
-activities consisting in writing short programs interactively. An experimental
-approach was adopted with this aim. The proposed method in this work obtained
+of use is a learning environment for programmers, in particular for learning
+assignments consisting in writing short programs interactively. An experimental
+approach was adopted with this objective, by effectively using the
+method with real subjects using the environment.
+
+The proposed method in this work obtained
 satisfactory results. It is usual for a classification method based on free-text
-to obtain an accuracy and $\kappa$ below their counterparts of fixed-text, and in this
+to obtain an accuracy and $\kappa$ measure below their counterparts of
+fixed-text; however, in this
 work the results obtained were similar to those works applied to fixed-text 
 dynamics. A possible explanation of this would be the addition of the mouse 
-dynamics features and the additional preprocessing performed on the feature vectors. 
+dynamics features and the additional preprocessing performed on the
+feature vectors.
+
 While these are promising results further experiments are needed. Other experiments
-should focus on novice programmers.
+should focus on novice programmers, but that is left as future
+work. This 
 % But you have to comment on how this addresses the problem. Which
@@
 % Is 80% effective? 
-
-Future work will focus on extracting other features from the learners'  interaction
-like behavioral data, as proposed by Blikstein \cite{blikstein2011using} ; the data includes the size
+future work will focus on extracting other features from the learners' interaction
+such as behavioral data, as proposed by Blikstein
+\cite{blikstein2011using}; the parameters proposed by this author include the size
 of the program and changes in the code. Other authors also propose treating the use
 of certain keys as features e.g., the delete key because is manly used for correcting
-sentences. Also a multi-modal approach is needed.
+sentences. Also a multi-modal approach is needed. Since all these need
+a completely new approach, a new methodology will be developed to
+gather data and take measures.
 

### 2017-06-12 paper.tex
Repositioned URLs

@@
 based learning environment including the KD and MD functions are open source and
-available as Github repositories. %at http://--------.--/----- also the code for the
-sandbox is in \url{https://github.com/mariosky/sandbox}.
+available as Github repositories at \url{https://github.com/mariosky/sandbox}; the code for the
+sandbox is also in \url{https://github.com/mariosky/sandbox}.
 

### 2017-06-12 paper.tex
Añade email de Ale
Few corrections
@jj there is an error with \maketitle

--- 2017-06-12 paper.tex
@@
   Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
-\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}}
+\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
 \affiliation{Geneura Team and CITIC, University of Granada (Spain), 
@@
 selecting an appropriate programming exercise, experienced instructors
-gauge a student’s affective state and skills to then assign an
+gauge a student\’s affective state and skills to then assign an
 activity with the appropriate level of difficulty. This work is
@@
 language. For this, an interactive web-based programming platform is
-proposed. The platform is designed to collect data from the students'
+proposed. The platform is designed to collect data from the students\'
 interaction for data analysis. Current work is focused on the
@@
 as another input to classification models in order to predict an
-individual's affective states.}
+individual\'s affective states.}
 
@@
 Introductory programming courses are
-generally regarded as difficult \cite{robins2003learning,lahtinen2005study} 
+generally regarded as difficult \cite{robins2003learning, lahtinen2005study} 
 and there is a common conception that they often have a high
@@
 difficulties often frustrate students. 
-Jenkins \cite{jenkins2001motivation,jenkins2002difficulty} notes that many
+Jenkins \cite{jenkins2001motivation, jenkins2002difficulty} notes that many
 students expect the course to be difficult and come with the notion that they
@@
 has gone far to understand the role of emotion in learning, both in the process
-and the outcome; for instance, in e-learning  \cite{kort2001affective,rossin2009effects}
-and in programming  \cite{rodrigo2009affective,jenkins2001motivation,
-bosch2013emotions,khan2007mood}.
+and the outcome; for instance, in e-learning  \cite{kort2001affective, rossin2009effects}
+and in programming  \cite{rodrigo2009affective, jenkins2001motivation ,
+bosch2013emotions, khan2007mood}.
 
@@
 that result challenging to students, but not much as to frustrate them. For
-this, experienced instructors gauge a student’s affective state and then assign
+this, experienced instructors gauge a student\'s affective state and then assign
 an activity with the appropriate level of difficulty. However, recognition of 
 emotions is a difficult task, even
-for humans. Instructors use their social intelligence often to recognize
-students’ affective states. In class a good instructor habitually reads the
+for humans. Instructors often use their social intelligence to recognize
+students\'' affective states. In class a good instructor habitually reads the
 faces of students in the classroom to see if they are confused, bored or
@@
 automatic perception of emotions. The recognition and simulation of human
-affects are becoming important fields of study, as many researchers have
-demonstrated affect-aware computers can provide better performance in assisting
+affects are becoming important fields of study, and many researchers have
+demonstrated that affect-aware computers can provide a better performance in assisting
 humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
-problem arises. In order to perform recognition of the users' affective states,
-sensors must be used to gather data. Some sensors rely on physiological readings
+problem arises. Sensors must be used to gather data, in order to perform the 
+recognition of users\' affective states. Some sensors rely on physiological readings
 and must be in physical contact with students. These sensors can be considered
-intrusive or invasive, and can disrupt the student’s learning experience 
-\cite{zhai2008stress,sidney2005integrating,arroyo2009emotion}.
+intrusive or invasive, and can disrupt the student\'s learning experience 
+\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
 Other sensors such as video cameras can also be considered invasive, since they always transmit the
@@
 
-In programming courses, students solve programming problems by implementing their
+In programming courses, students solve programming problems by programming their
 proposed solution in a particular language; in order to do this, they must write the code,
@@
 bring us a better understanding of the learning process, and also a better model to 
-both the selection of the appropriate content and to evaluate the learner’s competencies.
+both the selection of the appropriate content and to evaluate the learner\'s competencies.
 %But the problem is how to recognize this state so that it can be used
@@
 to their affective state when programming. A correlation of the 
-user’s interactions (keyboard and mouse) with an emotional state, has been 
+user\'s interactions (keyboard and mouse) with an emotional state, has been 
 established in the preliminary work by Zimmermann
 et al. \cite{zimmermann2003affective}, but the work experiments where
-conducted by asking users to write a predefined message. 
+conducted by asking users to write a predefined message, not programming
+execercises. 
 % since Bosch or whatever has already proved that this is possible (or
@@
 % advance from it - JJ
-For this, an interactive web-based programming platform
-is proposed. The platform is designed to collect data from the students'
-interaction for data analysis.
-The proposed method
-has three main components: a JavaScript library to capture keyboard and mouse
+In order to test the correlation of keyboard and mouse dynamics and an
+affective state when programming, an interactive web-based 
+programming platform is proposed. The platform is designed to collect
+keyboard and mouse data from the students\' interaction for data analysis.
+The proposed method has three main components: a JavaScript library to capture keyboard and mouse
 dynamics from a browser-based editor, a pre-processing step whose output is a
@@
 exercises using the web based learning environment. Using only the data from the
-student’s keyboard and mouse dynamics six affective states where recognized for
+student\'s keyboard and mouse dynamics six affective states where recognized for
 each of the attempts at solving the exercises. Results obtained from the
@@
 approach using sensory information from facial expressions and postural shifts
-of the learner combined with information about the learner's activity on the
+of the learner combined with information about the learner\'s activity on the
 computer; the learning activity was solving a puzzle on a computer. They report
@@
 the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
-behaviors relate to student's achievement within a basic computer science
+behaviors relate to student\'s achievement within a basic computer science
 course. The authors found that confusion, boredom and engagement in IDE-related
-(on-task) conversation are associated with lower achievement. Measuring mood is important, since it may have an impact on programmer’s performance according
+(on-task) conversation are associated with lower achievement. Measuring mood is important, 
+since it may have an impact on programmer\'s performance according
 to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
-on the basis of information regarding the programmer’s use of the keyboard and
+on the basis of information regarding the programmer\'s use of the keyboard and
 mouse, and to integrate them into development environments that can improve
@@
 They also stress that their methods must be validated further in
-other contexts. Moods may have an impact on programmer’s performance according
+other contexts. Moods may have an impact on programmer\'s performance according
 to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
-on the basis of information regarding the programmer’s use of the keyboard and
+on the basis of information regarding the programmer\'s use of the keyboard and
 mouse, and to integrate them into development environments that can improve
@@
 
-Research works related to Keystroke Dynamics (KD) are carried out either using
-fixed-texts, or free-texts Gunetti y Picardi \cite{gunetti2005keystroke}.
+Research works related to Keystroke Dynamics (KD) have been carried out by using either 
+fixed-texts or free-texts \cite{gunetti2005keystroke}.
 KD performed on fixed-texts
@@
 
-Although the use of Keystroke Dynamics (KD) is found in several research works
+Although the use of KD is found in several research works
 as a biometric measure, its use for identifying affective states is rare in
@@
 capture of these events records a time-stamp in milliseconds (using the method
-getTime() of JavaScript's built-in class Date) that describes when the event
+getTime() of JavaScript\'s built-in class Date) that describes when the event
 occurred. If the event is a keystroke, the script captures what key was
@@
 based learning environment including the KD and MD functions are open source and
-available as Github repositories. %at http://--------.--/----- also the code for the
-sandbox is in \url{https://github.com/mariosky/sandbox}.
+available as Github repositories at \url{https://github.com/amherag/keyboard-mouse-dynamics}; the code for the
+sandbox is also available in \url{https://github.com/mariosky/sandbox}.
 
@@
 \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
-\label{fig_KD} 
+\label{figKD} 
 \end{figure*}
@@
 down, key-up or mouse-move events triggered during an exercise. These events
-have the dynamic shown in Figure \ref{fig_KD}; for example when typing the word ‘key’ a
+have the dynamic shown in Figure \ref{figKD}; for example when typing the word ‘key’ a
 user first presses the letter K and triggers the key-down event, this is
@@
 the previous one. These additional features represent these particular cases,
-and could be meaningful for the estimation of a learner's affective states.
+and could be meaningful for the estimation of a learner\'s affective states.
 Regarding the mouse movements, the average and standard deviation of the
@@
 30 years. Although no experience in software programming was needed, as the
-web tutorial's course is of a very basic level, participants had
+web tutorial\'s course is of a very basic level, participants had
 different levels of experience, from 0 (8 persons) to more than 5 (22).
@@
 
-The participants' interaction generated a total of 142 feature vectors one for each
+The participants\' interaction generated a total of 142 feature vectors one for each
 successfully completed exercise. The affective states reported by learners after
@@
 
-As part of the data mining process a genetic algorithm was used for feature
+As part of the data mining process a canonical genetic algorithm was used for feature
 selection with the following parameters: population size 700, 30 generations with
@@
   % specific in this conference - JJ
+  % donde - Mario
 In the end, the subset for the frustration classifier
@@
 
-Table \ref{tab_performance} shows the performance of each of the four classifiers is together 
+Table \ref{tab_performance} shows the performance of each of the four classifiers together 
 with the $\kappa$ coefficients. A 10-Fold cross validation was used. 
@@
 % Is 80% effective? 
-future work will focus on extracting other features from the learners' interaction
+future work will focus on extracting other features from the learners\' interaction
 such as behavioral data, as proposed by Blikstein
@@
 \section*{Acknowledgment} 
-This work has been supported in part by: de Ministerio espa\~{n}ol de Econom\'{\i}a y Competitividad under project TIN2014-56494-C4-3-P (UGR-EPHEMECH).
+This work has been supported in part by: de Ministerio espa\~{n}ol de Econom\'{\i}a y Competitividad under project TIN2014-56494-C4-3-P (UGR-EPHEMECH) and by CONACYT PEI Project No. 220590.
 

### 2017-06-12 paper.tex
Error found

@@
 selecting an appropriate programming exercise, experienced instructors
-gauge a student\’s affective state and skills to then assign an
+gauge a student\'s affective state and skills to then assign an
 activity with the appropriate level of difficulty. This work is

### 2017-06-12 paper.tex
Minor changes ready for submission

@@
 
-\begin{table*}[!t]
+\begin{table*}[h!tb]
 \centering
@@
 
-\begin{figure*}[h!t] 
+\begin{figure*}[h!tb] 
 \centering 
@@
 
-\begin{figure*}[hb!] 
+\begin{figure*}[h!tb] 
 \centering 
@@
     \hline
-    Flow/engaged    & 79.00\% 0.50 & 80.48\% 0.51 & 76.14\% 0.43 & 78.43\% 0.45 \\
+    Flow/engaged    & 79.00\% (0.50) & 80.48\% (0.51) & 76.14\% (0.43) & 78.43\% (0.45) \\
     \hline
-    Relaxation      & 71.10\% 0.40 & 72.57\% 0.45 & 69.14\% 0.37 & 71.24\% 0.34\\
+    Relaxation      & 71.10\% (0.40) & 72.57\% (0.45) & 69.14\% (0.37) & 71.24\% (0.34)\\
     \hline
-    Distraction     & 70.33\% 0.43 & 71.00\% 0.43 & 74.00\% 0.50 & 72.52\% 0.39\\
+    Distraction     & 70.33\% (0.43) & 71.00\% (0.43) & 74.00\% (0.50) & 72.52\% (0.39)\\
     \hline
-    Frustration     & 74.71\% 0.49 & 73.86\% 0.45 & 72.62\% 0.42 & 78.00\% 0.50\\
+    Frustration     & 74.71\% (0.49) & 73.86\% (0.45) & 72.62\% (0.42) & 78.00\% (0.50)\\
     \hline
-    Boredom         & 74.71\% 0.47 & 84.57\% 0.59 & 83.81\% 0.58 & 76.19\% 0.45\\
+    Boredom         & 74.71\% (0.47) & 84.57\% (0.59) & 83.81\% (0.58) & 76.19\% (0.45)\\
     \hline
@@
 accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
-statistic for the kind of problem. The artificial neural network % Which one? - JJ
-give the highest
+statistic for the kind of problem. The artificial neural network (ANN)
+gives the highest
 accuracy in the classification of frustration.
@@
 dynamics features and the additional preprocessing performed on the
-feature vectors.
+feature vectors. Out of all methods tested, an artificial neural
+network (as provided by RapidMiner) gave the best classification
+results.
 

### 2017-06-12 paper.tex
Specifying ANN
Final draft

@@
 were writing a program, a task comparable with free-text writing, where there
-is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
+is no restriction on what has to be written. 
 In the case of the $\kappa$ coefficients, it is
@@
 dynamics features and the additional preprocessing performed on the
-feature vectors. Out of all methods tested, an artificial neural
-network (as provided by RapidMiner) gave the best classification
+feature vectors. Out of all methods tested, a multilayer perceptron
+trained with backpropagation, the one tagged ANN in RapidMiner,  gave the best classification
 results.

### 2017-09-22 paper.tex
Reduced to 9 pages
With mostly elimination of figures and some reduction to 1 column. Now
the real haircut begins.

@@
 
-\begin{figure*}[h!tbp] 
+\begin{figure}[h!tbp] 
 \centering 
-\includegraphics[width=0.9\textwidth]{KMDAffective.png} 
+\includegraphics[width=0.45\textwidth]{KMDAffective.png} 
 \caption{Code Execution and Affect Recognition Architecture.}
 \label{fig_process} 
-\end{figure*}
+\end{figure}
 
@@
 vector was an inspiration for the proposed method in this work. As for free-text
-KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
+KD, Bixler y D\'Mello \cite{bixler2013detecting} present a method for the identification of boredom
 and engagement based on several classification models. A similar situation exists in the case of Mouse Dynamics (MD), which is used mainly for authentication; however, Salmeron-Majadas, Santos
@@
 
-\begin{table*}[h!tb]
-\centering
-\caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
-\label{tab_features}
-    \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
-    \hline
-    ID          & Description           & ID               & Description  \\
-    \hline
-    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
-    \hline
-    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
-    \hline
-    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
-    \hline
-    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
-    \hline
-    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
-    \hline
-    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
-    \hline
-    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
-    \hline
-    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
-    \hline
-    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
-    \hline
-    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
-    \hline
-    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
-    \hline
-    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
-    \hline
-    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
-    \hline
-    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
-   \hline
-    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
-    \hline
-    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
-    \hline
-    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
-    \hline
-    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
-    \hline
-    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
-    \hline
-    20 & Standard deviation of the previous feature. &   &  \\
-    \hline
-  \end{tabular}
-\end{table*}
+% \begin{table*}[h!tb]
+% \centering
+% \caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+% \label{tab_features}
+%     \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
+%     \hline
+%     ID          & Description           & ID               & Description  \\
+%     \hline
+%     1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
+%     \hline
+%     2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
+%     \hline
+%     3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
+%     \hline
+%     4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
+%     \hline
+%     5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
+%     \hline
+%     6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
+%     \hline
+%     7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
+%     \hline
+%     8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
+%     \hline
+%     9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
+%     \hline
+%     10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
+%     \hline
+%     11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
+%     \hline
+%     12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
+%     \hline
+%     13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
+%     \hline
+%     14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
+%    \hline
+%     15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
+%     \hline
+%     16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
+%     \hline
+%     17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
+%     \hline
+%     18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
+%     \hline
+%     19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
+%     \hline
+%     20 & Standard deviation of the previous feature. &   &  \\
+%     \hline
+%   \end{tabular}
+% \end{table*}
 
-\subsubsection{Capturing the Keystroke and Mouse Data} 
+%\subsubsection{Capturing the Keystroke and Mouse Data} 
 While a student is trying
@@
 
-\begin{figure*}[h!tb] 
-\centering 
-\includegraphics[width=0.9\textwidth]{editorRresult.png} 
-\caption{Web-based editor used for code evaluation. The lower panel
-  indicates (in Spanish) the tests passed successfully and the program
-  output.}
-\label{fig_editor} 
-\end{figure*}
+% \begin{figure}[h!tb] 
+% \centering 
+% \includegraphics[width=0.45\textwidth]{editorRresult.png} 
+% \caption{Web-based editor used for code evaluation. The lower panel
+%   indicates (in Spanish) the tests passed successfully and the program
+%   output.}
+% \label{fig_editor} 
+% \end{figure}
 
-\begin{figure*}[h!tb] 
-\centering 
-\includegraphics[width=0.9\textwidth]{KeyDyn.png} 
-\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
-\label{figKD} 
-\end{figure*}
+% \begin{figure}[h!tb] 
+% \centering 
+% \includegraphics[width=0.45\textwidth]{KeyDyn.png} 
+% \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
+% \label{figKD} 
+% \end{figure}
 
-\subsubsection{Preprocessing of the Keystroke and Mouse Data}
+%\subsubsection{Preprocessing of the Keystroke and Mouse Data}
 
@@
 Basically, this pre-processing consists in measuring the delays between key-
-down, key-up or mouse-move events triggered during an exercise. These events
-have the dynamic shown in Figure \ref{figKD}; for example when typing the word ‘key’ a
+down, key-up or mouse-move events triggered during an exercise.
+% These events have the dynamic shown in Figure \ref{figKD};
+For example when typing the word ‘key’ a
 user first presses the letter K and triggers the key-down event, this is
@@
 %
-\begin{figure*}[!t] 
+\begin{figure}[!t] 
 \centering 
-\includegraphics[width=0.9\textwidth]{Completed.png} 
-\includegraphics[width=0.9\textwidth]{classDist.png} 
-\caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}
+\includegraphics[width=0.45\textwidth]{Completed.png} 
+\includegraphics[width=0.45\textwidth]{classDist.png} 
+\caption{Completed activities and exercises by learner in ascending order (left) and distribution of emotions reported (right).}
 \label{fig_completed} 
-\end{figure*}
+\end{figure}
 
@@
 exercises that students need to solve in sequential order. Exercises are of
-incremental difficulty.  In Figure \ref{fig_editor}, 
-the user interface of the programming editor is
-presented.
+incremental difficulty.
+% In Figure \ref{fig_editor}, 
+% the user interface of the programming editor is
+% presented.
 For this experiment Protoboard was configured to allow unlimited
@@
 
-\begin{table*}[!t]
-\centering
-\caption{ Survey question: How many years of programming experience do you have? }
-\label{tab_results}
-    \begin{tabular}{ | c | l | }
-    \hline
-    Years Programming          & Number of Learners \\
-    \hline
-   		0   &  8 \\
-    \hline
-    	1   &  13\\
-    \hline
-    	2-4  & 15\\
-    \hline
-    	5+   & 22\\
-    \hline
-    \end{tabular}
+% \begin{table}[!t]
+% \centering
+% \caption{ Survey question: How many years of programming experience do you have? }
+% \label{tab_results}
+%     \begin{tabular}{ | c | l | }
+%     \hline
+%     Years Programming          & Number of Learners \\
+%     \hline
+%    		0   &  8 \\
+%     \hline
+%     	1   &  13\\
+%     \hline
+%     	2-4  & 15\\
+%     \hline
+%     	5+   & 22\\
+%     \hline
+%     \end{tabular}
 
-\end{table*}
+% \end{table}
 
@@
 \centering
-\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy ($\kappa$) }
+\caption{Performance of every classifier with 10-fold cross validation. Accuracy ($\kappa$) }
 \label{tab_performance}

### 2017-09-22 paper.tex
Reduced to 8 pages
Closes #4

Mainly through elimination of references and state of the art and
rephrasing using shorter sentences.

@@
 problem arises. Sensors must be used to gather data, in order to perform the 
-recognition of users\' affective states. Some sensors rely on physiological readings
-and must be in physical contact with students. These sensors can be considered
-intrusive or invasive, and can disrupt the student\'s learning experience 
-\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
-Other sensors such as video cameras can also be considered invasive, since they always transmit the
-identity, appearance and behavior along with emotional data \cite{picard2001toward}.
+recognition of users\' affective states.
+%Some sensors rely on physiological readings
+%and must be in physical contact with students.
+% These sensors can be considered
+% intrusive or invasive, and can disrupt the student\'s learning experience 
+% \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
+% Other sensors such as video cameras can also be considered invasive, since they always transmit the
+% identity, appearance and behavior along with emotional data \cite{picard2001toward}.
 
@@
 the code in the keyboard, errors, corrections, elapsed time, number of attempts,
-compiling errors, exceptions, among others. Data mining of this data, could
+compiling errors, exceptions, among others. Data mining this could
 bring us a better understanding of the learning process, and also a better model to 
@@
 classifier determined if the student experienced the affective state during the
-exercise. The proposed method could be used in an ensemble with other sensor
-channels to improve the non-invasive recognition of the students’ affective
-states when they are working on a programming task. In order to determine what
+exercise.
+%
+% The proposed method could be used in an ensemble with other sensor
+% channels to improve the non-invasive recognition of the students’ affective
+% states when they are working on a programming task.
+%
+In order to determine what
 affective states a student was experiencing, an Experience Sampling Method (ESM)
@@
 
-If a relationship could be modeled, ILEs could adapt the instruction using the
-same devices required to input the program to the computer. This could be
-important because writing programs in a computer is an important learning
-activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
-\cite{lahtinen2005study} is reported that students rated ``working alone on programming
-coursework'' as a more useful situation than lectures.
+% If a relationship could be modeled, ILEs could adapt the instruction using the
+% same devices required to input the program to the computer. This could be
+% important because writing programs in a computer is an important learning
+% activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
+% \cite{lahtinen2005study} is reported that students rated ``working alone on programming
+% coursework'' as a more useful situation than lectures.
 
@@
 
-\begin{figure}[h!tbp] 
-\centering 
-\includegraphics[width=0.45\textwidth]{KMDAffective.png} 
-\caption{Code Execution and Affect Recognition Architecture.}
-\label{fig_process} 
-\end{figure}
+% \begin{figure}[h!tbp] 
+% \centering 
+% \includegraphics[width=0.45\textwidth]{KMDAffective.png} 
+% \caption{Code Execution and Affect Recognition Architecture.}
+% \label{fig_process} 
+% \end{figure}
 
-Research works related to Keystroke Dynamics (KD) have been carried out by using either 
-fixed-texts or free-texts \cite{gunetti2005keystroke}.
-KD performed on fixed-texts
-involves the recognition of typing patterns when typing a pre-established fixed-
-length text, e.g., a password. In the other case, free-text KD achieves the
-recognition of typing patterns when typing a text of arbitrary-length, e.g., a
-description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
-most of
-the research regarding KD is done on fixed-text input, the reason being that
-fixed-text KD usually yields better results than free-text KD. Yet, the authors
-of this work share the opinion with Janakiraman, R., and Sim, T., that it would
-be more useful if KD can handle free text as well as fixed text, this is also a
-requirement if the text is a program.
+% Research works related to Keystroke Dynamics (KD) have been carried out by using either 
+% fixed-texts or free-texts \cite{gunetti2005keystroke}.
+% KD performed on fixed-texts
+% involves the recognition of typing patterns when typing a pre-established fixed-
+% length text, e.g., a password. In the other case, free-text KD achieves the
+% recognition of typing patterns when typing a text of arbitrary-length, e.g., a
+% description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
+% most of
+% the research regarding KD is done on fixed-text input, the reason being that
+% fixed-text KD usually yields better results than free-text KD. Yet, the authors
+% of this work share the opinion with Janakiraman, R., and Sim, T., that it would
+% be more useful if KD can handle free text as well as fixed text, this is also a
+% requirement if the text is a program.
 
@@
 % These events have the dynamic shown in Figure \ref{figKD};
-For example when typing the word ‘key’ a
-user first presses the letter K and triggers the key-down event, this is
-indicated with an arrow that changes the state of the key, the time of the event
-is important and also recorded. Only the event data is received from the
-browser, in order to generate feature vectors, the patterns and rhythm users
-have when pressing consecutive keys are captured measuring the delays between
-events, as it is a common practice when dealing with keystroke dynamics. In this
-work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
-time (Ht) is defined as the time between a key-down and a key-up of the same
-key, this would be Ht(K) in the example. Inter key time (It) is defined as the
-time between two consecutive key-down events, this time could be negative is the
-second key is pressed before the first is released. A sequence is defined as a
-list of consecutive keystrokes. In the above example valid sequences could be
-[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
-third is a trigraph. While there can be sequences of any size in a text, working
-only with digraphs and trigraphs is preferred. Even if only digraphs and
-trigraphs are considered, the amount found in free-text is too large and not
-very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
-they propose to use statistical measures to capture the patterns; the same
-strategy is used in this work. The averages and standard deviations are
-calculated from the delays between the events of the digraphs and trigraphs in
-each program.
-
+% For example when typing the word ‘key’ a
+% user first presses the letter K and triggers the key-down event, this is
+% indicated with an arrow that changes the state of the key, the time of the event
+% is important and also recorded. Only the event data is received from the
+% browser, in order to generate feature vectors, the patterns and rhythm users
+% have when pressing consecutive keys are captured measuring the delays between
+% events, as it is a common practice when dealing with keystroke dynamics. In this
+% work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
+% time (Ht) is defined as the time between a key-down and a key-up of the same
+% key, this would be Ht(K) in the example. Inter key time (It) is defined as the
+% time between two consecutive key-down events, this time could be negative is the
+% second key is pressed before the first is released. A sequence is defined as a
+% list of consecutive keystrokes. In the above example valid sequences could be
+% [‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
+% third is a trigraph. While there can be sequences of any size in a text, working
+% only with digraphs and trigraphs is preferred. Even if only digraphs and
+% trigraphs are considered, the amount found in free-text is too large and not
+% very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
+% they propose to use statistical measures to capture the patterns; the same
+% strategy is used in this work. The averages and standard deviations are
+% calculated from the delays between the events of the digraphs and trigraphs in
+% each program.
+%
+%
 To calculate the average and standard deviations of these key presses, the delays
@@
 and mouse dynamics as an appropriate sensory input for an affective recognition
-system. The context of use is a learning environment for programmers. In
-particular for learning activities consisting in writing short programs
-interactively. An experimental approach was adopted with this aim: Sensory and
+system in a learning environment for programmers,
+who interactively write short programs. An experimental approach was adopted with this aim: Sensory and
 quantitative data was collected from learners as they were enrolled in a basic
@@
 and can be found online at \url{https://github.com/mariosky/protoboard}.
-The main functionality  used for this experiment is
-shown schematically in Figure \ref{fig_process} . % Don't know what's
+%The main functionality  used for this experiment is
+%shown schematically in Figure \ref{fig_process} . % Don't know what's
                                 % happened to this figure
@@
 (8\%) and boredom (2\%). This distribution is different from what was reported
-by D’Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
+by D\'Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
 Although Flow/Engagement was
@@
   % donde - Mario
-In the end, the subset for the frustration classifier
-consists of 11 features, and the subset for the boredom classifier consists of
+In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of
 13 features.
@@
 should focus on novice programmers, but that is left as future
-work. This 
+work. %This 
 % But you have to comment on how this addresses the problem. Which
@@
 % Is 80% effective? 
-future work will focus on extracting other features from the learners\' interaction
-such as behavioral data, as proposed by Blikstein
-\cite{blikstein2011using}; the parameters proposed by this author include the size
-of the program and changes in the code. Other authors also propose treating the use
-of certain keys as features e.g., the delete key because is manly used for correcting
-sentences. Also a multi-modal approach is needed. Since all these need
-a completely new approach, a new methodology will be developed to
-gather data and take measures.
+% future work will focus on extracting other features from the learners\' interaction
+% such as behavioral data, as proposed by Blikstein
+% \cite{blikstein2011using}; the parameters proposed by this author include the size
+% of the program and changes in the code. Other authors also propose treating the use
+% of certain keys as features e.g., the delete key because is manly used for correcting
+% sentences. Also a multi-modal approach is needed. Since all these need
+% a completely new approach, a new methodology will be developed to
+% gather data and take measures.
 

### 2017-10-31 paper.tex
Final version
Adds context

@@
 information about the session. In the server, the code is evaluated in an
-external virtual machine that provides a sand box to prevent malicious or
+external virtual machine that provides a sandbox to prevent malicious or
 erroneous code to halt the server. When the result is ready, it is recorded

### 2017-10-31 paper.tex
Changes exercise to assignment
Probably a better term for. Please @mariosky check that it's correct.

@@
 \abstract{Learning to program is often regarded as a difficult task. When
-selecting an appropriate programming exercise, experienced instructors
+selecting an appropriate programming assignment, experienced instructors
 gauge a student\'s affective state and skills to then assign an
@@
 successfully run a challenging program for the first time. They can also become
-bored if they found the exercises too repetitive or too easy. Learning to
+bored if they found the assignments too repetitive or too easy. Learning to
 program is a difficult task, where emotions play a significant role. Research
@@
 An experiment was conducted where students solved a series of programming
-exercises using the web based learning environment. Using only the data from the
+assignments using the web based learning environment. Using only the data from the
 student\'s keyboard and mouse dynamics six affective states where recognized for
-each of the attempts at solving the exercises. Results obtained from the
+each of the attempts at solving the assignments. Results obtained from the
 experiment are promising. The affective states recognized include the most
@@
 classifier determined if the student experienced the affective state during the
-exercise.
+assignment.
 %
@@
 was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
-solve a programming exercise, they are presented with an ESM survey that asks
-what they were feeling during their solving of the exercise.
+solve a programming assignment, they are presented with an ESM survey that asks
+what they were feeling during their solving of the assignment.
 
@@
 recognition method based on the sensory data provided by the keyboard and mouse
-dynamics generated by a learner as he/she types a programming exercise. A brief explanation of the
+dynamics generated by a learner as he/she types a programming assignment. A brief explanation of the
 whole process is explained next, details come later. The process starts when the
@@
 While a student is trying
-to solve an exercise, a script coded in JavaScript is running in the background,
+to solve an assignment, a script coded in JavaScript is running in the background,
 which captures every keystroke, mouse movement and mouse button press. Each
@@
 interpreters or compilers in learner’s computer. The code could even be written
-in a mobile device or any web-enabled device. Programming exercises are evaluated using unit
+in a mobile device or any web-enabled device. Programming assignments are evaluated using unit
 tests; the results of the evaluation are shown to users. If all tests the
@@
 Basically, this pre-processing consists in measuring the delays between key-
-down, key-up or mouse-move events triggered during an exercise.
+down, key-up or mouse-move events triggered during an assignment.
 % These events have the dynamic shown in Figure \ref{figKD};
@@
 added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
-a student required before successfully solving an exercise. The final feature vector
+a student required before successfully solving an assignment. The final feature vector
 consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
@@
 and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
-subjects was to solve as many exercises as they could in a period of two weeks.
+subjects was to solve as many assignments as they could in a period of two weeks.
 There was neither time limit nor a minimum amount of time required for a
-participant while trying to solve the exercises or complete the tutorial. The
+participant while trying to solve the assignments or complete the tutorial. The
 participants were able to stop and resume their interaction with the system at
@@
 \includegraphics[width=0.45\textwidth]{classDist.png} 
-\caption{Completed activities and exercises by learner in ascending order (left) and distribution of emotions reported (right).}
+\caption{Completed activities and assignments by learner in ascending order (left) and distribution of emotions reported (right).}
 \label{fig_completed} 
@@
 execute their code. What follows after these videos, are 13 programming
-exercises that students need to solve in sequential order. Exercises are of
+assignments that students need to solve in sequential order. Assignments are of
 incremental difficulty.
@@
 For this experiment Protoboard was configured to allow unlimited
-attempts to solve each exercise.
+attempts to solve each assignment.
 
@@
 Experience Sampling Method (ESM) \cite{kubey1996experience} was used.
-After the students successfully solve a programming exercise, they were presented
+After the students successfully solve a programming assignment, they were presented
 with an ESM survey that asks what they were feeling during their solving of the
-exercise. A very brief description is given about what to do in this survey,
+assignment. A very brief description is given about what to do in this survey,
 followed by statements the students need to answer according to how they were
@@
 After the two weeks of the experiment, only four learners completed all of the
-programming exercises and 22 did not completed any. Out of the total activities
+programming assignments and 22 did not completed any. Out of the total activities
 available (videos, survey and questionnaires) only two completed all. Figure \ref{fig_completed}
-shows the number of exercises and activities completed by each learner.
+shows the number of assignments and activities completed by each learner.
 
 The participants\' interaction generated a total of 142 feature vectors one for each
-successfully completed exercise. The affective states reported by learners after
-completing each exercise was grouped in three classes: ‘Yes’, ‘Neutral’ and
+successfully completed assignment. The affective states reported by learners after
+completing each assignment was grouped in three classes: ‘Yes’, ‘Neutral’ and
 ‘No’. The class distribution of the emotions reported is shown in Figure 6. This
@@
 frustrated or bored learners simply quit the tutorial. A high number of learners
-(49\%) did not completed any exercise and the once who completed more where also
+(49\%) did not completed any assignment and the once who completed more where also
 the more experienced.

### 2018-04-18 paper.tex
Suggestions for EvoStar paper
After all, we had to cut down a big part of this paper due to space constraints and it being accepted as a short paper. Please @mariosky take a look.
First extension

@@
 recognition of users\' affective states.
-%Some sensors rely on physiological readings
-%and must be in physical contact with students.
-% These sensors can be considered
-% intrusive or invasive, and can disrupt the student\'s learning experience 
-% \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
-% Other sensors such as video cameras can also be considered invasive, since they always transmit the
-% identity, appearance and behavior along with emotional data \cite{picard2001toward}.
+Some sensors rely on physiological readings
+and must be in physical contact with students.
+These sensors can be considered
+intrusive or invasive, and can disrupt the student\'s learning experience
+\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
+ Other sensors such as video cameras can also be considered invasive, since they always transmit the
+ identity, appearance and behavior along with emotional data \cite{picard2001toward}.
 
@@
 both the selection of the appropriate content and to evaluate the learner\'s competencies.
-%But the problem is how to recognize this state so that it can be used
-%to (select appropriate content| give grades| whatever ) - JJ
+% But the problem is how to recognize this state so that it can be used
+% to (select appropriate content| give grades| whatever ) - JJ
 % I highlited the Data Ming task for the track, and that there are other options
@@
 assignment.
-%
-% The proposed method could be used in an ensemble with other sensor
-% channels to improve the non-invasive recognition of the students’ affective
-% states when they are working on a programming task.
-%
+The proposed method could be used in an ensemble with other sensor
+channels to improve the non-invasive recognition of the students’ affective
+states when they are working on a programming task.
 In order to determine what
@@
 
-% If a relationship could be modeled, ILEs could adapt the instruction using the
-% same devices required to input the program to the computer. This could be
-% important because writing programs in a computer is an important learning
-% activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
-% \cite{lahtinen2005study} is reported that students rated ``working alone on programming
-% coursework'' as a more useful situation than lectures.
+If a relationship could be modeled, ILEs could adapt the instruction using the
+same devices required to input the program to the computer. This could be
+important because writing programs in a computer is an important learning
+activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
+\cite{lahtinen2005study} is reported that students rated ``working alone on programming coursework'' as a more useful situation than lectures.
 
@@
 
-% \begin{figure}[h!tbp] 
-% \centering 
-% \includegraphics[width=0.45\textwidth]{KMDAffective.png} 
-% \caption{Code Execution and Affect Recognition Architecture.}
-% \label{fig_process} 
-% \end{figure}
+\begin{figure}[h!tbp]
+\centering
+\includegraphics[width=0.45\textwidth]{KMDAffective.png}
+\caption{Code Execution and Affect Recognition Architecture.}
+\label{fig_process}
+\end{figure}
 
-% Research works related to Keystroke Dynamics (KD) have been carried out by using either 
-% fixed-texts or free-texts \cite{gunetti2005keystroke}.
-% KD performed on fixed-texts
-% involves the recognition of typing patterns when typing a pre-established fixed-
-% length text, e.g., a password. In the other case, free-text KD achieves the
-% recognition of typing patterns when typing a text of arbitrary-length, e.g., a
-% description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke},
-% most of
-% the research regarding KD is done on fixed-text input, the reason being that
-% fixed-text KD usually yields better results than free-text KD. Yet, the authors
-% of this work share the opinion with Janakiraman, R., and Sim, T., that it would
-% be more useful if KD can handle free text as well as fixed text, this is also a
-% requirement if the text is a program.
+Research works related to Keystroke Dynamics (KD) have been carried out by using either
+fixed-texts or free-texts \cite{gunetti2005keystroke}.
+KD performed on fixed-texts
+involves the recognition of typing patterns when typing a pre-established fixed-
+length text, e.g., a password. In the other case, free-text KD achieves the
+recognition of typing patterns when typing a text of arbitrary-length, e.g., a
+description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke}, most of
+the research regarding KD is done on fixed-text input, the reason being that
+fixed-text KD usually yields better results than free-text KD. Yet, the authors
+of this work share the opinion with Janakiraman, R., and Sim, T., that it would
+ be more useful if KD can handle free text as well as fixed text, this is also a
+ requirement if the text is a program.
 
@@
 
-% \begin{table*}[h!tb]
-% \centering
-% \caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
-% \label{tab_features}
-%     \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
-%     \hline
-%     ID          & Description           & ID               & Description  \\
-%     \hline
-%     1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
-%     \hline
-%     2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
-%     \hline
-%     3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
-%     \hline
-%     4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
-%     \hline
-%     5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
-%     \hline
-%     6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
-%     \hline
-%     7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
-%     \hline
-%     8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
-%     \hline
-%     9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
-%     \hline
-%     10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
-%     \hline
-%     11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
-%     \hline
-%     12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
-%     \hline
-%     13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
-%     \hline
-%     14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
-%    \hline
-%     15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
-%     \hline
-%     16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
-%     \hline
-%     17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
-%     \hline
-%     18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
-%     \hline
-%     19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
-%     \hline
-%     20 & Standard deviation of the previous feature. &   &  \\
-%     \hline
-%   \end{tabular}
-% \end{table*}
+\begin{table*}[h!tb]
+\centering
+\caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+\label{tab_features}
+    \begin{tabular}{ | l |p{0.4\textwidth} || l | p{0.4\textwidth} | }
+    \hline
+    ID          & Description           & ID               & Description  \\
+    \hline
+    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
+    \hline
+    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
+    \hline
+    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
+    \hline
+    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
+    \hline
+    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
+    \hline
+    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
+    \hline
+    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
+    \hline
+    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
+    \hline
+    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
+    \hline
+    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
+    \hline
+    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
+    \hline
+    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
+    \hline
+    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
+    \hline
+    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
+   \hline
+    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
+    \hline
+    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
+    \hline
+    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
+    \hline
+    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
+    \hline
+    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
+    \hline
+    20 & Standard deviation of the previous feature. &   &  \\
+    \hline
+  \end{tabular}
+\end{table*}
 
-%\subsubsection{Capturing the Keystroke and Mouse Data} 
+\subsubsection{Capturing the Keystroke and Mouse Data}
 While a student is trying
@@
 
-% \begin{figure}[h!tb] 
-% \centering 
-% \includegraphics[width=0.45\textwidth]{editorRresult.png} 
-% \caption{Web-based editor used for code evaluation. The lower panel
-%   indicates (in Spanish) the tests passed successfully and the program
-%   output.}
-% \label{fig_editor} 
-% \end{figure}
+\begin{figure}[h!tb]
+\centering
+\includegraphics[width=0.45\textwidth]{editorRresult.png}
+\caption{Web-based editor used for code evaluation. The lower panel
+  indicates (in Spanish) the tests passed successfully and the program
+  output.}
+\label{fig_editor}
+\end{figure}
 
-% \begin{figure}[h!tb] 
-% \centering 
-% \includegraphics[width=0.45\textwidth]{KeyDyn.png} 
-% \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
-% \label{figKD} 
-% \end{figure}
+\begin{figure}[h!tb]
+\centering
+\includegraphics[width=0.45\textwidth]{KeyDyn.png}
+\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
+\label{figKD}
+\end{figure}
 
-%\subsubsection{Preprocessing of the Keystroke and Mouse Data}
+\subsubsection{Preprocessing of the Keystroke and Mouse Data}
 
@@
 down, key-up or mouse-move events triggered during an assignment.
-% These events have the dynamic shown in Figure \ref{figKD};
-% For example when typing the word ‘key’ a
-% user first presses the letter K and triggers the key-down event, this is
-% indicated with an arrow that changes the state of the key, the time of the event
-% is important and also recorded. Only the event data is received from the
-% browser, in order to generate feature vectors, the patterns and rhythm users
-% have when pressing consecutive keys are captured measuring the delays between
-% events, as it is a common practice when dealing with keystroke dynamics. In this
-% work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
-% time (Ht) is defined as the time between a key-down and a key-up of the same
-% key, this would be Ht(K) in the example. Inter key time (It) is defined as the
-% time between two consecutive key-down events, this time could be negative is the
-% second key is pressed before the first is released. A sequence is defined as a
-% list of consecutive keystrokes. In the above example valid sequences could be
-% [‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
-% third is a trigraph. While there can be sequences of any size in a text, working
-% only with digraphs and trigraphs is preferred. Even if only digraphs and
-% trigraphs are considered, the amount found in free-text is too large and not
-% very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
-% they propose to use statistical measures to capture the patterns; the same
-% strategy is used in this work. The averages and standard deviations are
-% calculated from the delays between the events of the digraphs and trigraphs in
-% each program.
-%
-%
+These events have the dynamic shown in Figure \ref{figKD};
+For example when typing the word ‘key’ a
+user first presses the letter K and triggers the key-down event, this is
+indicated with an arrow that changes the state of the key, the time of the event
+is important and also recorded. Only the event data is received from the
+browser, in order to generate feature vectors, the patterns and rhythm users
+have when pressing consecutive keys are captured measuring the delays between
+events, as it is a common practice when dealing with keystroke dynamics. In this
+work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
+time (Ht) is defined as the time between a key-down and a key-up of the same
+key, this would be Ht(K) in the example. Inter key time (It) is defined as the
+time between two consecutive key-down events, this time could be negative is the
+second key is pressed before the first is released. A sequence is defined as a
+list of consecutive keystrokes. In the above example valid sequences could be
+[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
+third is a trigraph. While there can be sequences of any size in a text, working
+only with digraphs and trigraphs is preferred. Even if only digraphs and
+trigraphs are considered, the amount found in free-text is too large and not
+very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
+they propose to use statistical measures to capture the patterns; the same
+strategy is used in this work. The averages and standard deviations are
+calculated from the delays between the events of the digraphs and trigraphs in
+each program.
+
 To calculate the average and standard deviations of these key presses, the delays
@@
 incremental difficulty.
-% In Figure \ref{fig_editor}, 
-% the user interface of the programming editor is
-% presented.
+In Figure \ref{fig_editor},
+the user interface of the programming editor is
+presented.
 For this experiment Protoboard was configured to allow unlimited
@@
 
-% \begin{table}[!t]
-% \centering
-% \caption{ Survey question: How many years of programming experience do you have? }
-% \label{tab_results}
-%     \begin{tabular}{ | c | l | }
-%     \hline
-%     Years Programming          & Number of Learners \\
-%     \hline
-%    		0   &  8 \\
-%     \hline
-%     	1   &  13\\
-%     \hline
-%     	2-4  & 15\\
-%     \hline
-%     	5+   & 22\\
-%     \hline
-%     \end{tabular}
+\begin{table}[!t]
+\centering
+\caption{ Survey question: How many years of programming experience do you have? }
+\label{tab_results}
+    \begin{tabular}{ | c | l | }
+    \hline
+    Years Programming          & Number of Learners \\
+    \hline
+   		0   &  8 \\
+    \hline
+    	1   &  13\\
+    \hline
+    	2-4  & 15\\
+    \hline
+    	5+   & 22\\
+    \hline
+    \end{tabular}
 
-% \end{table}
+\end{table}
 
@@
 tournament selection, tournament size was 25\% of population size. A uniform crossover
-with 0.5 probability and mutation of 1/39, and a minimum feature size of 5.
+with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
   % Which one? You really have to be _way_ more
@@
   % donde - Mario
+% with th evolutionary algorithm - JJ
 In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of
 13 features.
+% It's not clear what's the fitness function here or what it really does.
+Maybe we could expand this part of the work, test different evolutionary algorithms and write a paper on feature selection using evolutionary algorithms for EvoStar - JJ
 

### 2018-04-18 paper.tex
New Title
New abstract for #7
@JJ please check

@@
 % Do not put math or special symbols in the title.
-\title{Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning}
+\title{Mining of Keystroke and Mouse Dynamics to Enhance Student Engagement with Programming Assignments}
 
@@
 
-\keywords{Affective computing, neural networks, learning analytics}
+\keywords{Affective computing, neural networks, machine learning}
 
-\abstract{Learning to program is often regarded as a difficult task. When
-selecting an appropriate programming assignment, experienced instructors
-gauge a student\'s affective state and skills to then assign an
-activity with the appropriate level of difficulty. This work is
-focused on the prediction of the affective states of programmers with
-different levels of expertise when learning a new programming
-language. For this, an interactive web-based programming platform is
-proposed. The platform is designed to collect data from the students\'
-interaction for data analysis. Current work is focused on the
-prediction of affective states using non-obtrusive
-sensors. Specifically, the aim of this research is to evaluate the use
-of keyboard and mouse dynamics as an appropriate sensory input for an
-affective recognition system. The proposed method uses feature vectors
-obtained by mining data generated from both keyboard and mouse
-dynamics of students as they work in basic Python programming
-assignments, which were used to train different classification
-algorithms to classify learners into five different affective states:
-boredom, frustration, distraction, relaxation and engagement. Accuracy
-achieved was around 75\% with J48 obtaining the best results, proving
-that data gathered from non-obtrusive sensors can successfully be used
-as another input to classification models in order to predict an
+\abstract{
+When choosing from a collection of programming assignments, instructors must
+gauge the student\'s affective state and skills to select a learning resource
+with the appropriate level of difficulty. This work is
+focused on capturing and pre-processing of data for the evaluation of several
+machine learning techniques aim at classifing the affective
+states of users with different levels of expertise
+when learning a new programming language.
+Currently, web-based platform is applied to collect data from the students\'
+interactions whith programming excersises. The aim of this research is to
+evaluate the use of keyboard and mouse dynamics as an appropriate
+non-obtrusive sensory input for an affective aware system.
+Feature vectors were obtained by pre-processing
+data from keyboard and mouse dynamics captured from students as
+they were coding basic Python programming assignments.
+Feature vectors were used for the classification of five affective states:
+boredom, frustration, distraction, relaxation and engagement.
+The following classification algorithms have been evaluated: k -nearest
+neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
+tree indiction algorithm). The best accuracy was around 75\% and was achieved
+by the J-48 algorithm. Results show that data gathered from ready-available,
+non-obtrusive sensors can be used successfully
+as another input to hyebrid classification models in order to predict an
 individual\'s affective states.}

### 2018-04-18 paper.tex
Introduction upadate
There is also a reference to the conference paper.

@@
 \section{Introduction}
-
-Introductory programming courses are
-generally regarded as difficult \cite{robins2003learning, lahtinen2005study}
+Programming courses are often regarded as difficult \cite{robins2003learning, lahtinen2005study}
 and there is a common conception that they often have a high
-failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that there
-are multiple factors involved in this lack of success:
+failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that this lack of success has multiple factors:
 \begin{itemize}
@@
 
-On the other hand, Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is very
-problem-solving intensive, which means that a high precision is required because even the
-slightest perturbation can render a program totally worthless. These
-difficulties often frustrate students.
-Jenkins \cite{jenkins2001motivation, jenkins2002difficulty} notes that many
-students expect the course to be difficult and come with the notion that they
-will have to struggle, others may have a stereotyped image of a programmer,
-these beliefs can negatively affect their initial motivation. During the course,
-novice programmers experience many emotions, for instance frustration and
+Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is
+very problem-solving intensive, which means that a high precision is required
+because even the slightest perturbation can render a program totally worthless.
+These difficulties often frustrate novice students. Instructors move to explain
+more advanced subjects, while a group of students is still struggling with
+basic syntax.
+Again Jenkins \cite{jenkins2001motivation, jenkins2002difficulty} notes that
+many students expect the course to be difficult and come with the notion that
+they will have to struggle, others may have a stereotyped image of a programmer,
+these beliefs can negatively affect their initial motivation and expectations.
+During the course, novice programmers experience many emotions,
+for instance frustration and
 confusion when they cannot find a bug in a program, but also joy when they
 successfully run a challenging program for the first time. They can also become
-bored if they found the assignments too repetitive or too easy. Learning to
-program is a difficult task, where emotions play a significant role. Research
-has gone far to understand the role of emotion in learning, both in the process
-and the outcome; for instance, in e-learning  \cite{kort2001affective, rossin2009effects}
+bored if they found the assignments too repetitive or too easy.
+
+Learning to program is a difficult task, where emotions play a significant
+role. Research has gone far to understand the role of emotion in learning, both
+in the process and the outcome; for instance, in e-learning
+\cite{kort2001affective, rossin2009effects}
 and in programming  \cite{rodrigo2009affective, jenkins2001motivation ,
-bosch2013emotions, khan2007mood}.
-
-In these works the emotion of flow/engagement is regarded as the ideal affective state in which
+bosch2013emotions, khan2007mood}. In these works the emotion of flow/engagement
+is regarded as the ideal affective state in which
 students tend to be most capable of acquiring meaningful information through the
 learning process. Flow is defined by Csikszentmihalyi \cite{csikszentmihalyi1990flow}
-as a mental state in which a person performing an activity is fully immersed in a feeling of
-energized focus, full involvement, and enjoyment. Engagement is also a positive
-affect where students are thought to be more involved behaviorally,
-intellectually, and emotionally in their learning tasks \cite{bangert2002teacher}.
+as a mental state in which a person performing an activity is fully immersed in a feeling of energized focus, full involvement, and enjoyment. Engagement is
+also a positive affect where students are thought to be more involved
+behaviorally, intellectually, and emotionally in their learning tasks
+\cite{bangert2002teacher}.
 
-In order to promote engagement, instruction is designed with the
-objective of assigning learning activities
-that result challenging to students, but not much as to frustrate them. For
-this, experienced instructors gauge a student\'s affective state and then assign
-an activity with the appropriate level of difficulty. However, recognition of
-emotions is a difficult task, even
+In order to promote engagement, sometimes instruction is designed with the
+objective of creating learning activities that are challenging to students, but
+not much as to frustrate them. Engagement is not only achieved through the
+design of single learning activities but also on the appropiate selection and
+sequence in which they are presented.
+For this, experienced instructors gauge a student\'s affective state and then
+assign an activity with the appropriate level of difficulty. However,
+recognition of emotions is a difficult task, even
 for humans. Instructors often use their social intelligence to recognize
-students\'' affective states. In class a good instructor habitually reads the
+students\'' affective states. In class, an instructor habitually reads the
 faces of students in the classroom to see if they are confused, bored or
@@
 
-In programming courses, students solve programming problems by programming their
-proposed solution in a particular language; in order to do this, they must write the code,
-some times compile it, and finally run it to see how it works.
-While doing this, data can be collected, all the dynamics of actually writing
-the code in the keyboard, errors, corrections, elapsed time, number of attempts,
-compiling errors, exceptions, among others. Data mining this could
-bring us a better understanding of the learning process, and also a better model to
-both the selection of the appropriate content and to evaluate the learner\'s competencies.
-% But the problem is how to recognize this state so that it can be used
+In programming courses, students solve programming problems by implementing
+their proposed solution in a particular language; in order to do this, they
+must write the code,
+some times compile it, to finally run it to see how it works. This cycle is
+often repeated many times, and studens can change their affective state
+accordingly.
+Data can be collected, while students are working on their assignments.
+All the dynamics of actually writing the code using the keyboard for instance
+the syntax errors they received from the compiler or interpreter, the
+corrections they tried, the elapsed time of each step, the number of attempts,
+compiling errors, exceptions, among others. Data mining of all these data could
+bring us a better understanding of the learning process, and also a better
+model to both the selection of the appropriate content and to evaluate the
+learner\'s competencies. But the problem that remains is to how to recognize
+the student's affective state so that it can be used by an adaptive system.
 % to (select appropriate content| give grades| whatever ) - JJ
@@
 affective state when programming, an interactive web-based
-programming platform is proposed. The platform is designed to collect
-keyboard and mouse data from the students\' interaction for data analysis.
-The proposed method has three main components: a JavaScript library to capture keyboard and mouse
-dynamics from a browser-based editor, a pre-processing step whose output is a
-feature vector that is sent to the third component, a classification algorithm.
+programming platform has been proposed \cite{ijcci17}.
+The platform is designed to collect keyboard and mouse data from the students\' interaction for data analysis.
+The proposed method has three main components: a JavaScript library to capture keyboard and mouse dynamics from a browser-based editor, a pre-processing step whose output is a feature vector that is sent to a third component, the classification algorithm.
+
 An experiment was conducted where students solved a series of programming
-assignments using the web based learning environment. Using only the data from the
-student\'s keyboard and mouse dynamics six affective states where recognized for
-each of the attempts at solving the assignments. Results obtained from the
-experiment are promising. The affective states recognized include the most
-common in novice programmers according to the study of Bosch, D'Mello \& Mills
-\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
+assignments using the web based learning environment.
+Using only the data from the student\'s keyboard and mouse dynamics six affective states where recognized for each of the attempts at solving the assignments.
+Results obtained from the experiment are promising.
+The affective states recognized include the most common in novice programmers according to the study of Bosch, D'Mello \& Mills \cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration.
+For this experiment
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour

### 2018-04-18 paper.tex
Other changes

@@
 \section{Introduction}
-Programming courses are often regarded as difficult \cite{robins2003learning, lahtinen2005study}
+Programming courses are often regarded as difficult
+\cite{robins2003learning, lahtinen2005study}
 and there is a common conception that they often have a high
-failure rate \cite{bennedsen2007failure}. Jenkins \cite{jenkins2001motivation} argues that this lack of success has multiple factors:
+failure rate \cite{bennedsen2007failure}.
+Jenkins \cite{jenkins2001motivation} argues that this lack of success has
+multiple factors:
 \begin{itemize}
@@
 affects are becoming important fields of study, and many researchers have
-demonstrated that affect-aware computers can provide a better performance in assisting
-humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
+demonstrated that affect-aware computers can provide a better performance in
+assisting humans \cite{picard2001toward}. When ILEs embrace these methodologies
+and techniques a
 problem arises. Sensors must be used to gather data, in order to perform the
@@
 \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
- Other sensors such as video cameras can also be considered invasive, since they always transmit the
- identity, appearance and behavior along with emotional data \cite{picard2001toward}.
+Other sensors such as video cameras can also be considered invasive, since
+they always transmit the user's identity, appearance and behavior along with
+the emotional data \cite{picard2001toward}.
 
 In programming courses, students solve programming problems by implementing
-their proposed solution in a particular language; in order to do this, they
-must write the code,
-some times compile it, to finally run it to see how it works. This cycle is
-often repeated many times, and studens can change their affective state
-accordingly.
-Data can be collected, while students are working on their assignments.
-All the dynamics of actually writing the code using the keyboard for instance
+their solution in a particular language; in order to do this, they
+must type the code, some times compile it, and finally run it to see how or if
+it works. This cycle can be repeated many times, as they are in this cycle,
+studens often change their affective state according to their performence.
+Data can be collected while students are working on their assignments.
+All the dynamics of actually typing the code and for instance
 the syntax errors they received from the compiler or interpreter, the
 corrections they tried, the elapsed time of each step, the number of attempts,
-compiling errors, exceptions, among others. Data mining of all these data could
+compiling errors, exceptions, among others. Data mining of all this data could
 bring us a better understanding of the learning process, and also a better
@@
 recognition method based on the sensory data provided by the keyboard and mouse
-dynamics generated by a learner as he/she types a programming assignment. A brief explanation of the
-whole process is explained next, details come later. The process starts when the
+dynamics generated by a learner as he or she types a programming assignment.
+A detailed explanation of the process outlined in a previous work \cite{ijcci17} is explained next.
+The process starts when the
 learner begins to type a program in a browser-based editor. As she types or
-moves the mouse the dynamics are recorded. When the learner submits the code to
-evaluation, the request includes the sensory data along with the code and
-information about the session. In the server, the code is evaluated in an
-external virtual machine that provides a sandbox to prevent malicious or
-erroneous code to halt the server. When the result is ready, it is recorded
-along with the sensory data and sent to the preprocessing module. The output is
-a feature vector, ready for classification. A previously trained classifier is
-responsible for the classification, which outputs the predicted affective state.
-The method does not consider other sensory data, but it could be integrated with
-other sensory inputs in a multi-modal approach.  Each of the steps is explained
-in detail next.
+moves the mouse all the data of the dynamics is recorded on the client.
+When the learner submits the code to evaluation, the HTTP request includes,
+in a JSON string, all the sensory data along with the code and
+information about the current session.
+In the web server, the code is evaluated by an
+external container that provides a virtual sandbox in which to execute the
+program. The sandbox is used to prevent malicious or
+erroneous code to halt the server. When a container is halted, it is simply
+removed and a new container is created.
+When the result is ready, it is recorded
+along with the sensory data and sent to the preprocessing module that it will
+be explained next. The output is a feature vector, ready for classification. A
+previously trained classifier is responsible for the classification, which
+outputs the predicted affective state.
+
+Currently, the method does not consider other sensory data,
+but it could be integrated with other sensory inputs in a multi-modal
+approach.  Each of the steps is explained in detail next.
 
@@
 \subsubsection{Capturing the Keystroke and Mouse Data}
-While a student is trying
-to solve an assignment, a script coded in JavaScript is running in the background,
-which captures every keystroke, mouse movement and mouse button press. Each
-capture of these events records a time-stamp in milliseconds (using the method
+As it was explained before, as a student is trying to solve an assignment,
+a script coded in JavaScript runs in the background,
+capturing every keystroke, mouse movement and mouse button pressed. Each record
+of these events includes a time-stamp in milliseconds (using the method
 getTime() of JavaScript\'s built-in class Date) that describes when the event
@@
 interpreters or compilers in learner’s computer. The code could even be written
-in a mobile device or any web-enabled device. Programming assignments are evaluated using unit
-tests; the results of the evaluation are shown to users. If all tests the
-program is considered to be correct. The source code for the Protoboard web
-based learning environment including the KD and MD functions are open source and
-available as Github repositories at \url{https://github.com/amherag/keyboard-mouse-dynamics}; the code for the
+in a mobile device or any web-enabled device. Programming assignments are
+evaluated using unit tests; the results of the evaluation are shown to users.
+If all tests the program is considered to be correct. The source code for the
+Protoboard web based learning environment including the KD and MD functions are
+open source and available as Github repositories
+at\url{https://github.com/amherag/keyboard-mouse-dynamics}; the code for the
 sandbox is also available in \url{https://github.com/mariosky/sandbox}.
@@
 
-The raw data
-obtained from the script needs to be preprocessed to obtain a feature vector.
+The raw data obtained from the script needs to be preprocessed to obtain a feature vector.
 Basically, this pre-processing consists in measuring the delays between key-
@@
 
-The selected classifiers where implemented using Rapid Miner, with the following parameters:
-For J48 decision tree algorithm a confidence threshold for pruning of 0.25, with a
-minimum of 2 instances per leaf, and 3 folds for reduced error pruning. The feed-forward
-neural network used a back-propagation learning algorithm with two hidden layer of 20 neurons
-each, with a learning rate and momentum of 0.6, and was trained for 200 generations.
+The selected classifiers where implemented using Rapid Miner, with the
+following parameters: For J48 decision tree algorithm a confidence threshold
+for pruning of 0.25, with a minimum of 2 instances per leaf, and 3 folds for
+reduced error pruning. The feed-forward neural network used a back-propagation
+learning algorithm with two hidden layer of 20 neurons each, with a learning
+rate and momentum of 0.6, and was trained for 200 generations.
 
-Table \ref{tab_performance} shows the performance of each of the four classifiers together
+Table \ref{tab_performance} shows the performance of each of the four
+classifiers together
 with the $\kappa$ coefficients. A 10-Fold cross validation was used.
@@
 In a fixed-text method, subjects are asked to write specific texts.
-The results obtained with this method are satisfactory, considering that learners
-were writing a program, a task comparable with free-text writing, where there
-is no restriction on what has to be written.
+The results obtained with this method are satisfactory, considering that
+learners were writing a program, a task comparable with free-text writing,
+where there is no restriction on what has to be written.
 In the case of the $\kappa$ coefficients, it is

### 2018-04-19 paper.tex
Revises abstract please @mariosky take a look

@@
 % Do not put math or special symbols in the title.
-\title{Mining of Keystroke and Mouse Dynamics to Enhance Student Engagement with Programming Assignments}
+\title{Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments}
 
@@
   Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
-\affiliation{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
+\affiliation{Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
 \affiliation{Geneura Team and CITIC, University of Granada (Spain),
@@
 
-\keywords{Affective computing, neural networks, machine learning}
+\keywords{Affective computing, neural networks, machine learning, learning analytics}
 
-\abstract{
-When choosing from a collection of programming assignments, instructors must
-gauge the student\'s affective state and skills to select a learning resource
-with the appropriate level of difficulty. This work is
-focused on capturing and pre-processing of data for the evaluation of several
-machine learning techniques aim at classifing the affective
+\abstract{The aim of this research is to 
+evaluate the use of keyboard and mouse dynamics as an appropriate
+non-obtrusive sensory input for an affective aware system, and is
+motivated by the lack of tools and methodologies for taking into
+account the affective state of the student in learning
+environments. In an ideal situation, when instructors have to choose
+from 
+  a collection of programming assignments, they should consider the
+  student\'s affective state and skills in order to select a learning resource
+  with the appropriate level of difficulty. However, neither the data
+  or the ability to process it are present in current learning
+  management systems. This work tries to address this problem, by 
+  focused on the capture and pre-processing of data that is going to
+  be fed to several machine learning techniques with the objective of classifying the affective
 states of users with different levels of expertise
-when learning a new programming language.
-Currently, web-based platform is applied to collect data from the students\'
-interactions whith programming excersises. The aim of this research is to
-evaluate the use of keyboard and mouse dynamics as an appropriate
-non-obtrusive sensory input for an affective aware system.
-Feature vectors were obtained by pre-processing
-data from keyboard and mouse dynamics captured from students as
-they were coding basic Python programming assignments.
-Feature vectors were used for the classification of five affective states:
+when learning a new programming language. We capture student data from
+a web-based platform, a learning management system where students
+interact whith programming excersises. We introduce in this paper a
+series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
+they were coding basic Python programming assignments  into feature
+vector, that have been later used for the classification into five affective states:
 boredom, frustration, distraction, relaxation and engagement.
@@
 neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
-tree indiction algorithm). The best accuracy was around 75\% and was achieved
+tree induction algorithm). The best accuracy was around 75\% and was achieved
 by the J-48 algorithm. Results show that data gathered from ready-available,
@@
 as another input to hyebrid classification models in order to predict an
-individual\'s affective states.}
+individual\'s affective states, and these in turn can be used to make
+students more engaged and thus learn more efficiently.}
 

### 2018-04-19 paper.tex
Improves bibliography
Extends introduction

@@
 Programming courses are often regarded as difficult
-\cite{robins2003learning, lahtinen2005study}
-and there is a common conception that they often have a high
+\cite{robins2003learning, lahtinen2005study,munson2018models}
+and there is a common conception, backed by data, that they often have a high
 failure rate \cite{bennedsen2007failure}.
-Jenkins \cite{jenkins2001motivation} argues that this lack of success has
+Jenkins \cite{jenkins2001motivation} argues that this high rate of failure has
 multiple factors:
@@
 very problem-solving intensive, which means that a high precision is required
-because even the slightest perturbation can render a program totally worthless.
-These difficulties often frustrate novice students. Instructors move to explain
-more advanced subjects, while a group of students is still struggling with
+since even the slightest deviation from syntax can render a program
+totally worthless. These are hurdles that often frustrate novice
+students. This is aggravated by the fact that the time devoted to a
+particular topic is fixed, and when it is exhausted, instructors move to explain
+more advanced subjects, while a group of students might still be struggling with
 basic syntax.
@@
 many students expect the course to be difficult and come with the notion that
-they will have to struggle, others may have a stereotyped image of a programmer,
-these beliefs can negatively affect their initial motivation and expectations.
-During the course, novice programmers experience many emotions,
-for instance frustration and
-confusion when they cannot find a bug in a program, but also joy when they
+they will have to struggle; others may have a stereotyped image of a programmer,
+and these beliefs can negatively affect their initial motivation and expectations.
+During the duration of the course and its delivery, novice programmers
+have been proved to experience  emotions such as frustration and
+confusion when a bug in a program eludes them, but also joy when they
 successfully run a challenging program for the first time. They can also become
@@
 
-Learning to program is a difficult task, where emotions play a significant
+Learning to program is a difficult task, and similarly to other
+learning tasks, emotions play a significant
 role. Research has gone far to understand the role of emotion in learning, both

### 2018-04-19 paper.tex
Revised and extended abstract, closes #7

@@
 
-\abstract{The aim of this research is to 
+\abstract{The aim of the experiments described in this paper is to 
 evaluate the use of keyboard and mouse dynamics as an appropriate
-non-obtrusive sensory input for an affective aware system, and is
-motivated by the lack of tools and methodologies for taking into
-account the affective state of the student in learning
+non-obtrusive sensory input for an system that is sensitive to the
+affective state of its user. Our motivation for starting this research
+line has been the lack of tools and methodologies for taking into
+account this affective state in learning
 environments. In an ideal situation, when instructors have to choose

### 2018-04-19 paper.tex
Extend citation of the paper and explain differences.
Refs #8. There are other references, will have to check them too later
on. Keep it open meanwhile.

@@
 environments. In an ideal situation, when instructors have to choose
-from 
-  a collection of programming assignments, they should consider the
+from  a collection of programming assignments, they should consider the
   student\'s affective state and skills in order to select a learning resource
@@
 the student's affective state so that it can be used by an adaptive system.
-% to (select appropriate content| give grades| whatever ) - JJ
-% I highlited the Data Ming task for the track, and that there are other options
-% for Data Analytics
+
 In this work, a method for the recognition of affective states through the
@@
 affective state when programming, an interactive web-based
-programming platform has been proposed \cite{ijcci17}.
-The platform is designed to collect keyboard and mouse data from the students\' interaction for data analysis.
-The proposed method has three main components: a JavaScript library to capture keyboard and mouse dynamics from a browser-based editor, a pre-processing step whose output is a feature vector that is sent to a third component, the classification algorithm.
+programming platform was initially proposed by us in
+\cite{ijcci17}. In this paper we complete the background and
+motivation for this research, and update the state of the art, as well
+as justify better the methdology used and the way data has been
+collected. The platform we presented in that paper originally was designed to collect keyboard and mouse data
+from the students\' interaction for data analysis. The proposed method
+has three main components: a JavaScript library to capture keyboard
+and mouse dynamics from a browser-based editor, a pre-processing step
+whose output is a feature vector that is sent to a third component,
+the classification algorithm.
 

### 2018-04-19 paper.tex
Revised introduction

@@
 
-In order to promote engagement, sometimes instruction is designed with the
+In order to promote engagement, sometimes instruction material is designed with the
 objective of creating learning activities that are challenging to students, but
-not much as to frustrate them. Engagement is not only achieved through the
+not so much as to frustrate them. Engagement is not only achieved through the
 design of single learning activities but also on the appropiate selection and
@@
 for humans. Instructors often use their social intelligence to recognize
-students\'' affective states. In class, an instructor habitually reads the
+students\'' affective states. During presential  classes, an instructor habitually reads the
 faces of students in the classroom to see if they are confused, bored or
-engaged; then decides what to do next. Interactive Learning Environments (ILEs)
-can be enhanced if they can offer an
-automatic perception of emotions. The recognition and simulation of human
-affects are becoming important fields of study, and many researchers have
+engaged and then decides what to do next. However, that is impossible
+in asynchronous and virtual learning environments, that is why there
+have been several ways of including automatic perception of emotions
+in Interactive Learning Environments (ILEs) and Learning Management
+Systems \cite{gottardo2017affective} and even in Integrated
+Development Environments \cite{hundhausen2017ide}.
+
+But this of course presupposes that the affective state of the student is available to the
+system. This detection and recognition can be done in different
+ways. In the wider context of the detection of emotions and affective
+state, many researchers have
 demonstrated that affect-aware computers can provide a better performance in
-assisting humans \cite{picard2001toward}. When ILEs embrace these methodologies
+assisting humans
+\cite{picard2001toward,munson2018models,bosch2017affective}, and that
+providing a richer experience while learning to program might improve
+the outcome by making the students more engaged with it
+\cite{asensio2014progamer}. However, when ILEs embrace these methodologies
 and techniques a
-problem arises. Sensors must be used to gather data, in order to perform the
+problem arises: Sensors which are not present in the default computer
+configuration must be used to gather data that will be used to perform the
 recognition of users\' affective states.
 Some sensors rely on physiological readings
-and must be in physical contact with students.
+and must be in physical contact with students \cite{atkinson2017assessing,swansianalyzing}.
 These sensors can be considered
-intrusive or invasive, and can disrupt the student\'s learning experience
+intrusive or invasive, not to mention that you have to acquire them
+and convice the student to wear them while learning.  They can also disrupt the student\'s learning experience
 \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
-Other sensors such as video cameras can also be considered invasive, since
+Other less cumbersome sensors, such as video cameras, which are in
+fact default equipment in most laptops, can also be considered invasive, since
 they always transmit the user's identity, appearance and behavior along with
-the emotional data \cite{picard2001toward}.
+the emotional data \cite{picard2001toward}. So what we would need in
+order to embed recognition of affective states in learning systems
+would be non-invasive, anonymous and also ubiquitous sensors. In this
+paper we are going to test such a kind of system.
 
-In programming courses, students solve programming problems by implementing
+We will use the dynamic of the learning environment itself. In programming courses, students solve programming problems by implementing
 their solution in a particular language; in order to do this, they
 must type the code, some times compile it, and finally run it to see how or if
-it works. This cycle can be repeated many times, as they are in this cycle,
+it works. This cycle can be repeated many times, and as they are in this cycle,
 studens often change their affective state according to their performence.
-Data can be collected while students are working on their assignments.
+Data can, then, be non-intrusively collected while students are
+working on their assignments and typing on the keyboard and
+interacting with the GUI via mouse or other pointing device.
 All the dynamics of actually typing the code and for instance
 the syntax errors they received from the compiler or interpreter, the
-corrections they tried, the elapsed time of each step, the number of attempts,
-compiling errors, exceptions, among others. Data mining of all this data could
+corrections they tried, the elapsed time between each step, the number of attempts,
+compiling errors, exceptions, among others, are collected by the ILE
+and can be used. Data mining of all this data could
 bring us a better understanding of the learning process, and also a better
-model to both the selection of the appropriate content and to evaluate the
+model of the student state for both the selection of the appropriate
+content to be shown next and to evaluate the
 learner\'s competencies. But the problem that remains is to how to recognize
@@
 user\'s interactions (keyboard and mouse) with an emotional state, has been
-established in the preliminary work by Zimmermann
+established in preliminary work by Zimmermann
 et al. \cite{zimmermann2003affective}, but the work experiments where
 conducted by asking users to write a predefined message, not programming
-execercises.
-% since Bosch or whatever has already proved that this is possible (or
-% whatever) --> Establish the state of the art and how you plan to
-% advance from it - JJ
+exercises.
 In order to test the correlation of keyboard and mouse dynamics and an
@@
 as justify better the methdology used and the way data has been
-collected. The platform we presented in that paper originally was designed to collect keyboard and mouse data
+collected. The platform we presented in that paper originally was
+designed to collect keyboard and mouse data 
 from the students\' interaction for data analysis. The proposed method
@@
 An experiment was conducted where students solved a series of programming
-assignments using the web based learning environment.
-Using only the data from the student\'s keyboard and mouse dynamics six affective states where recognized for each of the attempts at solving the assignments.
-Results obtained from the experiment are promising.
-The affective states recognized include the most common in novice programmers according to the study of Bosch, D'Mello \& Mills \cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration.
-For this experiment
+assignments using the web based learning environment. Using only the
+data from the student\'s keyboard and mouse dynamics six affective
+states where recognized for each of the attempts at solving the
+assignments. Results obtained from the experiment are promising. The
+affective states recognized include the most common in novice
+programmers according to the study of Bosch, D'Mello \& Mills
+\cite{bixler2013detecting}: boredom, engagement/flow, confusion and
+frustration. For this experiment 
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour

### 2018-04-20 paper.tex
Revises intro, leaves space for new results

@@
 
-But this of course presupposes that the affective state of the student is available to the
-system. This detection and recognition can be done in different
-ways. In the wider context of the detection of emotions and affective
+But this of course presupposes that the affective state of the student
+is available to the system. This detection and recognition can be done
+in different ways. In the wider context of the detection of emotions and affective
 state, many researchers have
@@
 
-We will use the dynamic of the learning environment itself. In programming courses, students solve programming problems by implementing
+And in order to do that, we will use the dynamic of the learning environment itself. In programming courses, students solve assignments by implementing
 their solution in a particular language; in order to do this, they
-must type the code, some times compile it, and finally run it to see how or if
-it works. This cycle can be repeated many times, and as they are in this cycle,
+must type the code, run it directly or by compiling it to check its results. This cycle can be repeated many times, and as they are in this cycle,
 studens often change their affective state according to their performence.
@@
 interacting with the GUI via mouse or other pointing device.
-All the dynamics of actually typing the code and for instance
-the syntax errors they received from the compiler or interpreter, the
+All the dynamics of actually typing the code and ILE feedback received
+by students, the
 corrections they tried, the elapsed time between each step, the number of attempts,
@@
 In this work, a method for the recognition of affective states through the
-mining of keystroke and mouse dynamics data is proposed.
-The hypothesis is that students can vary the dynamic of keystrokes according
-to their affective state when programming. A correlation of the
-user\'s interactions (keyboard and mouse) with an emotional state, has been
+mining of keystroke and mouse dynamics data is analyzed further after
+its initial presentation in \cite{ijcci17}.
+Since students can vary the dynamic of keystrokes according
+to their affective state when programming, a correlation between the
+user\'s interactions (keyboard and mouse) with an emotional state has been
 established in preliminary work by Zimmermann
@@
 as justify better the methdology used and the way data has been
-collected. The platform we presented in that paper originally was
+collected, extending the results and the methods used. The platform we
+presented in that paper originally was 
 designed to collect keyboard and mouse data 
@@
 \cite{bixler2013detecting}: boredom, engagement/flow, confusion and
-frustration. For this experiment 
+frustration. For the initial experiment 
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
@@
 assignment.
+% in this experiment we will also use... - JJ
 The proposed method could be used in an ensemble with other sensor
@@
 states when they are working on a programming task.
+
 In order to determine what
@@
 
-If a relationship could be modeled, ILEs could adapt the instruction using the
-same devices required to input the program to the computer. This could be
-important because writing programs in a computer is an important learning
-activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
-\cite{lahtinen2005study} is reported that students rated ``working alone on programming coursework'' as a more useful situation than lectures.
+If a relationship between interaction dynamics and affective state could be modeled, ILEs could adapt the instruction using the
+same devices required to input the program to the computer, which
+could constitute a breakthrough since writing programs in a computer
+is the best way to learn programming by actually doing it. A by Lahtinen, Ala-Mutka \& Järvinen
+\cite{lahtinen2005study} reports that students rated ``working alone on programming coursework'' as a more useful situation than lectures.
 
-The structure of this work is organized as follows: next Section
+The rest of the paper is structured as follows: next Section
 presents a series of works related to the proposed method in this paper; Section

### 2018-04-20 paper.tex
setup
Revises some parts of the introduction
Adding that we are looking for real-time systems, as specified in the beginning of the related results section.

@@
 is available to the system. This detection and recognition can be done
-in different ways. In the wider context of the detection of emotions and affective
+in real time in different ways. In the wider context of the detection
+of emotions and affective 
 state, many researchers have
@@
 
-And in order to do that, we will use the dynamic of the learning environment itself. In programming courses, students solve assignments by implementing
-their solution in a particular language; in order to do this, they
-must type the code, run it directly or by compiling it to check its results. This cycle can be repeated many times, and as they are in this cycle,
-studens often change their affective state according to their performence.
-Data can, then, be non-intrusively collected while students are
-working on their assignments and typing on the keyboard and
-interacting with the GUI via mouse or other pointing device.
-All the dynamics of actually typing the code and ILE feedback received
-by students, the
-corrections they tried, the elapsed time between each step, the number of attempts,
-compiling errors, exceptions, among others, are collected by the ILE
-and can be used. Data mining of all this data could
-bring us a better understanding of the learning process, and also a better
-model of the student state for both the selection of the appropriate
-content to be shown next and to evaluate the
-learner\'s competencies. But the problem that remains is to how to recognize
-the student's affective state so that it can be used by an adaptive system.
+And in order to do that, we will use the dynamic of the learning
+environment itself. In programming courses, students solve assignments
+by implementing their solution in a particular language; in order to
+do this, they must type the code, run it directly or by compiling it
+to check its results. This cycle can be repeated many times, and as
+they are in this cycle, studens often change their affective state
+according to their performence.  Data can, then, be non-intrusively
+collected while students are working on their assignments and typing
+on the keyboard and interacting with the GUI via mouse or other
+pointing device.  All the dynamics of actually typing the code and ILE
+feedback received by students, the corrections they tried, the elapsed
+time between each step, the number of attempts, compiling errors,
+exceptions, among others, are collected by the ILE and can be
+used. Data mining of all this data could bring us a better
+understanding of the learning process, and also a better model of the
+student state for both the selection of the appropriate content to be
+shown next and to evaluate the learner\'s competencies. But the
+problem that remains is to how to recognize the student's affective
+state so that it can be used by an adaptive system.
 
-In this work, a method for the recognition of affective states through the
+In this work, a method for the recognition of affective states in real
+time through the
 mining of keystroke and mouse dynamics data is analyzed further after
@@
 
-The rest of the paper is structured as follows: next Section
-presents a series of works related to the proposed method in this paper; Section
-\ref{sec:method} describes the proposed method for the recognition of affective
-states in a web based learning environment for the teaching of programming
-languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
-proposed method and results.
-Finally, we draw some conclusions about our methodology and
-experiments.
+The rest of the paper is structured as follows: next Section presents
+a series of works related to the proposed method in this paper;
+Section \ref{sec:method} describes the proposed method for the
+recognition of affective states in a web based learning environment
+for the teaching of programming languages; next in Section
+\ref{sec:exp} we explain the experimental evaluation of the proposed
+method and results.  Finally, we draw some conclusions about our
+methodology and experiments.
 
@@
 Affect recognition is an active field of research. Many
-methods have been proposed, some require the intervention of the user to fill up
-questionnaires or forms; these selections remain static until the user changes
-the values. These methods are easy to implement, but cannot detect dynamic
+methods have been proposed, some of them requiring the intervention of
+the user to fill up questionnaires or forms; these selections remain static until the user changes
+the values and are easy to implement, but cannot detect dynamic
 changes.  A more dynamic approach requires the use of sensors to capture
-affective states as they change. The context, the environment and the learning
+affective states in real time.
+
+The context, the environment and the learning
 activity determine what kind of sensors can be used. The most common learning

### 2018-04-20 paper.tex

@@
 frustration. For the initial experiment
+
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour

### 2018-04-20 paper.tex
Revising related work

--- 2018-04-20 paper.tex
@@
 changes.  A more dynamic approach requires the use of sensors to capture
-affective states in real time.
+affective states in real time, and that is what we are doing in this
+paper.
 
-The context, the environment and the learning
-activity determine what kind of sensors can be used. The most common learning
-environment is the classroom, a physical space with context to facilitate
+The actual sensors are determined by the context, the environment and the learning
+activity determine. The most common learning
+environment is the classroom, a physical space that provides a context to facilitate
 learning. But learning is possible in a wide variety of settings, such as
-outside-of-school locations and outdoor environments these are sometimes
-referred as ubiquitous learning environments, an example of such environments is
-the work of Yang \cite{yang2006context} in a context-aware environment, but missing affective
-information. There are also virtual learning environments such as the one proposed by Dillenburg
+out-of-school locations and outdoor environments which are sometimes
+referred to as ubiquitous learning environments. An analysis of such
+environments is done by Yang \cite{yang2006context} in a context-aware
+environment, but his work misses affective information. There are also
+virtual learning environments such as the one proposed by Dillenburg 
 \cite{dillenbourg2002virtual}  where learners can have avatars, and virtual places
-where they can play roles and socialize.
+where they can play roles and socialize. In general, the importance of
+affect in learning environments has been recognized at a later stage
+of research.
 
-In the general context of learning there have been some approaches for affect
-recognition. The work of Kapoor y Picard \cite{kapoor2005multimodal} uses a multi modal
-approach using sensory information from facial expressions and postural shifts
+For instance, the work of Kapoor y Picard \cite{kapoor2005multimodal}
+uses a multimodal 
+approach with sensory information from facial expressions and postural shifts
 of the learner combined with information about the learner\'s activity on the
@@
 subject material of that system consisted in lectures of computer
-literacy.
+literacy. In this early stage of research, the importance of privacy
+and non-intrusiveness is set aside.
 
-Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
+Some other works try to introduce affective computing features into
+interactive learning environments, without explicitly detecting of
+using the student's affective state. For instance, Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor}
+propose the integration of an affective reasoner 
 with an agent in a virtual environment. In this work an agent called Steve
@@
 speech. Affective reasoning agents were used for training, by putting students
-in work related situations. Agents could react to the behavior of students, for
+in work related situations. Agents could react to the behavior of students; for
 instance if a student was being careless in a task and was in a dangerous

### 2018-04-20 paper.tex
test
Kernel
Revising related work

@@
 
-There has been very little research reported on the effectiveness of the use of
-keyboard and mouse dynamics as a sensory channel for affective recognition, and
-the few have not been focused on programming. The preliminary work by Zimmerman
-et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
-mouse) with an emotional state, measuring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
-uses sensory data based on the time elapsed between each key press, the task
-given to users was to write a free-text and used linguistic features in order to
-recognize both physical and emotional stress. The results show a classification
-accuracy of 62.5\% for physical stress and 75\% from emotional stress; authors
-argue that these results are comparable with other approaches in affective
-computing.
+There has been very little research reported on the effectiveness of
+the use of keyboard and mouse dynamics as a sensory channel for
+affective recognition, and the few have not been focused on
+programming. The preliminary work by Zimmerman et
+al. \cite{zimmermann2003affective} describes a method to correlate
+user’s interactions (keyboard and mouse) with an emotional state,
+measuring different physiological parameters. The work of Vizer, Zhou
+and Sears \cite{vizer2009automated} also uses sensory data based on
+the time elapsed between each key press, the task given to users was
+to write a free-text and used linguistic features in order to
+recognize both physical and emotional stress. The results show a
+classification accuracy of 62.5\% for physical stress and 75\% from
+emotional stress; authors argue that these results are comparable with
+other approaches in affective computing.
 
@@
 other contexts. Moods may have an impact on programmer\'s performance according
-to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
+to Khan, Hierons and Brinkman \cite{khan2007mood}. It may be possible to detect moods
 on the basis of information regarding the programmer\'s use of the keyboard and
@@
 
-Research works related to Keystroke Dynamics (KD) have been carried out by using either
+It is obviously necessary, in order to detect the moods, to prove that
+there is an actual relationship between them and the keyboard
+activity. In fact, Keystroke Dynamics (KD) has been investigated  by using either
 fixed-texts or free-texts \cite{gunetti2005keystroke}.
@@
 recognition of typing patterns when typing a text of arbitrary-length, e.g., a
-description of an item. However, as noted by Janakiraman y Sim \cite{janakiraman2007keystroke}, most of
+description of an item. However, as noted by Janakiraman and Sim
+\cite{janakiraman2007keystroke}, most of 
 the research regarding KD is done on fixed-text input, the reason being that
@@
  be more useful if KD can handle free text as well as fixed text, this is also a
- requirement if the text is a program.
+ requirement if the text is a program. This is what we are actually
+ using in this paper, since programs are created by the student and do
+ not have to follow exactly the same pattern.
 
@@
 
+\begin{table*}[!t]
+\centering
+\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
+\label{tab_performance}
+    \begin{tabular}{ | l | l | l | }
+    \hline
+    Classifier   &  Accuracy            & kappa                 \\
+    \hline
+    Na\"ive Bayes     & 72.52\%  & 0.339  \\
+    \hline
+    Deep Neural Network     & 78.29\%  & 0.419  \\
+    \hline
+    Random Forest      & 78.81\%  & 0.360  \\
+    \hline
+    Gradient Boosted Trees     & 78.95\%  & 0.415  \\
+    \hline
+    KNN      & 76.19\%  & 0.242  \\
+    \hline
+    Na\"ive Bayes Kernel     & 80.22\%  & 0.346 \\
+    \hline
+    J48     & 78.19 \%  & 0.433  \\
+    \hline
+
+    \end{tabular}
+\end{table*}
+
 In order to determine what affective states a student was experiencing, the

### 2018-04-20 paper.tex
Extends the state of the art

@@
 method based on fuzzy models for the recognition of emotions through KD, MD and
-touch-screen interactions. For a broad review of emotion recognition methods
-based on KD and MD, the work by Kolakowska \cite{kolakowska2013review} is recommended.
+touch-screen interactions. Lim \cite{lim2017detecting} use several
+sensor sets and methods for detecting stress levels, including neural
+networks; the detected stress level is then fed into the inference
+system that decides on adaptation. However, this work focuses only on
+stress level, disregarding the more general affective state. Although
+it is no longer current, the state of the art was described quite
+comprehensively in Kolakowska's paper \cite{kolakowska2013review}.
+
+More recent papers, like the one by Carneiro and Novais \cite{CARNEIRO2017171} takes a
+more comprehensive approach, using also a non-intrusive system based
+on keystroke analysis; several affective states are measured: mental
+fatigue, stress and emotional valence. However, it is not focused
+particularly in programming, which has certain characteristics, but
+uses music to create a certain mood on the subjects. The KD
+methodology, however, is applicable to a wide range of
+situations. Also Kolakowska has recently published a paper
+\cite{Kolakowska2018} assessing the usefulness of KD not only in
+emotion recognition, but also in user authentication. Whatever the
+method, it is indeed proved that KD can be used for detecting a range
+of user mood and attitudes \cite{wrobel2018applicability}. In this
+paper we are more concerned with the data collection and the selection
+of a particular method for detecting the mood with the limited amount
+of data obtained via KD.
 

### 2018-04-20 paper.tex
Fixes bibliography and adds another grant

@@
 \section*{Acknowledgment}
-This work has been supported in part by: de Ministerio espa\~{n}ol de Econom\'{\i}a y Competitividad under project TIN2014-56494-C4-3-P (UGR-EPHEMECH) and by CONACYT PEI Project No. 220590.
+This work has been supported in part by: de Ministerio espa\~{n}ol de
+Econom\'{\i}a y Competitividad under project TIN2014-56494-C4-3-P
+(UGR-EPHEMECH),  DeepBio (TIN2017-85727-C4-2-P) and by CONACYT PEI Project No. 220590.
 

### 2018-04-20 paper.tex
Please @mariosky take a look
There are unresolved comments from the last version. You could take a
look and expand text here and there. There is one near this change,
but several more in conclusions.
Adding new algorithms

@@
 % paper title
-% Titles are generally capitalized except for words such as a, an, and, as,
-% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
-% not capitalized unless they are the first or last word of the title.
-% Linebreaks \\ can be used within to get better formatting as desired.
-% Do not put math or special symbols in the title.
+
 \title{Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments}
 
-\author{\authorname{Mario Garcia Valdez\sup{1}, Amaury Hernandez Aguila\sup{1},
-  Juan-J. Merelo\sup{2} and Alejandra Mancilla Soto\sup{1}}
+\author{\authorname{Mario Garcia Valdez\sup{1}, Juan-J. Merelo\sup{2}, Amaury Hernandez Aguila\sup{1} and Alejandra Mancilla Soto\sup{1}}
 \affiliation{Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag,alejandra.mancilla\)@tectijuana.edu.mx}}
@@
 neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
-tree induction algorithm). The best accuracy was around 75\% and was achieved
+tree induction , Deep Neural Network, Random Forest, Gradient Boosted Trees and Na\"ive Bayes Kernel ). The best accuracy was around 75\% and was achieved
 by the J-48 algorithm. Results show that data gathered from ready-available,
@@
 \cite{bixler2013detecting}: boredom, engagement/flow, confusion and
-frustration. For the initial experiment
+frustration. For the initial experiment \cite{ijcci17}
 
@@
 classifier determined if the student experienced the affective state during the
-assignment.
-% in this experiment we will also use... - JJ
+assignment. In this experiment we will also use 
+Deep Learning \cite{candel2016deep} and  Gradient Boosted Trees 
+via H2O version 3.8.2.6, and the Random Forest and Na\"ive Bayes Kernel 
+implementatios from RapidMiner Studio version 8.1.
+
 The proposed method could be used in an ensemble with other sensor
@@
 and can be found online at \url{https://github.com/mariosky/protoboard}.
-%The main functionality  used for this experiment is
-%shown schematically in Figure \ref{fig_process} . % Don't know what's
-                                % happened to this figure
+The main functionality  used for this experiment is
+shown schematically in Figure \ref{fig_process} . % Don't know what's
+                                % happened to this  
+                                % There it was.
 Users log in to
-Protoboard  by creating an account or by using their account credentials from the
-social network Facebook. The web tutorial begins with three introductory videos that explain
+Protoboard  by creating an account or by using their account 
+credentials from the
+social network Facebook. The web tutorial begins with three introductory 
+videos that explain
 the fundamentals of programming in Python, and how to use the editor to
@@
 with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
-  % Which one? You really have to be _way_ more
-  % specific in this conference - JJ
-  % donde - Mario
-% with th evolutionary algorithm - JJ
 In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of

### 2018-04-20 paper.tex
subsub fixed

@@
 
-\subsubsection{Capturing the Keystroke and Mouse Data}
+\subsection{Capturing the Keystroke and Mouse Data}
 As it was explained before, as a student is trying to solve an assignment,
@@
 
-\subsubsection{Preprocessing of the Keystroke and Mouse Data}
+\subsection{Preprocessing of the Keystroke and Mouse Data}
 

### 2018-04-21 paper.tex
figures
Comments on new experiments

@@
 boredom, frustration, distraction, relaxation and engagement.
-The following classification algorithms have been evaluated: k -nearest
-neighbors,feed forward neural networks, na\"ive Bayes classifier and the J-48
-tree induction , Deep Neural Network, Random Forest, Gradient Boosted Trees and Na\"ive Bayes Kernel ). The best accuracy was around 75\% and was achieved
-by the J-48 algorithm. Results show that data gathered from ready-available,
+The following classification algorithms have been evaluated: k-nearest
+neighbors,feed forward neural networks, na\"ive Bayes classifier, J-48
+tree induction algorithm, deep learning, random forest, gradient boosted trees and na\"ive Bayes Kernel). The best accuracy was around 78\% and was achieved
+by the tree induction algorithms. Results show that data gathered from ready-available,
 non-obtrusive sensors can be used successfully
@@
 
-\begin{figure}[h!tbp]
+\begin{figure*}[h!tbp]
 \centering
-\includegraphics[width=0.45\textwidth]{KMDAffective.png}
+\includegraphics[width=0.80\textwidth]{KMDAffective.png}
 \caption{Code Execution and Affect Recognition Architecture.}
 \label{fig_process}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h!tb]
+\begin{figure*}[h!tb]
 \centering
-\includegraphics[width=0.45\textwidth]{editorRresult.png}
+\includegraphics[width=0.60\textwidth]{editorRresult.png}
 \caption{Web-based editor used for code evaluation. The lower panel
   indicates (in Spanish) the tests passed successfully and the program
-  output.}
+  output \cite{ijcci17}.}
 \label{fig_editor}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[h!tb]
+\begin{figure*}[h!tb]
 \centering
-\includegraphics[width=0.45\textwidth]{KeyDyn.png}
+\includegraphics[width=0.70\textwidth]{KeyDyn.png}
 \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
 \label{figKD}
-\end{figure}
+\end{figure*}
 
@@
 %
-\begin{figure}[!t]
+\begin{figure*}[!t]
 \centering
-\includegraphics[width=0.45\textwidth]{Completed.png}
-\includegraphics[width=0.45\textwidth]{classDist.png}
-\caption{Completed activities and assignments by learner in ascending order (left) and distribution of emotions reported (right).}
+\includegraphics[width=0.48\textwidth]{Completed.png}
+\includegraphics[width=0.48\textwidth]{classDist.png}
+\caption{Completed activities and assignments by learner in ascending order (left) and distribution of emotions reported (right) \cite{ijcci17}.}
 \label{fig_completed}
-\end{figure}
+\end{figure*}
 
@@
 \centering
-\caption{Performance of every classifier with 10-fold cross validation. Accuracy ($\kappa$) }
+\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Accuracy ($\kappa$) }
 \label{tab_performance}
@@
 \caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
-\label{tab_performance}
+\label{tab_perf_flow}
     \begin{tabular}{ | l | l | l | }
     \hline
-    Classifier   &  Accuracy            & kappa                 \\
+    Classifier   &  Accuracy            & $\kappa$                 \\
     \hline
@@
 with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
-In the end, the subset for the frustration classifier includes 11 features, and the subset for the boredom classifier consists of
-13 features.
+In the end, the subset for the frustration classifier includes only 11 features, and 
+the subset for the boredom classifier consists of only 13 features.
+
 % It's not clear what's the fitness function here or what it really does.
@@
 accuracy in the classification of frustration.
+
+Table \ref{tab_perf_flow} shows the performance for the additional classifiers
+evaluated specifically for the Flow/engaged affective state. The goal was to
+increase the performance achieved earlier. Table 1shows the performance for the
+additional classifiers evaluated specifically for the Flow/engaged affective
+state. The goal was to increase the performance achieved earlier.  Again a ten-
+fold cross-validation was conducted. In this case, instead of using an
+evolutionary algorithm to conduct feature selection, a vector of weights was
+optimized. The feature-weight vector assignes a weight to each feature. The
+parameters used for the evolutionary optimization algorithm are the following:
+Population size= 10, generations= 30, selection= tournament, tournament size=
+0.25 (fraction of the population),  initialize probability=0.5,  mutation
+probability = 1/number of attributes,  crossover probability=0.5. Parameters
+used for the Random Forest algorithm: number of trees=10, criterion=gain_ratio,
+maximal depth=20, confidence=0.25, minimal gain=0.1, minimal leaf size=2.  The
+parameters used for are the J48 are the following:  confidence=0.79, minimal
+leaf size=2. For the kNN algorithm, the value k=4 was used. For gradient boosted
+trees: number of trees= 20,  maximal depth=5, minimum rows=10, minimum split
+improvement=0.0, number of bins=20, learning rate=0.1, sample rate=1.0.  Bayes
+Kernel:  laplace correction=True, estimation mode=greedy, minimum bandwidth=0.1,
+number of kernels=10.  Finally for deep learning: activation=Rectifier, hidden
+layer sizes= [50, 50], epochs=10. Again the accuracy and $\kappa$ values of the
+algorithms show the difficulty of having classes that are not uniformly
+distributed, with several algorithms reaching an acceptable accuracy but with a
+low value of  $\kappa$, this is an indication of algorithms classifying almost
+everything as the dominant class. The results of the tree induction algorithms
+are better when we consider both accuracy and  $\kappa$ values. In the case of
+kernel-based or deep learning methods they provide competitive results but at a
+higher computational cost.
 

### 2018-04-21 paper.tex
Changes to conclusions
will leave issues open but I am addressing issue #12.

@@
 \centering
-\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Accuracy ($\kappa$) }
+\caption{Performance of every classifier with 10-fold cross validation \cite{ijcci17}. Each cell shows accuracy and ($\kappa$)}
 \label{tab_performance}
@@
 \centering
-\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation. }
+\caption{Performance of classifiers for Flow/Engaged  with 10-fold cross validation and optimized feature-weights }
 \label{tab_perf_flow}
@@
 evaluated specifically for the Flow/engaged affective state. The goal was to
-increase the performance achieved earlier. Table 1shows the performance for the
-additional classifiers evaluated specifically for the Flow/engaged affective
-state. The goal was to increase the performance achieved earlier.  Again a ten-
+increase the performance achieved earlier. Again a ten-
 fold cross-validation was conducted. In this case, instead of using an
-evolutionary algorithm to conduct feature selection, a vector of weights was
-optimized. The feature-weight vector assignes a weight to each feature. The
+evolutionary algorithm to optimize feature selection, it was used for 
+optimizing a weight vector. 
+The feature-weight vector assigned a weight to each o the features. The
 parameters used for the evolutionary optimization algorithm are the following:
-Population size= 10, generations= 30, selection= tournament, tournament size=
-0.25 (fraction of the population),  initialize probability=0.5,  mutation
-probability = 1/number of attributes,  crossover probability=0.5. Parameters
-used for the Random Forest algorithm: number of trees=10, criterion=gain_ratio,
-maximal depth=20, confidence=0.25, minimal gain=0.1, minimal leaf size=2.  The
-parameters used for are the J48 are the following:  confidence=0.79, minimal
-leaf size=2. For the kNN algorithm, the value k=4 was used. For gradient boosted
-trees: number of trees= 20,  maximal depth=5, minimum rows=10, minimum split
-improvement=0.0, number of bins=20, learning rate=0.1, sample rate=1.0.  Bayes
-Kernel:  laplace correction=True, estimation mode=greedy, minimum bandwidth=0.1,
-number of kernels=10.  Finally for deep learning: activation=Rectifier, hidden
-layer sizes= [50, 50], epochs=10. Again the accuracy and $\kappa$ values of the
+Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
+probability = 1/number of attributes,  crossover probability = 0.5. Parameters
+used for the Random Forest algorithm: number of trees = 10, 
+criterion = gain ratio, maximal depth = 20, 
+confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2.  The
+parameters used for are the J48 are the following:  confidence = 0.79, minimal
+leaf size=2. For the kNN algorithm, the value k = 4 was used. 
+For gradient boosted
+trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
+minimum split 
+improvement = 0.0, number of bins = 20, 
+learning rate = 0.1, sample rate = 1.0.  Bayes
+Kernel: laplace correction = True, estimation mode = greedy, 
+minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.
+
+Again the accuracy and $\kappa$ values of the
 algorithms show the difficulty of having classes that are not uniformly
@@
 dynamics features and the additional preprocessing performed on the
-feature vectors. Out of all methods tested, a multilayer perceptron
-trained with backpropagation, the one tagged ANN in RapidMiner,  gave the best classification
-results.
+feature vectors.  Out of all methods tested, tree induction algorithms
+obtained better results when we take in to consideration both accuracy and  $\kappa$ values.
 
-While these are promising results further experiments are needed. Other experiments
-should focus on novice programmers, but that is left as future
-work. %This
+While these are promising results further experiments are needed. Other experiments should focus on novice programmers, but that is left as future
+work.
+
+Further experiments with more data and more representative examples for all  the
+clasess should give researchers the hability to do stratified samples to balence
+the class distribution and achieve better results. Also more experiments should
+give more insigth on the relationship between affective state  and interactive
+behavior. Results show near 80\% accuracy, these results  are competitive
+against other classifiers of a similar domain, but we do not know yet what is
+the effectiveness when applied to affective-aware learing environments.
+
+%This
 % But you have to comment on how this addresses the problem. Which
@@
 % Is 80% effective?
-% future work will focus on extracting other features from the learners\' interaction
-% such as behavioral data, as proposed by Blikstein
-% \cite{blikstein2011using}; the parameters proposed by this author include the size
-% of the program and changes in the code. Other authors also propose treating the use
-% of certain keys as features e.g., the delete key because is manly used for correcting
-% sentences. Also a multi-modal approach is needed. Since all these need
+%  - done  Mario 
+Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
+\cite{blikstein2011using}; the parameters proposed by this author include the size of the program and changes in the code.
+Other authors also propose treating the use of certain keys as features e.g., the delete key because is manly used for correcting sentences.
+% Also a multi-modal approach is needed. Since all these need
 % a completely new approach, a new methodology will be developed to

### 2018-04-21 paper.tex
Changes

@@
 frustration. For the initial experiment \cite{ijcci17} 
-
 four binary classifiers were compared: J-48 decision trees, k-nearest neighbour

### 2018-04-21 paper.tex
Reviewing corrections. Closes #12. Reviews changes. Closes #11

@@
 feature vectors.  Out of all methods tested, tree induction algorithms
-obtained better results when we take in to consideration both accuracy and  $\kappa$ values.
+obtained better results when we take into consideration both accuracy and  $\kappa$ values.
 
-While these are promising results further experiments are needed. Other experiments should focus on novice programmers, but that is left as future
-work.
-
-Further experiments with more data and more representative examples for all  the
-clasess should give researchers the hability to do stratified samples to balence
+While these are promising results further experiments are
+needed. Other experiments should focus on novice programmers, with
+which we expect to gather more data and more representative examples for all the
+clasess. These experiments should give researchers the hability to do stratified samples to balence
 the class distribution and achieve better results. Also more experiments should
@@
 Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
-\cite{blikstein2011using}; the parameters proposed by this author include the size of the program and changes in the code.
-Other authors also propose treating the use of certain keys as features e.g., the delete key because is manly used for correcting sentences.
-% Also a multi-modal approach is needed. Since all these need
-% a completely new approach, a new methodology will be developed to
-% gather data and take measures.
+\cite{blikstein2011using}; the parameters proposed by this author
+include the size of the program and changes in the code. Other authors
+also propose treating the use of certain keys as features e.g., the
+delete key because is manly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
+gather data and take measures.
 

### 2018-04-21 paper.tex
Created table for GA for feature selection
Please Mario check that what is said is correct. Also try to justify a
bit more the selection of parameters.

@@
 consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
-\cite{epp2011identifying}.
-%and are shown in Table \ref{tab_features}.
+\cite{epp2011identifying} and are shown in Table \ref{tab_features}.
 
@@
 
+\begin{table}[h!tb]
+\centering
+\caption{Parameters used in the evolutionary algorithm for feature
+  selection.}
+\label{tab_ga}
+    \begin{tabular}{ | l | l | }
+      \hline
+      Parameter & Value \\
+      Population size & 700 \\
+      Generations & 30 \\
+      Selection & Tournament \\
+      Tournament size & 25\% of population size \\
+      Crossover & Uniform \\
+      Crossover probability & 0.5 \\
+      Mutation rate & $1 \over 39$ \\
+      Minimum number of features & 5 \\
+    \end{tabular}
+  \end{table}
+
 As part of the data mining process a canonical genetic algorithm was used for feature
-selection with the following parameters: population size 700, 30 generations with
-tournament selection, tournament size was 25\% of population size. A uniform crossover
-with 0.5 probability and mutation rate of 1/39, and a minimum feature size of 5.
+selection with the parameters used shown in Table \ref{tab_ga}. These
+parameters were set heuristically, and were the same for selecting the
+features of all classification algorithms. % Mario check this - JJ
 In the end, the subset for the frustration classifier includes only 11 features, and 
@@
 criterion = gain ratio, maximal depth = 20, 
-confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2.  The
+confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
 parameters used for are the J48 are the following:  confidence = 0.79, minimal
@@
 also propose treating the use of certain keys as features e.g., the
-delete key because is manly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
+delete key because it is mainly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
 gather data and take measures.

### 2018-04-21 paper.tex
Check again new table and comments around it @mariosky

@@
       Parameter & Value \\
+      \hline
       Population size & 700 \\
@@
       Selection & Tournament \\
-      Tournament size & 25\% of population size \\
+      Tournament size & population size / 4 \\
       Crossover & Uniform \\
@@
       Minimum number of features & 5 \\
+            \hline
+
     \end{tabular}
@@
 accuracy in the classification of frustration.
+%
+\begin{table}[h!tb]
+\centering
+\caption{Parameters used in the evolutionary algorithm for feature
+  weight optimization.}
+\label{tab_ga_w}
+    \begin{tabular}{ | l | l | }
+      \hline
+      Parameter & Value \\
+        \hline
+      Population size & 10 \\
+      Generations & 30 \\
+      Selection & Tournament \\
+      Tournament size & population size / 4 \\
+      Crossover & Uniform \\
+      Crossover probability & 0.5 \\
+      Mutation rate & $1 \over 39$ \\
+      Minimum number of features & 5 \\
+        \hline
+    \end{tabular}
+  \end{table}
+% 
+\begin{table*}[!t]
+\centering
+\caption{Parameters used for the different classification algorithms }
+\label{tab_params_ml}
+\label{tab_perf_flow}
+    \begin{tabular}{ | l | l |  }
+
+      Algorithm & Parameters \\
+      \hline
+    \end{tabular}
+  \end{table}
 
@@
 optimizing a weight vector. 
-The feature-weight vector assigned a weight to each o the features. The
-parameters used for the evolutionary optimization algorithm are the following:
-Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
-probability = 1/number of attributes,  crossover probability = 0.5. Parameters
+      The feature-weight vector assigned a weight to each one of the
+      features, and the parameters that have been used in the
+      evolutionary algorithm are shown in Table \ref{tab_ga_w}.
+      % The
+% parameters used for the evolutionary optimization algorithm are the following:
+% Population size = 10, generations = 30, selection = tournament, tournament size = 0.25 (fraction of the population),  initialize probability = 0.5,  mutation
+%       probability = 1/number of attributes,  crossover probability =
+%       0.5.
+      % Please Mario check these parameters and how they have put to
+      % the table; I don't know what's the "initialization
+      % probability" - JJ
+
+      Parameters
 used for the Random Forest algorithm: number of trees = 10, 

### 2018-04-21 paper.tex
Puts parameters in table form

@@
   \end{table}
-
+  %
+  \begin{table*}[h!tb]
+\centering
+\caption{Parameters used in the classification algorithms.}
+\label{tab_clas}
+\begin{tabular}{ | l | p{10cm} | }
+  \hline
+  Algorithm & Parameters \\
+  \hline
+  J48 & Confidence threshold for pruning = 0.25; minimum instances per
+        leaf = 2; folds for reduced error pruning = 3 \\
+  Neural network & Learning algorithm: Back-propagation; hidden layers
+                   = 2, hidden layer size = 20 neurons; learning rate
+                   = 0.6; momentum = 0.6; training iterations = 200 \\
+  \hline
+\end{tabular}
+\end{table*}
+%
 As part of the data mining process a canonical genetic algorithm was used for feature
@@
 the subset for the boredom classifier consists of only 13 features. 
-
+%
 % It's not clear what's the fitness function here or what it really does.
 Maybe we could expand this part of the work, test different evolutionary algorithms and write a paper on feature selection using evolutionary algorithms for EvoStar - JJ
+%
+The selected classifiers where implemented using Rapid Miner, with the
+  parameters shown in Table \ref{tab_clas}.
 
-The selected classifiers where implemented using Rapid Miner, with the
-following parameters: For J48 decision tree algorithm a confidence threshold
-for pruning of 0.25, with a minimum of 2 instances per leaf, and 3 folds for
-reduced error pruning. The feed-forward neural network used a back-propagation
-learning algorithm with two hidden layer of 20 neurons each, with a learning
-rate and momentum of 0.6, and was trained for 200 generations.
+%   For J48 decision tree algorithm a confidence threshold
+% for pruning of 0.25, with a minimum of 2 instances per leaf, and 3 folds for
+% reduced error pruning. The feed-forward neural network used a back-propagation
+% learning algorithm with two hidden layer of 20 neurons each, with a learning
+% rate and momentum of 0.6, and was trained for 200 generations.
 
@@
 \centering
-\caption{Parameters used for the different classification algorithms }
+\caption{Parameters used for the different classification algorithms
+  used for recognizing the Flow/engaged state.}
 \label{tab_params_ml}
-\label{tab_perf_flow}
-    \begin{tabular}{ | l | l |  }
-
+    \begin{tabular}{ | l | p{10cm} |  }
+\hline
       Algorithm & Parameters \\
       \hline 
+      Random Forest &  number of trees = 10, 
+criterion = gain ratio, maximal depth = 20, 
+                      confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2 \\
+      J48 &  confidence = 0.79, minimal leaf size=2 \\
+      kNN & $k$ = 4 \\
+      Gradient boosted trees & number of trees = 20,  maximal depth = 5, minimum rows = 10, 
+minimum split 
+improvement = 0.0, number of bins = 20, 
+                               learning rate = 0.1, sample rate = 1.0 \\
+      Bayes Kernel & laplace correction = True, estimation mode = greedy, 
+                     minimum bandwidth = 0.1, number of kernels = 10 \\
+      Deep learning & activation = Rectifier, hidden layer sizes =
+                      [50, 50], epochs = 10 \\
+      \hline
     \end{tabular}
-  \end{table}
+  \end{table*}
 
@@
       % probability" - JJ
-
-      Parameters
-used for the Random Forest algorithm: number of trees = 10, 
-criterion = gain ratio, maximal depth = 20, 
-confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
-parameters used for are the J48 are the following:  confidence = 0.79, minimal
-leaf size=2. For the kNN algorithm, the value k = 4 was used. 
-For gradient boosted
-trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
-minimum split 
-improvement = 0.0, number of bins = 20, 
-learning rate = 0.1, sample rate = 1.0.  Bayes
-Kernel: laplace correction = True, estimation mode = greedy, 
-minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.
+%
+      The parameters used in this case for the classification
+      algorithms, and the algorithms used are shown in Table \ref{tab_params_ml}
+% used for the Random Forest algorithm: number of trees = 10, 
+% criterion = gain ratio, maximal depth = 20, 
+% confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2.  The
+% parameters used for are the J48 are the following:  confidence = 0.79, minimal
+% leaf size=2. For the kNN algorithm, the value k = 4 was used. 
+% For gradient boosted
+% trees: number of trees = 20,  maximal depth = 5, minimum rows = 10, 
+% minimum split 
+% improvement = 0.0, number of bins = 20, 
+% learning rate = 0.1, sample rate = 1.0.  Bayes
+% Kernel: laplace correction = True, estimation mode = greedy, 
+% minimum bandwidth = 0.1, number of kernels = 10.  Finally for deep learning: activation = Rectifier, hidden layer sizes = [50, 50], epochs = 10.
 

### 2018-04-21 paper.tex
Table correction
Review changes in commit ed2faa3

--- 2018-04-21 paper.tex
@@
 selection with the parameters used shown in Table \ref{tab_ga}. These
-parameters were set heuristically, and were the same for selecting the
-features of all classification algorithms. % Mario check this - JJ
+parameters were set heuristically and were the same for selecting the
+features of all classification algorithms. The population size was changed
+mainly because the feature vector is real-valued and is more computationally demanding.  
+% Mario check this - JJ
+% OK, added another sentence - Mario 
 In the end, the subset for the frustration classifier includes only 11 features, and 
@@
     \end{tabular}
-  \end{table}
+  \end{table*}
 

### 2018-04-21 paper.tex
Reviewed changes
Explaining the GA algorithm

@@
 features of all classification algorithms. The population size was changed
-mainly because the feature vector is real-valued and is more computationally demanding.  
+mainly because some of the algorithms are more computationally demanding. The
+genetic algorithm is included as a RapidMiner operator. In the algorithm, each
+individual consists of a binary list indicating if a feature will be used or
+not. The mutation operator simply changes the switch with the probability
+indicated in the mutation-probability parameter.  When the population is
+initialized each switch has 0.5 probability of being "on".
+
 % Mario check this - JJ
 % OK, added another sentence - Mario 
+% Added a mini explanation of the GA. - Mario
+
 In the end, the subset for the frustration classifier includes only 11 features, and 
@@
 increase the performance achieved earlier. Again a ten-
-fold cross-validation was conducted. In this case, instead of using an
-evolutionary algorithm to optimize feature selection, it was used for 
-optimizing a weight vector. 
-      The feature-weight vector assigned a weight to each one of the
-      features, and the parameters that have been used in the
-      evolutionary algorithm are shown in Table \ref{tab_ga_w}.
+fold cross-validation was conducted. Again, an evolutionary algorithm was used for feature with the parameters shown in Table \ref{tab_ga_w}.
       % The
@@
       % probability" - JJ
+      % Is the probability a feature has to be "on" (used by the algorithm)
+      % initially.
 

### 2018-04-21 paper.tex

--- 2018-04-21 paper.tex
@@
 selection with the parameters used shown in Table \ref{tab_ga}. These
-parameters were set heuristically, and were the same for selecting the
-features of all classification algorithms. % Mario check this - JJ
+parameters were set heuristically and were the same for selecting the
+features of all classification algorithms. The population size was changed
+mainly because some of the algorithms are more computationally demanding. The
+genetic algorithm is included as a RapidMiner operator. In the algorithm, each
+individual consists of a binary list indicating if a feature will be used or
+not. The mutation operator simply changes the switch with the probability
+indicated in the mutation-probability parameter.  When the population is
+initialized each switch has 0.5 probability of being "on".
+
+% Mario check this - JJ
+% OK, added another sentence - Mario 
+% Added a mini explanation of the GA. - Mario
+
 In the end, the subset for the frustration classifier includes only 11 features, and 
@@
       \hline
+      Initialize probability & 0.5 \\
       Population size & 10 \\
@@
       % probability" - JJ
-%
+     % Is the probability a feature has to be "on" (used by the algorithm)
+     % initially.
+% Does that mean that you make probablity = 0.5? Still don't
+% understand, but I have added to the table - JJ
       The parameters used in this case for the classification

### 2018-04-22 paper.tex
Comment
Changes bibliography and spell check

@@
 a web-based platform, a learning management system where students
-interact whith programming excersises. We introduce in this paper a
+interact with programming exorcises. We introduce in this paper a
 series of pre-processing techniques that are able to convert data from keyboard and mouse dynamics captured from students as
@@
 non-obtrusive sensors can be used successfully
-as another input to hyebrid classification models in order to predict an
+as another input to hybrid classification models in order to predict an
 individual\'s affective states, and these in turn can be used to make
@@
 not so much as to frustrate them. Engagement is not only achieved through the
-design of single learning activities but also on the appropiate selection and
+design of single learning activities but also on the appropriate selection and
 sequence in which they are presented.
@@
 intrusive or invasive, not to mention that you have to acquire them
-and convice the student to wear them while learning.  They can also disrupt the student\'s learning experience
+and convince the student to wear them while learning.  They can also disrupt the student\'s learning experience
 \cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
@@
 to check its results. This cycle can be repeated many times, and as
-they are in this cycle, studens often change their affective state
-according to their performence.  Data can, then, be non-intrusively
+they are in this cycle, students often change their affective state
+according to their performance.  Data can, then, be non-intrusively
 collected while students are working on their assignments and typing
@@
 motivation for this research, and update the state of the art, as well
-as justify better the methdology used and the way data has been
+as justify better the methodology used and the way data has been
 collected, extending the results and the methods used. The platform we
@@
 via H2O version 3.8.2.6, and the Random Forest and Na\"ive Bayes Kernel 
-implementatios from RapidMiner Studio version 8.1.
+implementations from RapidMiner Studio version 8.1.
 
@@
 criterion = gain ratio, maximal depth = 20, 
-                      confidence = 0.25, minimal gain = 0.1, minmail leaf size = 2 \\
+                      confidence = 0.25, minimal gain = 0.1, minimal leaf size = 2 \\
       J48 &  confidence = 0.79, minimal leaf size=2 \\
@@
                                learning rate = 0.1, sample rate = 1.0 \\
-      Bayes Kernel & laplace correction = True, estimation mode = greedy, 
+      Bayes Kernel & Laplace correction = True, estimation mode = greedy, 
                      minimum bandwidth = 0.1, number of kernels = 10 \\
@@
 % understand, but I have added to the table - JJ
+% Yes, is basically a binary random selection when creating
+% the population. - Mario
+
       The parameters used in this case for the classification
@@
 
-The aim of this research was to evaluate the use of keyboard and mouse dynamics
-as an appropriate sensory input for an affective recognition system. The context
-of use is a learning environment for programmers, in particular for learning
-assignments consisting in writing short programs interactively. An experimental
-approach was adopted with this objective, by effectively using the
-method with real subjects using the environment.
+The aim of this research was to evaluate the use of keyboard and mouse
+dynamics (KD) as an appropriate sensory input for an affective
+recognition system for students of different levels learning to
+program, specifically carrying out assignments consisting in writing
+short programs interactively. An experimental approach was adopted
+with this objective, by effectively using the method with real
+subjects using the environment and using KD measurements and survey
+responses as training data for several machine learning systems.
 
@@
 fixed-text; however, in this
-work the results obtained were similar to those works applied to fixed-text
+work the results obtained were similar to those works that worked with fixed-text
 dynamics. A possible explanation of this would be the addition of the mouse
@@
 feature vectors. Out of all methods tested, tree induction algorithms
-obtained better results when we take into consideration both accuracy and  $\kappa$ values.
-
-While these are promising results further experiments are
+obtained better results when we take into consideration both accuracy
+and $\kappa$ values. These methods, besides, are able to give us some
+insight on what are the KD features that effectively describe the
+different affective states.
+
+While these are promising results, further experiments are
 needed. Other experiments should focus on novice programmers, with
 which we expect to gather more data and more representative examples for all the
-clasess. These experiments should give researchers the hability to do stratified samples to balence
+classes. These experiments should give researchers the ability to do stratified samples to balance
 the class distribution and achieve better results. Also more experiments should
-give more insigth on the relationship between affective state  and interactive
+give more insight on the relationship between affective state  and interactive
 behavior. Results show near 80\% accuracy, these results  are competitive
 against other classifiers of a similar domain, but we do not know yet what is
-the effectiveness when applied to affective-aware learing environments.
+the effectiveness when applied to affective-aware learning environments.
 
@@
 also propose treating the use of certain keys as features e.g., the
-delete key because it is mainly used for correcting sentences. Also a multi-modal approach is needed. Since all these need a completely new approach, a new methodology will be developed to
+delete key because it is mainly used for correcting sentences. Also a
+multi-modal approach is needed, using attention detectors alongside KD
+dynamics to measure at least the boredom/attention state independently
+of the verbal assessment by the student. Since all these need a
+completely new approach, a new methodology will be developed to 
 gather data and take measures.
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{splncs03}
 % argument is your BibTeX string definitions and bibliography database(s)

### 2018-06-05 paper.tex
Adds citation for figure and table

@@
 interactive learning environments, without explicitly detecting of
-using the student's affective state. For instance, Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor}
-propose the integration of an affective reasoner 
-with an agent in a virtual environment. In this work an agent called Steve
-responds emotionally to other agents and interactive users. The agent simulates
-his emotions through different multimedia modes including facial expressions and
-speech. Affective reasoning agents were used for training, by putting students
-in work related situations. Agents could react to the behavior of students; for
-instance if a student was being careless in a task and was in a dangerous
-situation, Steve would show distress or fear. Also in a virtual environment the
-work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
-investigating the affective transitions that occur throughout narrative-centered
-learning experiences. The analysis of affective state transitions in this work
-replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al.
-\cite{rodrigo2009affective} where
-also engagement/flow dominated self-reported affect. For outdoor environments
-Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
-recognition.  The results about emotion recognition from physiological signals
-achieved a best-case accuracy (86.3\%) for four types of learning emotions.
+using the student's affective state. For instance, Elliott, Rickel y
+Lester \cite{elliott1999lifelike,d2008autotutor} propose the
+integration of an affective reasoner with an agent in a virtual
+environment. In this work an agent called Steve responds emotionally
+to other agents and interactive users. The agent simulates his
+emotions through different multimedia modes including facial
+expressions and speech. Affective reasoning agents were used for
+training, by putting students in work related situations. Agents could
+react to the behavior of students; for instance if a student was being
+careless in a task and was in a dangerous situation, Steve would show
+distress or fear. Also in a virtual environment the work of McQuiggan,
+Robison \& Lester \cite{mcquiggan2010affective} extends this line of
+research by investigating the affective transitions that occur
+throughout narrative-centered learning experiences. The analysis of
+affective state transitions in this work replicated the findings by
+D’Mello et al. \cite{d2008autotutor} and Baker et al.
+\cite{rodrigo2009affective} where also engagement/flow dominated
+self-reported affect. For outdoor environments Shen, Wang \& Shen
+\cite{shen2009affective} augment a pervasive e-learning platform with
+affective recognition.  The results about emotion recognition from
+physiological signals achieved a best-case accuracy (86.3\%) for four
+types of learning emotions.
 
@@
 \includegraphics[width=0.80\textwidth]{KMDAffective.png}
-\caption{Code Execution and Affect Recognition Architecture.}
+\caption{Code Execution and Affect Recognition Architecture. Figure
+  extracted from \cite{ijcci17}}
 \label{fig_process}
@@
 \centering
-\caption{The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
+\caption{The components of the feature vector; adapted from Epp,
+  Lippold \& Mandryk \cite{epp2011identifying}. These components were
+  used in \cite{ijcci17} too.}
 \label{tab_features}

### 2018-06-05 paper.tex
Fixes errors and adds conclusions

@@
 Some sensors rely on physiological readings
-and must be in physical contact with students \cite{atkinson2017assessing,swansianalyzing}.
+and must be in physical contact with students \cite{kula2017assessing,swansianalyzing,gonzalez2017affect}.
 These sensors can be considered
@@
 
-The aim of this research was to evaluate the use of keyboard and mouse
-dynamics (KD) as an appropriate sensory input for an affective
-recognition system for students of different levels learning to
-program, specifically carrying out assignments consisting in writing
+The aim of this research was to present an affective recognition
+system that uses keyboard and mouse
+dynamics (KD) as a non-intrusive sensory input. This system was
+applied to  for students of different levels learning to
+program by carrying out assignments consisting in writing
 short programs interactively. An experimental approach was adopted
-with this objective, by effectively using the method with real
-subjects using the environment and using KD measurements and survey
-responses as training data for several machine learning systems.
-
-The proposed method in this work obtained
-satisfactory results. It is usual for a classification method based on free-text
+with this objective, by effectively taking KD measurements on real
+subjects using the environment; these subjects they responded to
+surveys. Measurements and surveys were merged into a dataset that was
+used for training different machine learning system, with our
+objective being to find out which method yields the best results in
+this environment.
+
+Results obtained with these machine learning systems show that the
+method used for gathering affective data convey sufficient information
+for an adequate classification of affective states; in fact, although
+it is usual for a classification method based on free-text
 to obtain an accuracy and $\kappa$ measure below their counterparts of
-fixed-text; however, in this
-work the results obtained were similar to those works that worked with fixed-text
-dynamics. A possible explanation of this would be the addition of the mouse
-dynamics features and the additional preprocessing performed on the
-feature vectors. Out of all methods tested, tree induction algorithms
+fixed-text; in this work the classification accuracy obtained  with
+the best machine learning systems were similar to those using
+fixed-text. This level of accuracy obtained beyond the state of the
+art in free-text systems might be due to the addition of
+the mouse dynamics features and the additional preprocessing performed on the
+feature vectors.
+
+With data captured and preprocessed as indicated in the methodology,
+and out of all the machine learning algorithms tested, tree induction algorithms
 obtained better results when we take into consideration both accuracy
 and $\kappa$ values. These methods, besides, are able to give us some
-insight on what are the KD features that effectively describe the
+insight on the key KD features that effectively describe the
 different affective states.
@@
 While these are promising results, further experiments are
-needed. Other experiments should focus on novice programmers, with
+needed. In this work we have used programmers with different levels of
+expertise; however, it would be interesting to focus on novice programmers, with
 which we expect to gather more data and more representative examples for all the
 classes. These experiments should give researchers the ability to do stratified samples to balance
-the class distribution and achieve better results. Also more experiments should
+the class distribution and achieve better results. Additional experiments should
 give more insight on the relationship between affective state and interactive
@@
 
-%This
-% But you have to comment on how this addresses the problem. Which
-% should be in the title that is not there yet - JJ
-% You also have to draw a _hard_ conclusion. Is programming using free
-% text better than what? Typing a fixed text?
-% Is 80% effective?
-%  - done  Mario 
-Future work will focus on extracting other features from the learners\' interaction such as behavioral data, as proposed by Blikstein
+A different line of work will focus on extracting other features from
+the learners\' interaction such as behavioral data, as proposed by
+Blikstein 
 \cite{blikstein2011using}; the parameters proposed by this author
@@
 delete key because it is mainly used for correcting sentences. Also a
-multi-modal approach is needed, using attention detectors alongside KD
+multi-modal approach to the detection of the affective state might be
+needed, using attention detectors alongside KD 
 dynamics to measure at least the boredom/attention state independently
